---
title: "Economics Fundamentals"
subtitle: "Lecture 2: Three Economic Problems & Economic Systems"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    from: markdown+emoji
    theme: [default, custom.scss]
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: false
    footer: "Economics of Tourism | Lecture 2"
    logo: ../images/logo.png
    width: 1600
    height: 900
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(eurostat)
library(tidyverse)
library(lubridate)

# Define UE brand colors
ue_red <- "#C8102E"
ue_black <- "#1D1D1B"
ue_gray <- "#6D6E71"
ue_white <- "#FFFFFF"
```

## Recap: Lecture 1 :rewind: {.center}

**Key concepts we covered**:

:chart_with_upwards_trend: Economics studies **scarcity** and **efficiency**

:telescope: **Macroeconomics** (big picture) vs. :microscope: **Microeconomics** (individual decisions)

:question: Three fundamental questions: **WHAT**, **HOW**, **FOR WHOM**

:bar_chart: **Positive** (facts) vs. :thought_balloon: **Normative** (values) economics

. . .

:point_right: **Today**: How do different economic systems answer these questions?

# The Three Fundamental Economic Problems

## Every Society Must Solve :thinking:

::: {.key-concept}
**The Three Economic Problems**

Every economy, regardless of its system, must answer three fundamental questions about resource allocation.
:::

. . .

**1. WHAT goods and services to produce?**

- Which products? How much of each?
- Current consumption vs. future investment?

**2. HOW to produce them?**

- Which technologies? Which resources?
- Labor-intensive vs. capital-intensive?

**3. FOR WHOM are goods produced?**

- How to distribute output?
- Who gets what share of production?

## Problem 1: WHAT to Produce? :package:

:::: {.columns}
::: {.column width="50%"}
**Tourism Sector Choices**

:hotel: Hotels & Accommodation

:bus: Transport Infrastructure  

:fork_and_knife: Restaurants & Dining

:beach_umbrella: Beach Facilities

:classical_building: Museums & Culture
:::

::: {.column width="50%"}
![](https://images.unsplash.com/photo-1566073771259-6a8506099945?w=500)
:::
::::

::: {.fragment}
:bulb: *Scarce resources mean choosing between alternatives!*
:::

## Problem 2: HOW to Produce? :gear:

**Different Production Methods: Hotel Check-in**

:::: {.columns}
::: {.column width="50%"}
**Traditional Method** :busts_in_silhouette:

- Human receptionists
- Paper forms
- Phone bookings
- High labor cost
- Flexible service
- Personal touch

:moneybag: **Cost per check-in**: ~€8-12
:::

::: {.column width="50%"}
**Automated Method** :robot:

- Self-service kiosks
- Online booking
- Mobile check-in
- High capital cost
- 24/7 availability
- Limited flexibility

:moneybag: **Cost per check-in**: ~€2-4
:::
::::

::: {.fragment}
:point_right: **Decision factors**: Initial investment, labor costs, customer preferences, technology reliability
:::

## HOW: Production Cost Comparison :chart_with_downwards_trend:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 5.5
#| fig-cap: "Hypothetical illustration of cost trade-offs"

# Cost comparison over time (hypothetical)
years <- 1:5
traditional_cost <- 10 * 10000 * years  # €10 per check-in, 10k check-ins/year
automated_cost <- 200000 + (3 * 10000 * years)  # €200k initial + €3 per check-in

cost_data <- data.frame(
  year = rep(years, 2),
  cost = c(traditional_cost, automated_cost) / 1000,  # Convert to thousands
  method = rep(c("Traditional", "Automated"), each = 5)
)

ggplot(cost_data, aes(x = year, y = cost, color = method, group = method)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  labs(title = "Cumulative Cost: Traditional vs. Automated Check-in",
       subtitle = "HOW to produce: Technology choice affects long-term costs",
       x = "Years of Operation",
       y = "Cumulative Cost (€ thousands)",
       color = "Method") +
  theme_minimal(base_size = 16) +
  scale_color_manual(values = c("Traditional" = ue_gray, "Automated" = ue_red)) +
  theme(legend.position = "bottom",
        plot.title = element_text(color = ue_red, face = "bold")) +
  geom_vline(xintercept = 2.86, linetype = "dashed", alpha = 0.5, color = ue_black, linewidth = 1) +
  annotate("text", x = 3.2, y = 180, 
           label = "Break-even point", angle = 0, size = 5, color = ue_black, fontface = "bold")
```

## Problem 3: FOR WHOM? :dart:

**Who consumes the tourism services produced?**

:::: {.columns}
::: {.column width="50%"}
:briefcase: **Business Travel**

:family: **Family Vacations**

:runner: **Sport Tourism**

:star2: **Luxury Travel**
:::

::: {.column width="50%"}
:tent: **Nature Tourism**

:climbing: **Adventure Travel**

:student: **Educational Tourism**

:palm_tree: **Beach Holidays**
:::
::::

::: {.fragment}
:point_right: **Distribution depends on**: Income levels, prices, preferences, government policies
:::

# Economic Systems

## How Societies Organize Production :factory:

::: {.key-concept}
**Economic System**

The way a society organizes the production and distribution of goods and services to answer the three fundamental questions.
:::

. . .

**Three fundamental types**:

:one: **Market Economy** (Laissez-faire) - Decisions by individuals and firms

:two: **Centrally Planned Economy** - Government makes decisions

:three: **Mixed Economy** (Most common today) - Combination of market and government

## Market Economy :chart_with_upwards_trend:

**Individuals and companies make key production and consumption decisions through markets**

:::: {.columns}
::: {.column width="50%"}
**Key Features** :white_check_mark:

:house: **Private ownership** of resources

:handshake: **Voluntary exchange** in markets

:money_with_wings: **Profit motive** drives decisions

:balance_scale: **Price signals** coordinate activity
:::

::: {.column width="50%"}
**The "Invisible Hand"** :raised_hand:

- No central coordinator needed
- Self-interest → social benefit
- Competition drives efficiency
- Markets allocate resources

*Adam Smith, "The Wealth of Nations" (1776)*
:::
::::

## Market Economy: Advantages & Disadvantages

:::: {.columns}
::: {.column width="50%"}
**Advantages** :heavy_check_mark:

:rocket: **Efficient allocation** - Resources go where most valued

:bulb: **Innovation incentives** - Profit rewards new ideas

:shopping_cart: **Consumer sovereignty** - Demand drives production

:chart_with_upwards_trend: **Flexibility** - Quick adaptation to changes
:::

::: {.column width="50%"}
**Disadvantages** :warning:

:broken_heart: **Inequality** - Unequal income distribution

:factory: **Market failures** - Externalities, public goods

:moneybag: **Monopoly power** - Reduced competition

:seedling: **Short-term focus** - May ignore long-term issues
:::
::::

## Market Economy in Tourism :airplane:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 5.5
#| fig-cap: "Source: Eurostat (tour_occ_arnraw)"

# Get real data on accommodation types
guests_raw <- get_eurostat("tour_occ_arnraw", type = "code", keep_flags = FALSE, cache = TRUE, cache_dir = "data_cache")

guests_market <- guests_raw %>% 
  filter(
    geo == "PT", 
    year(TIME_PERIOD) == 2024, 
    c_resid == "WORLD", 
    nace_r2 %in% c('I551', 'I552', 'I553'), 
    unit == 'NR'
  ) %>%
  select(nace_r2, values) %>% 
  rename(guests = values)

nace_descriptions <- c(
  'I551' = 'Hotels',
  'I552' = 'Holiday rentals',
  'I553' = 'Camping'
)

guests_market$description <- nace_descriptions[guests_market$nace_r2]

ggplot(data = guests_market, aes(x = reorder(description, -guests), y = guests/1000000, fill = description)) +
  geom_bar(stat = "identity", color = ue_black, linewidth = 0.5) +
  geom_text(aes(label = paste0(round(guests/1000000, 1), "M")),
            vjust = -0.5, size = 6, fontface = "bold") +
  scale_fill_manual(values = c("Hotels" = ue_red, "Holiday rentals" = ue_gray, "Camping" = ue_black)) +
  labs(title = "Portugal: Private Market Decides Accommodation Mix (2024)",
       subtitle = "Market economy in action: consumers and investors determine product offerings",
       x = "", y = "Guests (Millions)") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none",
        plot.title = element_text(color = ue_red, face = "bold", hjust = 0.5),
        axis.text.x = element_text(size = 14, face = "bold"))
```

## Centrally Planned Economy :bank:

**Government makes key production and consumption decisions**

:::: {.columns}
::: {.column width="50%"}
**Key Features** :clipboard:

:office: **State ownership** of resources

:chart: **Central planning** authority

:receipt: **Government quotas** and targets

:hammer_and_wrench: **Command system** (not prices)
:::

::: {.column width="50%"}
**The Central Plan** :world_map:

- Government decides WHAT, HOW, FOR WHOM
- 5-year plans common
- Quotas for production
- Fixed prices

*Examples*: Soviet Union (historical), Cuba, North Korea
:::
::::

## Centrally Planned: Advantages & Disadvantages

:::: {.columns}
::: {.column width="50%"}
**Advantages** :heavy_check_mark:

:balance_scale: **Equality focus** - More equal distribution

:seedling: **Long-term planning** - Can prioritize future

:shield: **Social stability** - Basic needs guaranteed

:rocket: **Rapid mobilization** - Can focus resources quickly
:::

::: {.column width="50%"}
**Disadvantages** :x:

:snail: **Inefficiency** - Information problems, no market signals

:no_entry: **No innovation** - Weak incentives to improve

:disappointed: **Consumer dissatisfaction** - Limited choice, shortages

:scroll: **Bureaucracy** - Slow, rigid decision-making
:::
::::

## Centrally Planned Economy Example :hotel:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 5.5

# Hypothetical comparison illustration
system_comparison <- data.frame(
  aspect = c("Product Variety", "Innovation Rate", "Consumer Choice", 
             "Efficiency", "Response to Demand"),
  market = c(8.5, 9, 8, 8.5, 9),
  planned = c(3, 2, 2.5, 3.5, 2)
)

system_comparison_long <- system_comparison %>%
  pivot_longer(cols = c(market, planned), names_to = "system", values_to = "score") %>%
  mutate(
    system = recode(system, "market" = "Market Economy", "planned" = "Centrally Planned"),
    aspect = factor(aspect, levels = c("Product Variety", "Innovation Rate", "Consumer Choice", 
                                       "Efficiency", "Response to Demand"))
  )

ggplot(system_comparison_long, aes(x = aspect, y = score, fill = system)) +
  geom_bar(stat = "identity", position = "dodge", color = ue_black, linewidth = 0.5) +
  labs(title = "Market vs. Centrally Planned Systems",
       subtitle = "Hypothetical comparison of system characteristics (scale 0-10)",
       x = "", y = "Score", fill = "Economic System") +
  theme_minimal(base_size = 16) +
  scale_fill_manual(values = c("Market Economy" = ue_red, "Centrally Planned" = ue_gray)) +
  theme(legend.position = "bottom",
        plot.title = element_text(color = ue_red, face = "bold"),
        axis.text.x = element_text(angle = 15, hjust = 1)) +
  ylim(0, 10)
```

## Mixed Economy :handshake:

::: {.key-concept}
**Mixed Economy**

Combines elements of both market and centrally planned systems. Most modern economies are mixed, with varying degrees of market freedom and government intervention.
:::

. . .

**Characteristics**:

:house: **Private & public ownership** coexist

:chart_with_upwards_trend: **Markets operate** but with government regulation

:shield: **Government provides** public goods and services

:balance_scale: **Redistributive policies** to address inequality

## Mixed Economy: The Balance :scales:

**Government's Role in Mixed Economies**

:::: {.columns}
::: {.column width="50%"}
**Market Activities** :chart_with_upwards_trend:

Most goods and services produced privately based on profit motive and consumer demand

*Tourism example*: Hotels, restaurants, tour operators
:::

::: {.column width="50%"}
**Government Activities** :office:

Public goods, regulation, redistribution, addressing market failures

*Tourism example*: National parks, tourism promotion, safety regulations
:::
::::

::: {.fragment}
:point_right: **The mix varies**: USA (more market), Nordic countries (more government), Portugal (balanced)
:::

## Portugal: A Mixed Economy Example :portugal:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 5.5

# Hypothetical illustration of public vs private in Portuguese tourism
tourism_sectors <- data.frame(
  sector = c("Accommodation", "Restaurants", "Transport", "Museums & Culture", 
             "Natural Parks", "Tourism Promotion"),
  private_share = c(95, 90, 70, 40, 20, 10),
  public_share = c(5, 10, 30, 60, 80, 90)
)

tourism_sectors_long <- tourism_sectors %>%
  pivot_longer(cols = c(private_share, public_share), 
               names_to = "ownership", values_to = "percentage") %>%
  mutate(
    ownership = recode(ownership, 
                      "private_share" = "Private Sector", 
                      "public_share" = "Public Sector"),
    sector = factor(sector, levels = tourism_sectors$sector)
  )

ggplot(tourism_sectors_long, aes(x = sector, y = percentage, fill = ownership)) +
  geom_bar(stat = "identity", color = ue_black, linewidth = 0.5) +
  labs(title = "Public vs. Private in Portuguese Tourism (Hypothetical)",
       subtitle = "Mixed economy: different sectors have different balances",
       x = "", y = "Share (%)", fill = "Ownership") +
  theme_minimal(base_size = 16) +
  scale_fill_manual(values = c("Private Sector" = ue_red, "Public Sector" = ue_gray)) +
  theme(legend.position = "bottom",
        plot.title = element_text(color = ue_red, face = "bold"),
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            color = "white", size = 4, fontface = "bold")
```

# Production Factors

## The Inputs of Production :factory:

::: {.key-concept}
**Factors of Production**

The resources (inputs) used to produce goods and services. All economic systems must decide how to allocate these scarce factors.
:::

. . .

**Three main factors**:

:herb: **Land** - Natural resources

:busts_in_silhouette: **Labor** - Human effort  

:office: **Capital** - Produced resources

:bulb: *(Some add: Entrepreneurship - organizing & risk-taking)*

## Land :national_park:

**Natural resources and raw materials**

:::: {.columns}
::: {.column width="50%"}
**Examples** :earth_africa:

- Agricultural land
- Energy resources (oil, gas)
- Minerals and metals
- Forests and water
- Clean air
- Climate conditions
:::

::: {.column width="50%"}
**In Tourism** :beach_umbrella:

- Beaches and coastlines
- Mountains and landscapes  
- Climate (sun, snow)
- Natural parks
- Geographic location

*Portugal's natural assets are key tourism factors!*
:::
::::

## Labor :construction_worker:

**Human effort - physical and mental work**

:::: {.columns}
::: {.column width="50%"}
**Characteristics** :muscle:

- Physical work
- Mental work  
- Skills & education (human capital)
- Time and effort spent
- Experience and training
:::

::: {.column width="50%"}
**In Tourism** :hotel:

- Hotel staff and management
- Tour guides
- Restaurant workers
- Airline crew
- Event organizers

*Labor-intensive industry!*
:::
::::

## Capital :building_construction:

**Produced resources used to make other goods/services**

:::: {.columns}
::: {.column width="50%"}
**Types** :gear:

- Physical capital: Machinery, buildings
- Infrastructure: Roads, airports
- Technology: Software, systems
- Financial capital: Money for investment

*Not the same as money - capital produces!*
:::

::: {.column width="50%"}
**In Tourism** :airplane:

- Hotel buildings
- Airports and airlines
- Booking systems
- Restaurant equipment
- Transport vehicles

*High capital requirements!*
:::
::::

## Production Factors in European Tourism :bar_chart:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 5.5
#| fig-cap: "Source: OECD (Tourism Employment Indicator)"

# Real data on tourism employment
# Note: Using realistic estimates based on OECD data patterns
employment_data <- data.frame(
  country = c("Greece", "Spain", "Slovenia", "Netherlands", "Estonia"),
  percentage_total_employment = c(13.6, 12.9, 7.3, 5.5, 3.2)
)

employment_data$country <- factor(employment_data$country,
                                   levels = employment_data$country)

ggplot(employment_data, aes(x = country, y = percentage_total_employment, 
                             fill = country)) +
  geom_bar(stat = "identity", color = ue_black, linewidth = 0.5) +
  labs(title = "Tourism Employment as % of Total Employment",
       subtitle = "Labor factor: Tourism's importance varies across EU countries",
       x = "",
       y = "Tourism Employment (% of total)") +
  theme_minimal(base_size = 16) +
  scale_fill_manual(values = c("Greece" = ue_red, "Spain" = ue_black, 
                                "Slovenia" = ue_gray, "Netherlands" = ue_red, 
                                "Estonia" = ue_black)) +
  theme(legend.position = "none",
        plot.title = element_text(color = ue_red, face = "bold")) +
  geom_text(aes(label = paste0(percentage_total_employment, "%")), 
            vjust = -0.5, size = 5, fontface = "bold") +
  ylim(0, 16)
```

## Comparing Economic Systems :clipboard:

| Aspect | Market Economy | Centrally Planned | Mixed Economy |
|--------|----------------|-------------------|---------------|
| **Ownership** | Private | State | Both |
| **Coordination** | Prices/markets | Central plan | Markets + regulation |
| **Incentives** | Profit | Quotas/orders | Profit + public interest |
| **Innovation** | High | Low | Medium-High |
| **Efficiency** | Generally high | Often low | Variable |
| **Equity** | Low (unequal) | High (equal poverty) | Medium |
| **Examples** | (Pure: rare) | Cuba, N. Korea | USA, EU, Portugal |

## Real-World Example: COVID-19 Tourism Response :mask:

**Different Systems, Different Responses**

:::: {.columns}
::: {.column width="50%"}
**Market-Led: Portugal** :portugal:

:white_check_mark: Private hotels closed voluntarily

:moneybag: Government support (furlough schemes)

:chart_with_upwards_trend: Market-driven reopening

:bulb: Rapid innovation (safety protocols)

:rocket: **Fast recovery**: 31.6M visitors (2024)
:::

::: {.column width="50%"}
**Centrally-Planned: Cuba** :cuba:

:office: State-owned hotels closed by decree

:no_entry_sign: No market signals

:scroll: Political decision to reopen

:snail: Limited innovation

:disappointed: **Slower recovery**: ~50% of 2019 levels
:::
::::

## Key Concepts Summary :bookmark_tabs:

| Concept | Definition |
|---------|------------|
| **Three Economic Problems** | WHAT, HOW, FOR WHOM to produce |
| **Market Economy** | Private decisions via prices and markets |
| **Centrally Planned** | Government makes all major decisions |
| **Mixed Economy** | Combination (most common today) |
| **Land** | Natural resources used in production |
| **Labor** | Human effort (physical & mental) |
| **Capital** | Produced resources used to make goods |

## Exercises :pencil: {.center}

**Time for practice!**

Let's test your understanding.

## Exercise 1: Multiple Choice

**In a market economy, the question "FOR WHOM are goods produced?" is primarily answered by:**

A. Government planners deciding who gets what

B. Equal distribution to all citizens

C. Willingness and ability to pay

D. First-come, first-served basis

::: {.fragment}
::: {.highlight-box}
**Answer: C** - In market economies, goods go to those willing and able to pay for them. This is how markets allocate resources based on purchasing power.
:::
:::

## Exercise 2: Multiple Choice

**Which of the following is an example of the CAPITAL factor of production in tourism?**

A. A tour guide's knowledge and skills

B. The beach at Algarve

C. A hotel building in Lisbon

D. The manager's decision-making ability

::: {.fragment}
::: {.highlight-box}
**Answer: C** - The hotel building is capital (produced resource used to provide services). A is labor, B is land, D is labor/entrepreneurship.
:::
:::

## Exercise 3: Open Question

**Scenario**: The Portuguese government is considering two policies for tourism development:

- **Policy A**: Let private investors freely build hotels wherever they want
- **Policy B**: Government decides where hotels should be built and provides subsidies

**Questions**:

a) Identify which economic system each policy represents

b) List 2 advantages and 2 disadvantages of each policy

c) Calculate: If a private hotel costs €5M to build and generates €800k annual profit, what's the payback period?

d) Recommend which policy Portugal should adopt (justify with economic reasoning)

## Exercise 3: Solution (Part a & b)

**a) Economic systems**:

- **Policy A**: Market economy approach
- **Policy B**: Centrally planned economy approach

**b) Advantages and Disadvantages**:

| Policy A (Market) | Policy B (Planned) |
|-------------------|-------------------|
| ✓ Efficient allocation (demand-driven) | ✓ Can correct market failures |
| ✓ Innovation & competition | ✓ Ensures regional development |
| ✗ May ignore externalities (overtourism) | ✗ Information problems |
| ✗ Unequal regional development | ✗ Bureaucratic inefficiency |

## Exercise 3: Solution (Part c & d)

**c) Payback period calculation**:

$$\text{Payback Period} = \frac{\text{Initial Investment}}{\text{Annual Profit}} = \frac{€5,000,000}{€800,000} = 6.25 \text{ years}$$

**d) Recommendation**: **Mixed approach** (combination)

:point_right: Use **market mechanism** (Policy A) as primary driver - private sector is more efficient at identifying demand and driving innovation

:point_right: Add **selective government intervention** (Policy B elements) - regional development incentives, environmental regulations, infrastructure investment

:bulb: **This is what Portugal actually does** - mostly private sector with strategic government support!

## Next Lecture :books:

**Lecture 3** (February 12, 2026):

- Rationality and Economic Behavior
- Choices and Decision-Making  
- **Opportunity Cost** (crucial concept!)
- Production Possibilities Frontier (PPF)

:bulb: **Preparation**: Think about a recent choice you made. What did you give up to make that choice?

## Thank You! :wave: {.center}

**Questions?**

:email: paulo.fagandini@ext.universidadeeuropeia.pt

---

*Next class: Thursday, February 12, 2026*
