---
title: "Economics Fundamentals"
subtitle: "Lecture 4: Production Factors & Production Possibilities Frontier"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    theme: default
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: false
    footer: "Economics of Tourism | Lecture 4"
    logo: ../images/logo.png
    include-in-header:
      text: |
        <style>
        .reveal .caption {
          font-size: 0.5em;
        }
        /* This targets the figure captions specifically */
        figcaption {
          font-size: 0.5em !important;
        }
        </style>
---

## Recap: Lecture 3 {.center}

**Key Concepts from Last Class**:

- **Rationality**: Cost-benefit decision making
- **Marginal Analysis**: Decisions at the margin (MB vs MC)
- **Opportunity Cost**: Value of best alternative foregone
- **Sunk Costs**: Ignore them in decisions!

**Today**: We build on opportunity cost to understand:

- Production factors (inputs)
- **Production Possibilities Frontier (PPF)**
- Society-level trade-offs and efficiency

## Production Factors (Inputs)

**To produce goods and services, economies use three main factors**:

:::: {.columns}
::: {.column width="33%"}
### Land
**Natural Resources**

- Agricultural land
- Minerals & energy
- Water resources
- Natural beauty
- Climate

*Tourism*: Beaches, mountains, heritage sites
:::

::: {.column width="33%"}
### Labor
**Human Effort**

- Physical work
- Mental work
- Skills & training
- Time & effort
- Entrepreneurship

*Tourism*: Staff, guides, managers, chefs
:::

::: {.column width="33%"}
### Capital
**Produced Goods**

- Buildings
- Machinery
- Infrastructure
- Technology
- Equipment

*Tourism*: Hotels, planes, booking systems
:::
::::

## Factor Payments

**Each factor receives compensation**:

| Factor | Payment Name | Example in Tourism |
|--------|-------------|-------------------|
| **Land** | Rent | Payment for beach concession, property lease |
| **Labor** | Wages/Salaries | Hotel staff wages, tour guide fees |
| **Capital** | Interest | Return on hotel investment, equipment loans |
| **Entrepreneurship** | Profit | Hotel owner's profit, airline earnings |

**Income Distribution**: Sum of all factor payments determines who gets what (FOR WHOM question)

## Tourism Production: Factor Intensive Example

**Different tourism products use factors in different proportions**:

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

library(ggplot2)
library(dplyr)

# Factor intensity by tourism type
tourism_types <- data.frame(
  type = rep(c("Beach Resort", "Cultural Tour", "Adventure Tourism", "Cruise Ship"), each = 3),
  factor = rep(c("Land", "Labor", "Capital"), 4),
  percentage = c(
    # Beach Resort
    40, 30, 30,
    # Cultural Tour
    15, 60, 25,
    # Adventure Tourism
    50, 35, 15,
    # Cruise Ship
    5, 25, 70
  )
)

tourism_types$type <- factor(tourism_types$type,
                              levels = c("Beach Resort", "Cultural Tour", 
                                        "Adventure Tourism", "Cruise Ship"))

ggplot(tourism_types, aes(x = type, y = percentage, fill = factor)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Factor Intensity in Different Tourism Products",
       subtitle = "Different activities require different combinations of Land, Labor, and Capital",
       x = "",
       y = "Percentage of Total Input Cost",
       fill = "Production Factor") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("Land" = "#27ae60", 
                                "Labor" = "#e74c3c",
                                "Capital" = "#3498db")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 20, hjust = 1)) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_fill(vjust = 0.5), 
            color = "white", size = 4, fontface = "bold")
```

**Insight**: Cruise tourism is **capital-intensive**, cultural tours are **labor-intensive**, beach resorts balanced

## Production Possibilities Frontier (PPF)

**Definition**: The PPF shows the **maximum combinations** of two goods an economy can produce given:

- Fixed quantity of production factors (land, labor, capital)
- Current technology level
- Full employment of resources

**Key Insight**: With scarcity, producing more of one good requires producing less of another

→ **This is opportunity cost visualized!**

## Simple PPF Example: Tourism Economy

**Imagine an economy producing only**:

- **Beach Tourism** (hotels, resorts)
- **Cultural Tourism** (museums, tours, heritage sites)

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Create PPF curve
beach <- seq(0, 100, by = 1)
cultural <- 100 * (1 - (beach/100)^0.7)

ppf_basic <- data.frame(beach, cultural)

# Points
points <- data.frame(
  beach = c(0, 30, 60, 85, 100, 50),
  cultural = c(100, 81, 57, 32, 0, 40),
  label = c("A", "B", "C", "D", "E", "U"),
  type = c(rep("On PPF", 5), "Inefficient")
)

ggplot(ppf_basic, aes(x = beach, y = cultural)) +
  geom_line(size = 2, color = "#3498db") +
  geom_ribbon(aes(ymin = 0, ymax = cultural), alpha = 0.15, fill = "#3498db") +
  geom_point(data = points, aes(x = beach, y = cultural, color = type, shape = type), 
             size = 6) +
  geom_text(data = points, aes(x = beach, y = cultural, label = label), 
            vjust = -1.5, size = 6, fontface = "bold") +
  labs(title = "Production Possibilities Frontier: Tourism Economy",
       subtitle = "Maximum combinations of beach and cultural tourism capacity",
       x = "Beach Tourism Capacity (units)",
       y = "Cultural Tourism Capacity (units)") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("On PPF" = "#e74c3c", "Inefficient" = "#95a5a6")) +
  scale_shape_manual(values = c("On PPF" = 16, "Inefficient" = 4)) +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  xlim(-5, 110) + ylim(-5, 110) +
  annotate("text", x = 85, y = 75, 
           label = "Efficient\n(On frontier)", size = 5, color = "#27ae60") +
  annotate("text", x = 50, y = 30, 
           label = "Inefficient\n(Unemployed resources)", size = 5, color = "#95a5a6")
```

## Interpreting the PPF

**Point A** (0, 100): All resources → cultural tourism, no beach tourism

**Point E** (100, 0): All resources → beach tourism, no cultural tourism

**Point C** (60, 57): Mixed economy, balanced allocation

**Point U** (50, 40): **Inefficient** – inside PPF

- Resources unemployed or misallocated
- Could produce more of both goods!

**Beyond PPF**: **Impossible** with current resources & technology

## Opportunity Cost on the PPF

**Moving along the PPF reveals opportunity costs**:

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Detailed PPF with opportunity cost calculation
beach_detailed <- seq(0, 100, by = 5)
cultural_detailed <- 100 * (1 - (beach_detailed/100)^0.7)

ppf_detailed <- data.frame(beach_detailed, cultural_detailed)

# Movement from B to C
move_points <- data.frame(
  beach = c(30, 60),
  cultural = c(81, 57),
  label = c("B", "C")
)

ggplot(ppf_detailed, aes(x = beach_detailed, y = cultural_detailed)) +
  geom_line(size = 2, color = "#3498db") +
  geom_point(data = move_points, aes(x = beach, y = cultural), 
             size = 7, color = "#e74c3c") +
  geom_text(data = move_points, aes(x = beach, y = cultural, label = label), 
            vjust = -1.5, size = 7, fontface = "bold") +
  labs(title = "Opportunity Cost: Moving from B to C",
       subtitle = "To gain 30 units of beach tourism, must sacrifice 24 units of cultural tourism",
       x = "Beach Tourism Capacity (units)",
       y = "Cultural Tourism Capacity (units)") +
  theme_minimal(base_size = 14) +
  # Horizontal arrow (gain in beach)
  annotate("segment", x = 30, xend = 60, y = 81, yend = 81,
           arrow = arrow(length = unit(0.4, "cm"), type = "closed"), 
           size = 1.5, color = "#27ae60") +
  annotate("text", x = 45, y = 85, 
           label = "GAIN: +30 beach units", size = 5, color = "#27ae60", fontface = "bold") +
  # Vertical arrow (loss in cultural)
  annotate("segment", x = 60, xend = 60, y = 81, yend = 57,
           arrow = arrow(length = unit(0.4, "cm"), type = "closed"), 
           size = 1.5, color = "#e74c3c") +
  annotate("text", x = 70, y = 69, 
           label = "COST: -24 cultural units\n(Opportunity Cost)", 
           size = 5, color = "#e74c3c", fontface = "bold") +
  xlim(-5, 110) + ylim(-5, 110)
```

**Opportunity Cost of moving B → C**: 24 cultural units

## Increasing Opportunity Costs

**Key Feature**: PPF is typically **bowed outward** (concave)

**Why?** → **Law of Increasing Opportunity Costs**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Calculate marginal opportunity costs
calc_points <- data.frame(
  beach = c(0, 20, 40, 60, 80, 100),
  cultural = 100 * (1 - (c(0, 20, 40, 60, 80, 100)/100)^0.7)
)

calc_points$beach_change <- c(NA, diff(calc_points$beach))
calc_points$cultural_change <- c(NA, diff(calc_points$cultural))
calc_points$opp_cost <- -calc_points$cultural_change / calc_points$beach_change

# Create visualization
opp_cost_data <- calc_points[2:6, ]

ggplot(opp_cost_data, aes(x = beach, y = opp_cost)) +
  geom_line(size = 1.5, color = "#e74c3c") +
  geom_point(size = 5, color = "#e74c3c") +
  labs(title = "Increasing Opportunity Costs Along PPF",
       subtitle = "Each additional unit of beach tourism costs MORE cultural tourism foregone",
       x = "Beach Tourism Production Level",
       y = "Opportunity Cost\n(Cultural units per Beach unit)") +
  theme_minimal(base_size = 14) +
  geom_text(aes(label = round(opp_cost, 2)), 
            vjust = -1, size = 5, fontface = "bold") +
  ylim(0, max(opp_cost_data$opp_cost) * 1.2)
```

**Explanation**: Resources are not perfectly adaptable. As you shift more resources to beach tourism, you use resources less suited for it (higher cost).

## Why PPF is Bowed: Intuition

**Example**: Converting resources from cultural to beach tourism

**Early conversions (A → B)**:

- Use coastal land perfect for beaches
- Relocate tour guides who can also work in resorts
- Low opportunity cost (~0.95 cultural per beach unit)

**Later conversions (D → E)**:

- Must use inland locations (poor beaches)
- Relocate museum curators (not suited for resort work)
- High opportunity cost (~2.25 cultural per beach unit)

**Result**: **Bowed PPF** reflects increasing difficulty of conversion

## Productive Efficiency

**An economy is productively efficient when**:

- It operates **ON** the PPF (not inside)
- Cannot produce more of one good without producing less of another
- Resources are fully employed and optimally allocated

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Efficiency comparison
efficiency_points <- data.frame(
  beach = c(50, 50, 90),
  cultural = c(70, 40, 25),
  status = c("Efficient (on PPF)", "Inefficient (inside)", "Efficient (on PPF)"),
  label = c("Point A", "Point B", "Point C")
)

beach_curve <- seq(0, 100, by = 1)
cultural_curve <- 100 * (1 - (beach_curve/100)^0.7)
ppf_eff <- data.frame(beach_curve, cultural_curve)

ggplot(ppf_eff, aes(x = beach_curve, y = cultural_curve)) +
  geom_line(size = 2, color = "#3498db") +
  geom_ribbon(aes(ymin = 0, ymax = cultural_curve), alpha = 0.1, fill = "#27ae60") +
  geom_point(data = efficiency_points, aes(x = beach, y = cultural, color = status), 
             size = 7) +
  geom_text(data = efficiency_points, aes(x = beach, y = cultural, label = label), 
            vjust = -1.5, size = 6, fontface = "bold") +
  labs(title = "Productive Efficiency: On vs. Inside PPF",
       subtitle = "Efficient points (A, C) on frontier; Inefficient point (B) inside",
       x = "Beach Tourism",
       y = "Cultural Tourism") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Efficient (on PPF)" = "#27ae60", 
                                 "Inefficient (inside)" = "#e74c3c")) +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  annotate("rect", xmin = 0, xmax = 100, ymin = 0, ymax = 100, 
           alpha = 0, color = "black", size = 0.5)
```

## Causes of Inefficiency

**Why might an economy operate inside the PPF?**

1. **Unemployment**
   - Workers without jobs
   - Idle factories and equipment
   - Example: COVID-19 lockdowns (2020)

2. **Resource Misallocation**
   - Resources in wrong industries
   - Poor management decisions
   - Example: Over-investment in declining sectors

3. **Lack of Technology Adoption**
   - Not using best available methods
   - Example: Manual booking when online systems exist

## Real Example: Portugal Tourism Efficiency

**Portugal Tourism 2019 vs 2020-2021**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Portugal tourism data with efficiency
portugal_efficiency <- data.frame(
  year = c(2019, 2020, 2021, 2022, 2023, 2024),
  actual_output = c(100, 39, 59, 98, 113, 117),  # Index: 2019 = 100
  potential_output = c(100, 95, 97, 105, 108, 112),  # Adjusted for capacity
  status = c("Efficient", "Major Inefficiency", "Recovering", 
             "Near-Efficient", "Efficient", "Efficient")
)

ggplot(portugal_efficiency, aes(x = year)) +
  geom_line(aes(y = potential_output, color = "Potential (PPF)"), 
            size = 1.5, linetype = "dashed") +
  geom_line(aes(y = actual_output, color = "Actual Output"), 
            size = 1.5) +
  geom_point(aes(y = potential_output), size = 4, color = "#3498db") +
  geom_point(aes(y = actual_output, shape = status, color = "Actual Output"), size = 5) +
  labs(title = "Portugal Tourism: Efficiency Analysis 2019-2024",
       subtitle = "COVID-19 pushed economy inside PPF (inefficient); recovery moved back to frontier",
       x = "Year",
       y = "Tourism Output (Index: 2019 = 100)",
       color = "",
       shape = "Status") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Potential (PPF)" = "#3498db", 
                                 "Actual Output" = "#e74c3c")) +
  scale_shape_manual(values = c("Efficient" = 16, 
                                 "Major Inefficiency" = 4,
                                 "Recovering" = 17,
                                 "Near-Efficient" = 15)) +
  theme(legend.position = "bottom") +
  annotate("rect", xmin = 2019.8, xmax = 2021.2, ymin = 0, ymax = 120, 
           alpha = 0.1, fill = "red") +
  annotate("text", x = 2020.5, y = 75, 
           label = "Inside PPF:\nUnemployment,\nClosed facilities", 
           size = 4, color = "#e74c3c", fontface = "bold")
```

## Economic Growth: Shifting the PPF

**PPF can shift outward due to**:

1. **More Resources**
   - Population growth → more labor
   - Discovery of new natural resources
   - Capital accumulation (investment)

2. **Better Technology**
   - Innovation and technological progress
   - Improved productivity

3. **Better Institutions**
   - Rule of law, property rights
   - Efficient markets

## Economic Growth Visualization

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Original and shifted PPF
beach_orig <- seq(0, 100, by = 1)
cultural_orig <- 100 * (1 - (beach_orig/100)^0.7)

beach_new <- seq(0, 130, by = 1)
cultural_new <- 130 * (1 - (beach_new/130)^0.7)

ppf_growth <- data.frame(
  beach = c(beach_orig, beach_new),
  cultural = c(cultural_orig, cultural_new),
  period = rep(c("2024 (Current PPF)", "2030 (After Growth)"), 
               c(length(beach_orig), length(beach_new)))
)

# Sample points
growth_points <- data.frame(
  beach = c(60, 80),
  cultural = c(57, 74),
  period = c("2024 (Current PPF)", "2030 (After Growth)"),
  label = c("Point A (2024)", "Point B (2030)")
)

ggplot(ppf_growth, aes(x = beach, y = cultural, color = period)) +
  geom_line(size = 2) +
  geom_point(data = growth_points, aes(x = beach, y = cultural), size = 7) +
  geom_text(data = growth_points, aes(x = beach, y = cultural, label = label), 
            vjust = -1.5, size = 5, fontface = "bold", color = "black") +
  labs(title = "Economic Growth: PPF Shifts Outward",
       subtitle = "With more resources & technology, can produce MORE of BOTH goods",
       x = "Beach Tourism Capacity",
       y = "Cultural Tourism Capacity",
       color = "") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("2024 (Current PPF)" = "#e74c3c", 
                                 "2030 (After Growth)" = "#27ae60")) +
  theme(legend.position = "bottom") +
  annotate("segment", x = 60, xend = 80, y = 57, yend = 74,
           arrow = arrow(length = unit(0.4, "cm"), type = "closed"),
           size = 1.5, color = "#3498db", linetype = "dashed") +
  annotate("text", x = 70, y = 62, 
           label = "Growth allows\nmore of both", 
           size = 4.5, color = "#3498db", fontface = "bold")
```

**Key**: Economic growth makes previously impossible combinations achievable!

## Investment vs. Consumption Trade-off

**Classic PPF Application**: Choose between current consumption and investment

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Two countries with different investment choices
consumption <- seq(0, 100, by = 1)
investment <- 100 * (1 - (consumption/100)^0.8)

countries <- data.frame(consumption, investment)

# Country choices
country_choices <- data.frame(
  consumption = c(70, 40),
  investment = c(47, 73),
  country = c("Country A:\nHigh Consumption", "Country B:\nHigh Investment"),
  future_growth = c(1.15, 1.35)
)

ggplot(countries, aes(x = consumption, y = investment)) +
  geom_line(size = 2, color = "#3498db") +
  geom_point(data = country_choices, aes(x = consumption, y = investment, 
                                          color = country), size = 8) +
  geom_text(data = country_choices, aes(x = consumption, y = investment, label = country), 
            vjust = -1.8, size = 4.5, fontface = "bold") +
  labs(title = "Current Consumption vs. Future Investment Trade-off",
       subtitle = "Country B sacrifices consumption now → more investment → faster future growth",
       x = "Current Consumption",
       y = "Investment (Capital Goods)") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Country A:\nHigh Consumption" = "#e74c3c", 
                                 "Country B:\nHigh Investment" = "#27ae60")) +
  theme(legend.position = "none") +
  annotate("text", x = 70, y = 35, 
           label = "→ Slower future growth\n(Small PPF shift)", 
           size = 4, color = "#e74c3c") +
  annotate("text", x = 40, y = 60, 
           label = "→ Faster future growth\n(Large PPF shift)", 
           size = 4, color = "#27ae60")
```

## Tourism Application: Investment Choice

**Portugal's Tourism Investment Strategy (2020-2024)**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Investment vs immediate revenue
investment_choice <- data.frame(
  category = c("Infrastructure\nInvestment", "Marketing\n(Quick Returns)", "Sustainability\nProjects", "Digital\nTransformation"),
  immediate_revenue = c(0, 85, 15, 30),
  long_term_growth = c(95, 40, 70, 80),
  chosen_2020_2024 = c(280, 120, 150, 180)
)

library(tidyr)
investment_long <- investment_choice %>%
  select(category, immediate_revenue, long_term_growth) %>%
  pivot_longer(cols = c(immediate_revenue, long_term_growth),
               names_to = "impact_type",
               values_to = "score")

investment_long$impact_type <- factor(investment_long$impact_type,
                                       levels = c("immediate_revenue", "long_term_growth"),
                                       labels = c("Immediate Revenue Impact", 
                                                 "Long-term Growth Potential"))

ggplot(investment_long, aes(x = category, y = score, fill = impact_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Tourism Investment Categories: Trade-off Analysis",
       subtitle = "Infrastructure & Digital have high long-term impact but low immediate returns",
       x = "",
       y = "Impact Score (0-100)",
       fill = "") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("Immediate Revenue Impact" = "#e74c3c", 
                                "Long-term Growth Potential" = "#27ae60")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 11)) +
  geom_text(aes(label = score), position = position_dodge(width = 0.9), 
            vjust = -0.5, size = 4)
```

**Strategic choice**: Prioritize long-term growth (infrastructure, digital) despite lower immediate returns

## Technological Change: Example

**Digital Transformation in Tourism (2015-2025)**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Technology shifts PPF for booking services
bookings_traditional <- seq(0, 100, by = 1)
customer_service_trad <- 100 * (1 - (bookings_traditional/100)^0.75)

bookings_digital <- seq(0, 150, by = 1)
customer_service_digital <- 150 * (1 - (bookings_digital/150)^0.75)

tech_ppf <- data.frame(
  bookings = c(bookings_traditional, bookings_digital),
  customer_service = c(customer_service_trad, customer_service_digital),
  era = rep(c("2015: Manual Systems", "2025: Digital + AI"), 
            c(length(bookings_traditional), length(bookings_digital)))
)

ggplot(tech_ppf, aes(x = bookings, y = customer_service, color = era)) +
  geom_line(size = 2) +
  labs(title = "Technology Impact: Travel Agency PPF Shift",
       subtitle = "Digital transformation + AI enables more bookings AND better customer service",
       x = "Number of Bookings Processed (per day)",
       y = "Quality of Customer Service (index)",
       color = "") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("2015: Manual Systems" = "#e74c3c", 
                                 "2025: Digital + AI" = "#27ae60")) +
  theme(legend.position = "bottom") +
  annotate("segment", x = 60, xend = 90, y = 60, yend = 90,
           arrow = arrow(length = unit(0.5, "cm"), type = "closed"),
           size = 2, color = "#3498db") +
  annotate("text", x = 75, y = 72, 
           label = "Technology\nShift", 
           size = 6, color = "#3498db", fontface = "bold",
           angle = 30)
```

## PPF Summary: Key Insights

1. **Shows maximum production** given resources & technology
2. **Points on PPF** = efficient (can't produce more without trade-off)
3. **Points inside PPF** = inefficient (unemployment, waste)
4. **Points outside PPF** = impossible (currently unattainable)
5. **Shape (bowed)** = increasing opportunity costs
6. **Moving along PPF** = reallocating resources (opportunity cost)
7. **Shifting PPF outward** = economic growth (more resources, better technology)
8. **Investment trade-off** = consumption today vs. growth tomorrow

## Real-World Applications

**PPF thinking helps answer**:

- Should Portugal build more hotels or preserve more heritage sites?
- How much should airlines invest in fuel efficiency vs. passenger comfort?
- Trade-off between mass tourism revenue and sustainable tourism?
- How much GDP to allocate to tourism vs. other sectors?

**All involve opportunity costs and efficiency considerations!**

## Summary: Complete Picture

**Today's Lecture Integration**:

1. **Production Factors** (Land, Labor, Capital) are scarce inputs
2. **Scarcity** forces choices between alternatives
3. **PPF** visualizes society's production possibilities
4. **Opportunity Cost** = moving along PPF (what's foregone)
5. **Efficiency** = operating on (not inside) PPF
6. **Economic Growth** = PPF shifts outward
7. **Investment** today enables faster growth tomorrow

**This connects to**: Lecture 3 (opportunity cost), Lecture 2 (WHAT/HOW/FOR WHOM), Lecture 1 (scarcity & efficiency)

## Exercises {.center}

**Application Time!**

PPF and opportunity cost calculations.

## Exercise 1: Multiple Choice

**Question**: An economy is producing at a point **inside** its PPF. This indicates:

A. The economy is efficiently using all resources  
B. The economy has economic growth  
C. There is unemployment or resource misallocation  
D. It's impossible to produce more of any good

::: {.fragment}
**Answer: C**

Explanation: Inside PPF means inefficiency – resources are unemployed or misallocated. Could produce more of at least one good without sacrificing the other (move toward frontier).
:::

## Exercise 2: Multiple Choice

**Question**: A country's PPF shifts outward. This could be caused by:

A. Higher unemployment  
B. Decrease in the labor force  
C. Technological innovation  
D. Producing less capital goods

::: {.fragment}
**Answer: C**

Explanation: Outward PPF shift = economic growth. Caused by: more resources, better technology, improved productivity. Options A, B, D would shift PPF inward or cause movement inside PPF.
:::

## Exercise 3: Open Question - Part 1

**Scenario**: The Portuguese Algarve region has limited land and labor. It can allocate resources between golf tourism and beach tourism. Current production possibilities:

| Combination | Golf Resorts | Beach Hotels |
|-------------|-------------|--------------|
| A | 0 | 50 |
| B | 5 | 48 |
| C | 10 | 44 |
| D | 15 | 38 |
| E | 20 | 30 |
| F | 25 | 0 |

## Exercise 3: Open Question - Part 2

**Questions**:

a) Draw the PPF with Golf Resorts on x-axis, Beach Hotels on y-axis

b) Calculate the opportunity cost of moving from:
   - Point B to Point C
   - Point D to Point E

c) Does the PPF exhibit increasing opportunity costs? Explain.

d) If region currently operates at point U (12 golf, 30 beach), what does this indicate?

e) A new tourism development technology is invented. How would this affect the PPF? Draw the new curve.

## Exercise 3: Solution - Part a & b

**a) PPF Graph**: 

Plots points A through F with Golf (x-axis: 0-25) and Beach (y-axis: 0-50). Curve is bowed outward (concave).

**b) Opportunity Cost Calculations**:

**B → C** (5 golf → 10 golf):
$$\text{Gain in Golf} = 10 - 5 = 5 \text{ resorts}$$
$$\text{Loss in Beach} = 48 - 44 = 4 \text{ hotels}$$
$$\text{Opportunity Cost} = \frac{4 \text{ hotels}}{5 \text{ resorts}} = 0.8 \text{ beach hotels per golf resort}$$

**D → E** (15 golf → 20 golf):
$$\text{Gain in Golf} = 20 - 15 = 5 \text{ resorts}$$
$$\text{Loss in Beach} = 38 - 30 = 8 \text{ hotels}$$
$$\text{Opportunity Cost} = \frac{8 \text{ hotels}}{5 \text{ resorts}} = 1.6 \text{ beach hotels per golf resort}$$

## Exercise 3: Solution - Part c, d, e

**c) Increasing Opportunity Costs?**

**YES!** Opportunity cost increases from 0.8 (B→C) to 1.6 (D→E).

**Explanation**: As more resources shift to golf, we use resources less suited for golf (originally better for beaches). Each additional golf resort requires sacrificing more beach hotels.

**d) Point U (12 golf, 30 beach)**:

This is **INSIDE the PPF** (compare to point D: 15 golf, 38 beach).

**Indicates**: 

- **Inefficiency** (unemployment or resource misallocation)
- Could produce more golf without reducing beach (move to D: 15 golf, 38 beach)
- Could produce more beach without reducing golf (move toward 12 golf, ~43 beach)

## Exercise 3: Solution - Part e

**e) New Technology Effect**:

**Technology shifts PPF OUTWARD** (economic growth).

New PPF might be:

| Combination | Golf Resorts | Beach Hotels |
|-------------|-------------|--------------|
| A' | 0 | 60 (+10) |
| C' | 12 | 55 (+11) |
| E' | 24 | 38 (+8) |
| F' | 30 | 0 (+5) |

**Effect**: Can now produce MORE of both types of tourism with same resources!

**Graph**: Draw original PPF, then new PPF further from origin (parallel or biased depending on where technology applies).

## Next Lecture

**February 19, 2026**: Doubts and Exercises Revision

**Preparation for Test 1 (February 20)**:

- Review Lectures 1-4
- Practice opportunity cost calculations
- Understand PPF shifts vs. movements
- Know production factors and economic systems

**Test 1 covers**: All Fundamentals material (15% of final grade)

## Thank You! {.center}

**Questions?**

*Contact*: paulo.fagandini@ext.universidadeeuropeia.pt

---

*Next class: Wednesday, February 19, 2026 (Review Session)*

*Test 1: Thursday, February 20, 2026*

::: {.notes}
This completes the Fundamentals block. Students should understand: scarcity → trade-offs → opportunity cost → PPF → efficiency. These concepts are foundational for all future topics.
:::