---
title: "Producer Theory"
subtitle: "Lecture 16: Short Run & Long Run Market Equilibrium"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    from: markdown+emoji
    theme: [default, custom.scss]
    transition: slide
    slide-number: true
    chalkboard: false
    scrollable: false
    footer: "Economics of Tourism | Lecture 16"
    logo: ../images/logo.png
    width: 1600
    height: 900
    embed-resources: true
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyverse)
library(patchwork)

ue_red <- "#C8102E"
ue_black <- "#1D1D1B"
ue_gray <- "#6D6E71"
ue_white <- "#FFFFFF"
ue_blue <- "#2E86AB"
ue_green <- "#27ae60"
ue_orange <- "#E67E22"
```

## Recap: Where We Are :rewind:

<div style="font-size: 78%;">

Over Lectures 5–15, we built **both sides** of the market:

:::: {.columns}
::: {.column width="50%"}
:shopping_cart: **Demand side** (L5–L9):

- Utility maximization → demand curve
- Market demand (horizontal sum)
- Consumer surplus
- Demand elasticity
:::

::: {.column width="50%"}
:factory: **Supply side** (L10–L15):

- Cost structure → $P = MC$ → supply curve
- Market supply (horizontal sum)
- Producer surplus
- Supply elasticity
:::
::::

. . .

:dart: **Today**: We put them **together** to find the market equilibrium — the price and quantity where buyers and sellers agree. This is the **payoff** of the entire course so far!

</div>

# Short-Run Market Equilibrium

## Where Supply Meets Demand :handshake:

<div style="font-size: 78%;">

::: {.key-concept}
**Market Equilibrium**

The price and quantity at which **quantity demanded equals quantity supplied**:

$$Q_d(P^*) = Q_s(P^*)$$

At the equilibrium price $P^*$, every buyer who wants to buy at that price can, and every seller who wants to sell can. **No one is left unsatisfied.**
:::

**Finding equilibrium algebraically** (linear curves):

Given $D: P = 50 - 0.5Q$ and $S: P = 10 + 0.25Q$

Set equal: $50 - 0.5Q = 10 + 0.25Q$ → $40 = 0.75Q$ → $Q^* = 53.3$ → $P^* = €23.3$

:point_right: At this price, the quantity buyers want to buy **exactly matches** the quantity sellers want to sell.

</div>

## Equilibrium Graphically :bar_chart:

```{r}
#| echo: false
#| warning: false
#| fig-width: 11
#| fig-height: 5.5
#| fig-align: center

# D: P = 50 - 0.5Q, S: P = 10 + 0.25Q
# Eq: Q* = 53.33, P* = 23.33
Q_eq <- 40/0.75
P_eq <- 50 - 0.5 * Q_eq

df_d <- data.frame(Q = seq(0, 100, by = 1)) %>% mutate(P = 50 - 0.5 * Q)
df_s <- data.frame(Q = seq(0, 100, by = 1)) %>% mutate(P = 10 + 0.25 * Q)

ggplot() +
  # CS triangle
  geom_polygon(data = data.frame(x = c(0, Q_eq, 0), y = c(50, P_eq, P_eq)),
               aes(x = x, y = y), fill = ue_blue, alpha = 0.12) +
  # PS triangle
  geom_polygon(data = data.frame(x = c(0, Q_eq, 0), y = c(10, P_eq, P_eq)),
               aes(x = x, y = y), fill = ue_red, alpha = 0.12) +
  # Curves
  geom_line(data = df_d %>% filter(P >= 0), aes(x = Q, y = P),
            color = ue_blue, linewidth = 1.5) +
  geom_line(data = df_s, aes(x = Q, y = P), color = ue_red, linewidth = 1.5) +
  # Equilibrium
  geom_point(aes(x = Q_eq, y = P_eq), size = 4, color = ue_black) +
  geom_segment(aes(x = Q_eq, y = 0, xend = Q_eq, yend = P_eq),
               linetype = "dashed", color = ue_gray) +
  geom_segment(aes(x = 0, y = P_eq, xend = Q_eq, yend = P_eq),
               linetype = "dashed", color = ue_gray) +
  # Labels
  annotate("text", x = 18, y = 38, label = "CS", color = ue_blue,
           fontface = "bold", size = 5) +
  annotate("text", x = 18, y = 16, label = "PS", color = ue_red,
           fontface = "bold", size = 5) +
  annotate("text", x = 90, y = 8, label = "D", color = ue_blue,
           fontface = "bold.italic", size = 6) +
  annotate("text", x = 90, y = 35, label = "S", color = ue_red,
           fontface = "bold.italic", size = 6) +
  annotate("text", x = Q_eq + 2, y = P_eq + 2.5, label = "E",
           fontface = "bold", size = 5) +
  annotate("text", x = Q_eq, y = -2, label = paste0("Q* ≈ ", round(Q_eq, 1)),
           fontface = "bold", size = 3.8) +
  annotate("text", x = -4, y = P_eq, label = paste0("P* ≈ ", round(P_eq, 1)),
           fontface = "bold", size = 3.8) +
  coord_cartesian(ylim = c(0, 55), xlim = c(0, 105)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Quantity (Q)", y = "Price (€)",
       title = "Market Equilibrium with CS and PS (Hypothetical)") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(color = ue_red, face = "bold", hjust = 0.5, size = 15),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## What Happens Away from Equilibrium? :warning:

<div style="font-size: 76%;">

:::: {.columns}
::: {.column width="50%"}
**Price too HIGH** ($P > P^*$): **Surplus** (excess supply)

- $Q_s > Q_d$ → goods pile up
- Sellers compete by **lowering prices**
- Price falls toward $P^*$

:hotel: *Tourism*: Hotels with too many empty rooms in off-season → prices drop on Booking.com
:::

::: {.column width="50%"}
**Price too LOW** ($P < P^*$): **Shortage** (excess demand)

- $Q_d > Q_s$ → not enough goods
- Buyers compete by **bidding prices up**
- Price rises toward $P^*$

:airplane: *Tourism*: Flight demand exceeds seats during holidays → prices rise, last-minute fares skyrocket
:::
::::

. . .

::: {.key-concept}
**Self-Correcting Mechanism**

Markets **naturally tend toward equilibrium**. Surpluses push prices down; shortages push prices up. This process continues until $Q_d = Q_s$.
:::

</div>

## The Four Shift Rules :arrows_counterclockwise:

<div style="font-size: 68%;">

```{r}
#| echo: false
#| warning: false
#| fig-width: 14
#| fig-height: 5
#| fig-align: center

make_shift_panel <- function(shift_type, title_str) {
  # Base: D: P = 50 - 0.5Q, S: P = 10 + 0.25Q, Eq: Q=53.3, P=23.3
  d1 <- function(Q) 50 - 0.5 * Q
  s1 <- function(Q) 10 + 0.25 * Q
  
  if (shift_type == "D_up") {
    d2 <- function(Q) 65 - 0.5 * Q  # demand shifts right
    s2 <- s1
    # New eq: 65 - 0.5Q = 10 + 0.25Q => 55 = 0.75Q => Q=73.3, P=28.3
    d2_col <- ue_blue; s2_col <- ue_red
    arrow_x <- 60; arrow_y <- 20
    d_label <- expression(D[2]); d_old_label <- expression(D[1])
    result <- "P↑ Q↑"
  } else if (shift_type == "D_down") {
    d2 <- function(Q) 35 - 0.5 * Q
    s2 <- s1
    d2_col <- ue_blue; s2_col <- ue_red
    result <- "P↓ Q↓"
  } else if (shift_type == "S_up") {
    d2 <- d1
    s2 <- function(Q) 10 + 0.25 * Q + 10  # supply shifts left (up)
    d2_col <- ue_blue; s2_col <- ue_red
    result <- "P↑ Q↓"
  } else {  # S_down
    d2 <- d1
    s2 <- function(Q) 10 + 0.25 * Q - 8  # supply shifts right (down)
    d2_col <- ue_blue; s2_col <- ue_red
    result <- "P↓ Q↑"
  }
  
  Q_seq <- seq(0, 100, by = 1)
  
  p <- ggplot() +
    # Old curves (gray)
    geom_line(data = data.frame(Q = Q_seq, P = d1(Q_seq)) %>% filter(P >= 0),
              aes(x = Q, y = P), color = ue_gray, linewidth = 0.8, linetype = "dashed") +
    geom_line(data = data.frame(Q = Q_seq, P = s1(Q_seq)),
              aes(x = Q, y = P), color = ue_gray, linewidth = 0.8, linetype = "dashed") +
    # New curves (bold)
    geom_line(data = data.frame(Q = Q_seq, P = d2(Q_seq)) %>% filter(P >= 0),
              aes(x = Q, y = P), color = ue_blue, linewidth = 1.3) +
    geom_line(data = data.frame(Q = Q_seq, P = s2(Q_seq)),
              aes(x = Q, y = P), color = ue_red, linewidth = 1.3)
  
  # Old equilibrium
  # Solve d1 = s1: 50 - 0.5Q = 10 + 0.25Q => Q = 53.3
  q_old <- 40 / 0.75; p_old <- d1(q_old)
  p <- p + geom_point(aes(x = q_old, y = p_old), size = 2, color = ue_gray)
  
  # New equilibrium (solve d2 = s2 numerically)
  # Find intersection
  f_diff <- function(Q) d2(Q) - s2(Q)
  q_new <- uniroot(f_diff, c(0, 150))$root
  p_new <- d2(q_new)
  p <- p +
    geom_point(aes(x = q_new, y = p_new), size = 3, color = ue_black) +
    annotate("text", x = q_new + 4, y = p_new + 3,
             label = result, fontface = "bold", size = 4, color = ue_black)
  
  p <- p +
    coord_cartesian(ylim = c(0, 55), xlim = c(0, 100)) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x = "Q", y = "P", title = title_str) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(color = ue_red, face = "bold", hjust = 0.5, size = 11),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
  return(p)
}

p1 <- make_shift_panel("D_up", "(a) Demand ↑: P↑ Q↑")
p2 <- make_shift_panel("D_down", "(b) Demand ↓: P↓ Q↓")
p3 <- make_shift_panel("S_up", "(c) Supply ↓: P↑ Q↓")
p4 <- make_shift_panel("S_down", "(d) Supply ↑: P↓ Q↑")

(p1 + p2) / (p3 + p4) +
  plot_annotation(title = "The Four Shift Rules",
                  theme = theme(plot.title = element_text(color = ue_red,
                                face = "bold", hjust = 0.5, size = 14)))
```

</div>

## The Four Rules Summarized :clipboard:

<div style="font-size: 78%;">

| Shift | Price | Quantity | Tourism Example |
|---|:---:|:---:|---|
| **Demand ↑** (right) | ↑ | ↑ | Summer season → more tourists want hotels |
| **Demand ↓** (left) | ↓ | ↓ | Pandemic → fewer tourists travel |
| **Supply ↑** (right) | ↓ | ↑ | New technology → cheaper flights |
| **Supply ↓** (left) | ↑ | ↓ | Fuel crisis → airline costs rise |

. . .

:warning: **When both curves shift simultaneously**, the direction of **one** variable (P or Q) becomes **ambiguous** — it depends on the **relative size** of the shifts.

**Example**: Summer in the Algarve — demand shifts right (tourists) AND supply shifts right (seasonal workers). Quantity definitely rises, but price? Depends on which shift is bigger!

</div>

## Welfare at Equilibrium :star:

<div style="font-size: 78%;">

::: {.key-concept}
**Total Economic Welfare**

$$W = CS + PS$$

At the competitive equilibrium, total welfare is **maximized**. Any other price/quantity combination would reduce the combined surplus.
:::

From our example ($D: P = 50 - 0.5Q$, $S: P = 10 + 0.25Q$, $Q^* \approx 53.3$, $P^* \approx 23.3$):

$$CS = \frac{1}{2} \times 53.3 \times (50 - 23.3) = \frac{1}{2} \times 53.3 \times 26.7 \approx €711$$

$$PS = \frac{1}{2} \times 53.3 \times (23.3 - 10) = \frac{1}{2} \times 53.3 \times 13.3 \approx €356$$

$$W = 711 + 356 \approx €1{,}067$$

:point_right: Any **tax**, **price ceiling**, or **price floor** would reduce total welfare by creating **deadweight loss**.

</div>

# Long-Run Market Equilibrium

## The Short Run vs Long Run Revisited :hourglass:

<div style="font-size: 78%;">

:::: {.columns}
::: {.column width="50%"}
:lock: **Short-run equilibrium**:

- Number of firms is **fixed**
- Firms may earn **positive profit**, **zero profit**, or **losses**
- Some factors of production are fixed
- This is what we've been analyzing!
:::

::: {.column width="50%"}
:unlock: **Long-run equilibrium**:

- Firms can **enter** or **exit** the market
- All factors of production are variable
- A special condition emerges: **zero economic profit**
- The market adjusts to a new, more fundamental equilibrium
:::
::::

. . .

::: {.key-concept}
**The Key Long-Run Question**

What happens over time when firms in a competitive market are earning positive profits? Or making losses?
:::

</div>

## The Entry and Exit Mechanism :door:

<div style="font-size: 76%;">

:::: {.columns}
::: {.column width="50%"}

**If firms earn positive economic profit** ($P > ATC$):

1. :arrow_right: New firms are attracted → they **enter** the market
2. :chart_with_upwards_trend: Market supply shifts **right** (more firms = more supply)
3. :arrow_down: Equilibrium price **falls**
4. :repeat: Process continues until $P = ATC_{\min}$ → **zero economic profit**

:::

::: {.column width="50%"}

**If firms make losses** ($P < ATC$):

1. :arrow_left: Some firms **exit** the market
2. :chart_with_downwards_trend: Market supply shifts **left** (fewer firms = less supply)
3. :arrow_up: Equilibrium price **rises**
4. :repeat: Process continues until $P = ATC_{\min}$ → **zero economic profit**
:::
::::

. . .

::: {.key-concept}
**Long-Run Equilibrium Condition**

In a perfectly competitive market: $P = MC = ATC_{\min}$

Firms produce at minimum average cost and earn **zero economic profit**.
:::

</div>

## Visualizing the Long-Run Adjustment :cinema:

<div style="font-size: 68%;">

```{r}
#| echo: false
#| warning: false
#| fig-width: 14
#| fig-height: 5
#| fig-align: center

# Left: Market with profit -> entry -> new equilibrium
# Right: Individual firm goes from profit to zero profit

# Market panel
df_d <- data.frame(Q = seq(0, 200, by = 1)) %>% mutate(P = 50 - 0.15 * Q)
df_s1 <- data.frame(Q = seq(0, 200, by = 1)) %>% mutate(P = 10 + 0.15 * Q)
df_s2 <- data.frame(Q = seq(0, 200, by = 1)) %>% mutate(P = 5 + 0.15 * Q)
# S1 eq: 50 - 0.15Q = 10 + 0.15Q => 40 = 0.3Q => Q=133.3, P=30
# S2 eq: 50 - 0.15Q = 5 + 0.15Q => 45 = 0.3Q => Q=150, P=27.5

p_market <- ggplot() +
  geom_line(data = df_d %>% filter(P >= 0), aes(x = Q, y = P),
            color = ue_blue, linewidth = 1.3) +
  geom_line(data = df_s1, aes(x = Q, y = P),
            color = ue_gray, linewidth = 1, linetype = "dashed") +
  geom_line(data = df_s2, aes(x = Q, y = P),
            color = ue_red, linewidth = 1.3) +
  # Old eq
  geom_point(aes(x = 133.3, y = 30), size = 3, color = ue_gray) +
  # New eq
  geom_point(aes(x = 150, y = 27.5), size = 3, color = ue_black) +
  # Arrow
  annotate("segment", x = 133.3, y = 30, xend = 148, yend = 27.8,
           arrow = arrow(length = unit(0.2, "cm")), color = ue_black, linewidth = 0.8) +
  annotate("text", x = 175, y = 45, label = "D", color = ue_blue,
           fontface = "bold.italic", size = 5) +
  annotate("text", x = 175, y = 22, label = expression(S[1]),
           color = ue_gray, fontface = "italic", size = 4) +
  annotate("text", x = 175, y = 32, label = expression(S[2] ~ "(entry)"),
           color = ue_red, fontface = "bold", size = 4) +
  annotate("text", x = 80, y = 5,
           label = "Entry → Supply shifts right → Price falls",
           color = ue_black, fontface = "italic", size = 3.5) +
  coord_cartesian(ylim = c(0, 55), xlim = c(0, 200)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Market Quantity", y = "Price (€)",
       title = "(a) Market: Entry Shifts Supply Right") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(color = ue_red, face = "bold", hjust = 0.5, size = 12),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Firm panel: MC and ATC, two price levels
Q_f <- seq(1, 55, by = 0.5)
MC_f <- 0.06 * Q_f^2 - 2.8 * Q_f + 60
ATC_f <- 0.02 * Q_f^3 / Q_f - 1.4 * Q_f + 60 + 500 / Q_f
# Simplify: ATC = 0.02*Q^2 - 1.4*Q + 60 + 500/Q

p_firm <- ggplot() +
  geom_line(data = data.frame(Q = Q_f, MC = MC_f),
            aes(x = Q, y = MC), color = ue_red, linewidth = 1.2) +
  geom_line(data = data.frame(Q = Q_f, ATC = ATC_f),
            aes(x = Q, y = ATC), color = ue_black, linewidth = 1) +
  # High price (SR with profit)
  geom_hline(yintercept = 60, color = ue_green, linewidth = 0.9, linetype = "dashed") +
  # LR price = ATC min ≈ 48.38
  geom_hline(yintercept = 48.38, color = ue_orange, linewidth = 1.2) +
  # Profit rectangle at high price (approx)
  annotate("rect", xmin = 0, xmax = 46.7, ymin = 48.38, ymax = 60,
           fill = ue_green, alpha = 0.1) +
  # Labels
  annotate("text", x = 52, y = 62, label = expression(P[SR] ~ "(profit)"),
           color = ue_green, fontface = "bold", size = 3.5) +
  annotate("text", x = 52, y = 46, label = expression(P[LR] == ATC[min]),
           color = ue_orange, fontface = "bold", size = 3.5) +
  annotate("text", x = 52, y = 72, label = "MC", color = ue_red,
           fontface = "bold.italic", size = 4.5) +
  annotate("text", x = 52, y = 52, label = "ATC", color = ue_black,
           fontface = "italic", size = 4) +
  annotate("text", x = 20, y = 54, label = "Profit\n(SR)", color = ue_green,
           fontface = "bold", size = 3.5) +
  annotate("segment", x = 25, y = 57, xend = 25, yend = 49,
           arrow = arrow(length = unit(0.15, "cm"), ends = "both"),
           color = ue_green, linewidth = 0.5) +
  coord_cartesian(ylim = c(25, 80), xlim = c(0, 55)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Firm Quantity (q)", y = "€/unit",
       title = "(b) Firm: Profit Attracts Entry → Price Falls to ATC min") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(color = ue_red, face = "bold", hjust = 0.5, size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p_market + p_firm
```

**(a)** Positive profits attract **entry** → supply shifts right → price falls. **(b)** Price falls until $P = ATC_{\min}$ → zero economic profit → **entry stops**.

</div>

## Zero Profit ≠ Failure! :bulb:

<div style="font-size: 78%;">

**"Zero economic profit"** sounds bad — but remember what it means:

::: {.key-concept}
**Zero Economic Profit**

The firm covers **all** costs, including the **opportunity cost** of the owner's time and capital. The owner earns exactly what they could earn in their **best alternative** — a **normal return**.

$$\text{Economic Profit} = 0 \implies \text{Accounting Profit} = \text{Normal Return}$$
:::

:::: {.columns}
::: {.column width="50%"}
:white_check_mark: **What zero economic profit means**:

- The business is sustainable
- The owner has no reason to leave
- No new firms want to enter
- The market is in **long-run rest**
:::

::: {.column width="50%"}
:x: **What it does NOT mean**:

- The owner earns no money
- The business is failing
- The owner should close
- There are no accounting profits
:::
::::

</div>

## The Long-Run Supply Curve :chart_with_upwards_trend:

<div style="font-size: 76%;">

If all firms have the **same cost structure** (constant-cost industry):

::: {.key-concept}
**Long-Run Supply in a Constant-Cost Industry**

The long-run supply curve is **horizontal** at $P = ATC_{\min}$.

- If demand increases → new firms enter, supply shifts right, price returns to $ATC_{\min}$
- If demand decreases → firms exit, supply shifts left, price returns to $ATC_{\min}$
- **Quantity adjusts**, but **price stays the same** in the long run!
:::

:point_right: In reality, costs may **rise** as the industry grows (increasing-cost industry → upward-sloping LR supply) or **fall** (decreasing-cost industry → downward-sloping LR supply). But the constant-cost case is the baseline.

**Tourism**: The long-run supply of standardized tour packages may be nearly horizontal — if tour operators can easily enter/exit and all face similar costs, the long-run price settles at the minimum cost of providing a tour.

</div>

# Tourism Application

## Tourism Markets: SR vs LR :airplane:

<div style="font-size: 72%;">

**What happened to Lisbon's hotel market after the tourism boom (post-2015)?**

:::: {.columns}
::: {.column width="50%"}
:one: **Short run** (2015–2017):

- Demand surged (low-cost airlines, Lisbon's popularity)
- Supply was **fixed** (existing hotels)
- Result: **prices skyrocketed**, occupancy hit record highs
- Existing hotels earned **large profits**
- $P >> ATC_{\min}$ → well above breakeven
:::

::: {.column width="50%"}
:two: **Long run** (2018–2024):

- Profits attracted **entry**: new hotels, Airbnb, hostels
- Supply shifted **right** significantly
- Result: prices stabilized (or even fell for some segments)
- Competition increased, profit margins **narrowed**
- Market moved toward $P \approx ATC_{\min}$

:point_right: This is the **entry mechanism** in action!
:::
::::

. . .

:bulb: The pandemic (2020–2021) reversed this: demand collapsed → firms exited → supply shifted left → the cycle restarted.

</div>

## Summary :memo:

<div style="font-size: 70%;">

**Today's Key Takeaways**:

1. **SR equilibrium**: where $Q_d = Q_s$. Price too high → surplus (price falls). Price too low → shortage (price rises).
2. **Four shift rules**: D↑ → P↑Q↑; D↓ → P↓Q↓; S↑ → P↓Q↑; S↓ → P↑Q↓
3. **Simultaneous shifts**: direction of one variable is ambiguous — depends on magnitude
4. **Total welfare** = CS + PS, maximized at competitive equilibrium
5. **LR equilibrium**: entry (if profit > 0) and exit (if profit < 0) drive $P → ATC_{\min}$
6. **Zero economic profit** in the LR = normal return = sustainable. It's not failure!
7. **LR supply** (constant-cost): horizontal at $P = ATC_{\min}$

**This completes the Producer Theory block!** :tada:

Lectures 10–16: Costs → Profit max → Shutdown → Supply rule → PS & Market supply → Elasticity → Equilibrium

**Next (Lecture 17, April 17)**: Introduction to Game Theory — Prisoner's Dilemma & Nash Equilibrium :chess_pawn:

</div>

## Exercises {.center}

**Practice Time!** :pencil2:

Market equilibrium, shifts, and long-run adjustment.

## Exercise 1: Multiple Choice

<div style="font-size: 82%;">

**Question**: In the Algarve hotel market, a new low-cost airline begins flying to Faro from Northern Europe. At the same time, the Portuguese government introduces a new tourist tax on hotel stays. What is the effect on the equilibrium **quantity** of hotel room-nights?

A. Quantity definitely increases

B. Quantity definitely decreases

C. Quantity definitely stays the same

D. The effect on quantity is ambiguous

::: {.fragment}
::: {.highlight-box}
**Answer: D**
:::

The new airline increases demand (shift right → Q↑). The tourist tax increases costs (supply shifts left → Q↓). Since one shift pushes Q up and the other pushes Q down, the net effect is **ambiguous** — it depends on which shift is larger. However, we can say that price **definitely rises** (both shifts push P upward).
:::

</div>

## Exercise 2: Multiple Choice

<div style="font-size: 82%;">

**Question**: In a perfectly competitive market, if firms are currently earning positive economic profits in the short run, what will happen in the long run?

A. The government will impose a price ceiling

B. New firms enter, supply shifts right, price falls until profit = 0

C. Existing firms raise prices to increase profits further

D. Demand shifts left as consumers find substitutes

::: {.fragment}
::: {.highlight-box}
**Answer: B**
:::

Positive profit → **entry** → supply shifts right → equilibrium price falls → profit is competed away until $P = ATC_{\min}$ → **zero economic profit**. This is the self-correcting mechanism of competitive markets. No government action needed (A), firms can't set prices (C — they're price takers!), and the question is about the supply side, not demand (D).
:::

</div>

## Exercise 3: Open Question :writing_hand:

<div style="font-size: 60%;">

**The market for guided walking tours in Sintra** has the following curves:

- **Demand**: $P = 40 - 0.2Q$ (€ per tour)
- **Short-run supply**: $P = 8 + 0.12Q$ (€ per tour)

All firms are identical with $ATC_{\min} = €20$.

**a)** Find the short-run equilibrium price and quantity.

**b)** Calculate CS, PS, and total welfare at the SR equilibrium.

**c)** Are firms making positive economic profit, zero profit, or losses? How do you know?

**d)** Describe what will happen in the **long run**. What will the new equilibrium price be? Will the number of firms increase or decrease?

**e)** A surge in tourist arrivals shifts demand to $P = 52 - 0.2Q$. Find the new SR equilibrium. Are firms now earning profits or losses?

**f)** After the demand surge, describe the long-run adjustment process. What is the final LR equilibrium price?

</div>

## Exercise 3: Solution — Parts a & b

<div style="font-size: 68%;">

**a)** SR equilibrium: $40 - 0.2Q = 8 + 0.12Q$

$32 = 0.32Q$ → $Q^* = 100$ tours, $P^* = 40 - 0.2(100) = €20$

**b)** At ($Q^* = 100$, $P^* = €20$):

$$CS = \frac{1}{2} \times 100 \times (40 - 20) = \frac{1}{2} \times 100 \times 20 = €1{,}000$$

$$PS = \frac{1}{2} \times 100 \times (20 - 8) = \frac{1}{2} \times 100 \times 12 = €600$$

$$W = CS + PS = 1{,}000 + 600 = €1{,}600$$

</div>

## Exercise 3: Solution — Parts c & d

<div style="font-size: 70%;">

**c)** $P^* = €20 = ATC_{\min} = €20$

Since price equals minimum ATC, firms are earning **zero economic profit**. The market is already at its **long-run equilibrium**! :white_check_mark:

**d)** Since profit is already zero, there is **no incentive** for firms to enter or exit. The long-run equilibrium price **stays at €20**. The number of firms remains the same.

:point_right: This is a special case where SR equilibrium **coincides** with LR equilibrium — the market is at rest.

</div>

## Exercise 3: Solution — Parts e & f

<div style="font-size: 68%;">

**e)** New demand: $P = 52 - 0.2Q$

New SR equilibrium: $52 - 0.2Q = 8 + 0.12Q$ → $44 = 0.32Q$ → $Q^* = 137.5$, $P^* = 52 - 0.2(137.5) = €24.50$

Since $P^* = €24.50 > ATC_{\min} = €20$: firms are now earning **positive economic profit**!

Profit per firm = $(24.50 - 20) \times q_i = €4.50 \times q_i > 0$ (exact firm quantity depends on individual MC).

**f)** Long-run adjustment:

1. Positive profits attract **new firms** → they enter the market
2. Supply shifts **right** (more firms)
3. Equilibrium price **falls** from €24.50 toward $ATC_{\min}$
4. Entry continues until $P = ATC_{\min} = €20$

**Final LR equilibrium**: $P_{LR} = €20$. At this price with the new demand: $Q = \frac{52 - 20}{0.2} = 160$ tours.

The market serves **more tours** (160 vs 100 originally) at the **same price** (€20). The adjustment came entirely through **more firms** entering, not through higher prices. :white_check_mark:

</div>

## Next Lecture {.center}

**April 17, 2026**: Introduction to Game Theory — Prisoner's Dilemma & Nash Equilibrium :chess_pawn:

A brand new topic — what happens when firms are **not** price takers and **interact strategically**?

. . .

:wave: **This completes the Producer Theory block!** Time to move beyond perfect competition.

## Thank You! {.center}

**Questions?** :raising_hand:

:email: paulo.fagandini@ext.universidadeeuropeia.pt

---

*Next class: Friday, April 17, 2026*

::: {.notes}
This is the capstone lecture of the producer theory block. It ties together everything from Lectures 10–15: cost curves determine supply, supply meets demand at equilibrium, and the long-run entry/exit mechanism drives profits to zero. The Lisbon hotel boom example makes the LR adjustment process tangible. Students should leave understanding that competitive markets self-correct through entry and exit, and that zero economic profit is the natural long-run state — not a sign of failure. The exercise deliberately starts at a long-run equilibrium, then shocks demand to show the full SR→LR adjustment process.
:::
