---
title: "Economics Fundamentals"
subtitle: "Lecture 1: Introduction, Macro vs Micro, Scarcity and Efficiency"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    from: markdown+emoji
    theme: [default, custom.scss]
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: false
    footer: "Economics of Tourism | Lecture 1"
    logo: ../images/logo.png
    width: 1600
    height: 900
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(eurostat)
library(tidyverse)
library(lubridate)

# Define UE brand colors
ue_red <- "#C8102E"
ue_black <- "#1D1D1B"
ue_gray <- "#6D6E71"
ue_white <- "#FFFFFF"
```

## Welcome to Economics! :wave: {.center}

:::: {.columns}
::: {.column width="55%"}
**Why study Economics?** :thinking:

:classical_building: Understand government policies

:globe_with_meridians: Navigate global markets

:briefcase: Achieve business success

:chart_with_upwards_trend: :chart_with_downwards_trend: Master supply & demand

:balance_scale: Understand inequality
:::

::: {.column width="45%"}
![](https://images.unsplash.com/photo-1526304640581-d334cdbbf45e?w=500)
:::
::::

::: {.notes}
Economics is essential for management professionals to make informed decisions in an increasingly complex business environment.
:::

## Introduction :man_teacher:

:man_teacher: **Paulo Fagandini**

:email: paulo.fagandini@ext.universidadeeuropeia.pt

:mortar_board: **In Canvas:** Syllabus, Calendar, Group Project

:calendar: **Schedule:** Thursdays & Fridays

## What is Economics? :bulb:

::: {.key-concept}
**Economics** studies how societies use **scarce resources** to produce valuable goods and distribute them among individuals.
:::

. . .

Economics as a **social science** uses simplified hypotheses and relates variables to understand economic phenomena.

## What Does Economics Study? :mag:

:::: {.columns}
::: {.column width="50%"}
:bank: **Institutions & Markets**

How technology influences prices and resource allocation

:money_with_wings: **Finance**

Behavior of financial markets and capital flows
:::

::: {.column width="50%"}
:busts_in_silhouette: **Society**

Income redistribution and social welfare

:chart_with_downwards_trend: **Cycles**

Unemployment, inflation, growth
:::
::::

# The Two (main) Branches of Economics

## Macroeconomics :telescope:

**The Big Picture** — Studies the economy as a whole

:::: {.columns}
::: {.column width="50%"}
:chart_with_upwards_trend: **GDP & Economic Growth**

:briefcase: **Unemployment rates**

:moneybag: **Inflation trends**
:::

::: {.column width="50%"}
:bank: **Monetary policy**

:globe_with_meridians: **International trade**

:receipt: **Government budgets**
:::
::::

::: {.fragment}
*Example:* "Why did Portugal's tourism revenue fall 50% in 2020?"
:::

## Microeconomics :microscope:

**Individual Decisions** — Studies specific agents and markets

:::: {.columns}
::: {.column width="50%"}
:shopping_cart: **Consumer behavior**

:factory: **Firm decisions**

:label: **Pricing strategies**
:::

::: {.column width="50%"}
:crossed_swords: **Competition analysis**

:handshake: **Market interactions**

:chart: **Supply & Demand**
:::
::::

::: {.fragment}

*Example:* "How should a hotel set its room prices?"

:::

## Macroeconomics in Tourism :airplane:

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 13
#| fig-height: 5.5
#| fig-cap: "Source: Eurostat (tour_occ_arnraw, bop_its6_det)"

guests_raw <- get_eurostat("tour_occ_arnraw", type = "code", keep_flags = FALSE, cache = TRUE, cache_dir = "data_cache")
revenue_raw <- get_eurostat("bop_its6_det", type = "code", keep_flags = FALSE, cache = TRUE, cache_dir = "data_cache")

# Clean Guest Data
guests_PT <- guests_raw %>%
  filter(
    geo=='PT',
    nace_r2=='I551-I553',
    unit %in% c('NR','MIO_EUR'),
    c_resid == "WORLD",
    year(TIME_PERIOD)>=2019
  ) %>%
  select(TIME_PERIOD, values)

guests_PT$values <- guests_PT$values/1000000
guests_PT <- guests_PT %>% rename(guests = values)

# Clean Revenue Data
revenue_PT <- revenue_raw %>% 
  filter(
    geo == "PT",
    year(TIME_PERIOD) >= 2019,
    partner == "WRL_REST",
    stk_flow == "CRE",
    bop_item == "SD"
  ) %>% 
  select(TIME_PERIOD, values)

revenue_PT$values <- revenue_PT$values/1000
revenue_PT <- revenue_PT %>% rename(revenue = values)

# Merge and Plot
df_plot <- inner_join(guests_PT, revenue_PT, by = "TIME_PERIOD") %>%
  mutate(year_val = year(TIME_PERIOD)) %>%
  pivot_longer(cols = c(guests, revenue), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, 
                         "guests" = "Guests (Millions)", 
                         "revenue" = "Revenue (Billions €)"))

ggplot(df_plot, aes(x = year_val, y = value, color = metric, group = metric)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_color_manual(values = c("Guests (Millions)" = ue_red, "Revenue (Billions €)" = ue_black)) +
  theme_minimal(base_size = 16) +
  labs(title = "Portugal Tourism Sector (2019-2024)",
       subtitle = "COVID-19 impact and recovery visible in macroeconomic data",
       x = "Year", y = "Scale (Millions / Billions)", color = "Indicator") +
  theme(legend.position = "bottom",
        plot.title = element_text(color = ue_red, face = "bold"))
```

## Microeconomics in Tourism :hotel:

```{r}
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 5
#| fig-cap: "Source: Eurostat (tour_occ_arnraw)"

guests_micro <- guests_raw %>% 
  filter(geo == "PT", year(TIME_PERIOD)==2024, c_resid == "WORLD", 
         nace_r2 %in% c('I551', 'I552', 'I553'), unit == 'NR')
guests_micro <- guests_micro %>% select(nace_r2, values) %>% rename(guests = values)

nace_descriptions <- c(
  'I551' = 'Hotels',
  'I552' = 'Holiday rentals',
  'I553' = 'Camping'
)

guests_micro$description <- nace_descriptions[guests_micro$nace_r2]

ggplot(data = guests_micro, aes(x = reorder(description, -guests), y = guests/1000000, fill = description)) +
  geom_bar(stat = "identity", color = ue_black, linewidth = 0.5) +
  geom_text(aes(label = paste0(round(guests/1000000, 1), "M")),
            vjust = -0.5, size = 6, fontface = "bold") +
  scale_fill_manual(values = c("Hotels" = ue_red, "Holiday rentals" = ue_gray, "Camping" = ue_black)) +
  labs(title = "Guests by Accommodation Type — Portugal 2024",
       subtitle = "Microeconomic analysis: which market segments attract more tourists?",
       x = "", y = "Guests (Millions)") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none",
        plot.title = element_text(color = ue_red, face = "bold", hjust = 0.5),
        axis.text.x = element_text(size = 14, face = "bold"))
```

## The Fundamental Economic Problem :warning:

::: {.key-concept}
**Scarcity**

Human wants are unlimited, but resources are limited
:::

. . . 

This leads to:

:point_right: **Trade-offs** — Choices involve compromises

:point_right: **Cost-Benefit Analysis** — Act only if benefits exceed costs

:point_right: **Efficiency** — Most effective use of resources

## Scarcity in Tourism: Lisbon Airport :flight_departure:

:::: {.columns}
::: {.column width="55%"}
**Capacity Constraints (2024)**

:warning: Design capacity: ~31M passengers/year

:airplane: Actual 2024: ~35M passengers

:chart_with_upwards_trend: Growing demand, limited slots

**Trade-off:** More flights vs. congestion & delays
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| warning: false
#| fig-width: 5
#| fig-height: 4

# Hypothetical illustrative data
capacity_data <- data.frame(
  category = c("Capacity", "Actual 2024", "Projected 2025"),
  passengers = c(31, 35, 36.5)
)
capacity_data$category <- factor(capacity_data$category, 
                                  levels = c("Capacity", "Actual 2024", "Projected 2025"))

ggplot(capacity_data, aes(x = category, y = passengers, fill = category)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_hline(yintercept = 31, linetype = "dashed", color = ue_red, linewidth = 1.2) +
  labs(y = "Passengers (M)", x = "") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c(ue_black, ue_red, ue_gray)) +
  theme(legend.position = "none") +
  annotate("text", x = 2.5, y = 32.5, label = "Capacity Limit", color = ue_red, size = 4, fontface = "bold")
```

*Hypothetical illustration based on news reports*
:::
::::

## Efficiency :dart:

::: {.key-concept}
**Economic Efficiency**

Situation where the economy produces and consumes goods at *socially optimal levels*:

:white_check_mark: Getting the most out of available resources

:white_check_mark: Using the least for current consumption level
:::

## Pareto Efficiency :balance_scale:

**Cannot increase one person's welfare without harming another**

:::: {.columns}
::: {.column width="50%"}
:heavy_check_mark: **Efficient allocation**

- Resources fully utilized
- No waste in production
- Optimal distribution
:::

::: {.column width="50%"}
:x: **Inefficient allocation**

- Resources underutilized
- Production waste exists
- Room for improvement
:::
::::

## Efficiency in Tourism :hotel:

**Dynamic Pricing Example**

```{r}
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 4.5

# Hypothetical illustrative data
pricing_data <- data.frame(
  strategy = rep(c("Fixed Pricing", "Dynamic Pricing"), each = 12),
  month = rep(month.abb, 2),
  occupancy = c(45, 48, 55, 52, 50, 85, 95, 92, 70, 55, 50, 60,
                65, 68, 75, 72, 70, 90, 98, 95, 85, 75, 70, 78),
  month_num = rep(1:12, 2)
)

ggplot(pricing_data, aes(x = month_num, y = occupancy, color = strategy, group = strategy)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  labs(title = "Hotel Occupancy: Fixed vs. Dynamic Pricing",
       subtitle = "Dynamic pricing improves efficiency through better resource allocation (hypothetical)",
       x = "", y = "Occupancy Rate (%)", color = "") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Fixed Pricing" = ue_gray, "Dynamic Pricing" = ue_red)) +
  theme(legend.position = "bottom",
        plot.title = element_text(color = ue_red, face = "bold"))
```

## The Three Economic Questions :question:

Every economy must answer:

| Question | Description | Tourism Example |
|----------|-------------|-----------------|
| **WHAT?** | What goods to produce? | More hotels or museums? |
| **HOW?** | How to produce them? | Automated or human agents? |
| **FOR WHOM?** | Who gets the output? | Luxury or mass tourism? |

::: {.fragment}
*These require decisions based on: resources, technology, preferences, markets*
:::

## Positive vs. Normative Economics :mag_right:

:::: {.columns}
::: {.column width="50%"}
### Positive Economics :bar_chart:
**Describes facts** — Can be tested

- Why do managers earn more than cleaners?
- Does tourism increase local incomes?
- What happens to prices in high season?

:point_right: *Answered with data and analysis*
:::

::: {.column width="50%"}
### Normative Economics :thought_balloon:
**Value judgments** — Opinions

- Should government limit Airbnb?
- Should tourism revenue be redistributed?
- Should airlines reduce emissions?

:point_right: *Requires ethical/political debate*
:::
::::

## Key Concepts Summary :clipboard:

| Concept | Definition |
|---------|------------|
| **Economics** | Study of scarce resource allocation |
| **Macroeconomics** | Whole economy (GDP, inflation) |
| **Microeconomics** | Individual decisions (firms, consumers) |
| **Scarcity** | Unlimited wants, limited resources |
| **Efficiency** | Optimal resource use (no waste) |
| **Trade-off** | Giving up one thing for another |

## Exercises :pencil: {.center}

**Time for practice!**

Let's test your understanding.

## Exercise 1: Multiple Choice

**Which of the following is a MACROECONOMIC question?**

A. How should a hotel set its room prices?

B. Why did Portugal's tourism revenue fall in 2020?

C. Should a restaurant hire more waiters?

D. How do travelers choose between two airlines?

::: {.fragment}
::: {.highlight-box}
**Answer: B** — Portugal's tourism revenue is an aggregate measure affecting the entire economy.
:::
:::

## Exercise 2: Multiple Choice

**A museum has limited exhibition space. Choosing between a modern art or historical exhibition illustrates:**

A. Efficiency

B. Scarcity

C. Positive economics

D. Macroeconomics

::: {.fragment}
::: {.highlight-box}
**Answer: B** — Limited space (scarce resource) forces a choice between alternatives.
:::
:::

## Exercise 3: Open Question

**Scenario:** A low-cost airline charges €50/ticket, sells 150 tickets (capacity: 180).

**Questions:**

a) Calculate current revenue per flight

b) If price increases to €60, demand falls to 120 tickets. Calculate new revenue.

c) Which pricing is more efficient? Why?

d) Is this microeconomic or macroeconomic analysis?

## Exercise 3: Solution

**a) Current revenue:**
$$\text{Revenue} = 50 \times 150 = €7,500$$

**b) New revenue:**
$$\text{Revenue} = 60 \times 120 = €7,200$$

**c) Efficiency:**

- €50 price: Revenue €7,500, occupancy 83%
- €60 price: Revenue €7,200, occupancy 67%

:point_right: **First is more efficient**: higher revenue AND better utilization

**d)** This is **microeconomic** analysis (individual firm's decision)

## Next Lecture :books:

**Lecture 2** (February 6, 2026):

- Three Fundamental Economic Problems (detailed)
- Market, Centralized, and Mixed Economies
- Economic Systems Comparison

:bulb: **Preparation:** Think about how tourism businesses make decisions in different economic systems!

## Thank You! :wave: {.center}

**Questions?**

:email: paulo.fagandini@ext.universidadeeuropeia.pt

---

*Next class: Friday, February 6, 2026*
