---
title: "Economics Fundamentals"
subtitle: "Lecture 4: Production Factors & Production Possibilities Frontier"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    from: markdown+emoji
    theme: [default, custom.scss]
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: false
    footer: "Economics of Tourism | Lecture 4"
    logo: ../images/logo.png
    width: 1600
    height: 900
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyverse)
library(lubridate)

# Define UE brand colors
ue_red <- "#C8102E"
ue_black <- "#1D1D1B"
ue_gray <- "#6D6E71"
ue_white <- "#FFFFFF"
```

## Recap: Lecture 3 :rewind: {.center}

**Key Concepts from Last Class**:

:brain: **Rationality**: Cost-benefit decision making

:chart_with_upwards_trend: **Marginal Analysis**: Decisions at the margin (MB vs MC)

:moneybag: **Opportunity Cost**: Value of best alternative foregone

:no_entry_sign: **Sunk Costs**: Ignore them in decisions!

. . .

:dart: **Today**: We build on opportunity cost to understand:

- Production factors (inputs)
- **Production Possibilities Frontier (PPF)**
- Society-level trade-offs and efficiency

# Production Factors

## The Three Main Inputs :factory:

::: {.key-concept}
**Production Factors**

To produce goods and services, economies use three main factors: **Land** (natural resources), **Labor** (human effort), and **Capital** (produced goods used in production).
:::

:::: {.columns}
::: {.column width="33%"}
**Land** :herb:

- Agricultural land
- Minerals & energy
- Water resources
- Natural beauty
- Climate

*Tourism*: Beaches, mountains, heritage sites
:::

::: {.column width="33%"}
**Labor** :busts_in_silhouette:

- Physical work
- Mental work
- Skills & training
- Time & effort
- Entrepreneurship

*Tourism*: Staff, guides, managers, chefs
:::

::: {.column width="33%"}
**Capital** :office:

- Buildings
- Machinery
- Infrastructure
- Technology
- Equipment

*Tourism*: Hotels, planes, booking systems
:::
::::

## Factor Payments :money_with_wings:

**Each factor receives compensation**:

| Factor | Payment Name | Example in Tourism |
|--------|-------------|-------------------|
| **Land** | Rent | Payment for beach concession, property lease |
| **Labor** | Wages/Salaries | Hotel staff wages, tour guide fees |
| **Capital** | Interest | Return on hotel investment, equipment loans |
| **Entrepreneurship** | Profit | Hotel owner's profit, airline earnings |


## Tourism Production: Example :bar_chart:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 6
#| fig-cap: "Hypothetical illustration of factor intensity"

# Factor intensity by tourism type
tourism_types <- data.frame(
  type = rep(c("Beach Resort", "Cultural Tour", "Adventure Tourism", "Cruise Ship"), each = 3),
  factor = rep(c("Land", "Labor", "Capital"), 4),
  percentage = c(
    # Beach Resort
    40, 30, 30,
    # Cultural Tour
    15, 60, 25,
    # Adventure Tourism
    50, 35, 15,
    # Cruise Ship
    5, 25, 70
  )
)

tourism_types$type <- factor(tourism_types$type,
                              levels = c("Beach Resort", "Cultural Tour", 
                                        "Adventure Tourism", "Cruise Ship"))

ggplot(tourism_types, aes(x = type, y = percentage, fill = factor)) +
  geom_bar(stat = "identity", position = "fill", color = ue_black, linewidth = 0.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Factor Intensity in Different Tourism Products",
       subtitle = "Different activities require different combinations of Land, Labor, and Capital",
       x = "",
       y = "Percentage of Total Input Cost",
       fill = "Production Factor") +
  theme_minimal(base_size = 16) +
  scale_fill_manual(values = c("Land" = ue_black, 
                                "Labor" = ue_red,
                                "Capital" = ue_gray)) +
  theme(legend.position = "bottom",
        plot.title = element_text(color = ue_red, face = "bold"),
        axis.text.x = element_text(angle = 15, hjust = 1)) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_fill(vjust = 0.5), 
            color = "white", size = 5, fontface = "bold")
```

:bulb: **Insight**: Cruise tourism is **capital-intensive**, cultural tours are **labor-intensive**, beach resorts balanced

# Production Possibilities Frontier (PPF)

## What is the PPF? :world_map:

::: {.key-concept}
**Production Possibilities Frontier (PPF)**

The PPF shows the **maximum combinations** of two goods an economy can produce given fixed resources, current technology, and full employment.
:::

. . .

**Key Insight** :bulb:

With scarcity, producing more of one good requires producing less of another

→ **This is opportunity cost visualized!**

## Simple PPF Example: Tourism :beach_umbrella:

**Imagine an economy producing only**:

- **Beach Tourism** (hotels, resorts)
- **Cultural Tourism** (museums, tours, heritage sites)

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 6

# Create Linear PPF data
ppf_data <- data.frame(
  beach = c(0, 100),
  cultural = c(100, 0)
)

# Define the Points
points <- data.frame(
  beach =    c(0, 30, 60, 85, 100, 40, 80),
  cultural = c(100, 70, 40, 15, 0, 30, 80),
  label =    c("A", "B", "C", "D", "E", "U", "X"),
  status =   c(rep("Efficient", 5), "Inefficient", "Unattainable")
)

# Plotting
ggplot(ppf_data, aes(x = beach, y = cultural)) +
  # The Linear Frontier
  geom_line(linewidth = 2, color = ue_red) +
  # Shaded feasible region
  geom_ribbon(aes(ymin = 0, ymax = cultural), alpha = 0.1, fill = ue_gray) +
  # Points
  geom_point(data = points, aes(color = status, shape = status), size = 7) +
  # Labels for points
  geom_text(data = points, aes(label = label), 
            nudge_y = 4, nudge_x = 3, size = 6, fontface = "bold") +
  # Styling
  scale_color_manual(values = c(
    "Efficient" = ue_black, 
    "Inefficient" = ue_gray, 
    "Unattainable" = ue_red
  )) +
  scale_shape_manual(values = c(
    "Efficient" = 16, 
    "Inefficient" = 17, 
    "Unattainable" = 15
  )) +
  labs(
    title = "Linear Production Possibilities Frontier",
    subtitle = "Constant Opportunity Cost (1:1 Swap Ratio)",
    x = "Beach Tourism Capacity",
    y = "Cultural Tourism Capacity"
  ) +
  coord_cartesian(xlim = c(0, 110), ylim = c(0, 110)) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(color = ue_red, face = "bold")
  ) +
  # Zone Annotations
  annotate("text", x = 20, y = 20, label = "FEASIBLE &\nINEFFICIENT", 
           color = ue_gray, alpha = 0.7, size = 5, fontface = "bold") +
  annotate("text", x = 85, y = 85, label = "UNATTAINABLE", 
           color = ue_red, alpha = 0.7, size = 5, fontface = "bold")
```

## Interpreting the PPF :mag:

**Point A** (0, 100): All resources → cultural tourism, no beach tourism

**Point E** (100, 0): All resources → beach tourism, no cultural tourism

**Point C** (60, 40): Mixed economy, balanced allocation

**Point U** (40, 30): **Inefficient** — inside PPF

- Resources unemployed or misallocated
- Could produce more of both goods!

**Point X** (80, 80): **Unattainable** — outside PPF

- Not enough resources with current technology
- Would need economic growth

## Moving Along the PPF :arrow_right:

**From B (30, 70) to C (60, 40)**:

:white_check_mark: Gain: 30 units of beach tourism

:x: Loss: 30 units of cultural tourism

:moneybag: **Opportunity Cost**: 30 cultural units

. . .

**Key Point** :key:

- Moving along PPF = **reallocation** of resources
- All points on PPF are **efficient** (no waste)
- But different distributions (different answers to WHAT question)
- Trade-offs are unavoidable!

## Bowed-Out PPF: Increasing Opportunity Cost :arrows_counterclockwise:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 6

# Create concave PPF
beach_vals <- seq(0, 100, by = 1)
cultural_vals <- 100 * sqrt(1 - (beach_vals/100)^2)

ppf_concave <- data.frame(
  beach = beach_vals,
  cultural = cultural_vals
)

# Points on concave PPF
points_concave <- data.frame(
  beach = c(0, 20, 40, 60, 80, 100),
  cultural = c(100, 98, 92, 80, 60, 0),
  label = c("A", "B", "C", "D", "E", "F")
)

ggplot(ppf_concave, aes(x = beach, y = cultural)) +
  geom_line(linewidth = 2, color = ue_red) +
  geom_ribbon(aes(ymin = 0, ymax = cultural), alpha = 0.1, fill = ue_gray) +
  geom_point(data = points_concave, size = 7, color = ue_black) +
  geom_text(data = points_concave, aes(label = label), 
            nudge_y = 4, size = 6, fontface = "bold") +
  labs(title = "Bowed-Out PPF: Increasing Opportunity Costs",
       subtitle = "As you produce more of one good, opportunity cost increases",
       x = "Beach Tourism Capacity",
       y = "Cultural Tourism Capacity") +
  coord_cartesian(xlim = c(0, 110), ylim = c(0, 110)) +
  theme_minimal(base_size = 16) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(color = ue_red, face = "bold"))
```

:bulb: **Why bowed out?** Resources aren't perfectly adaptable — some resources are better suited for beach tourism, others for cultural tourism

## Calculating Opportunity Cost on PPF :calculator:

**Example: Moving from C to D**

:::: {.columns}
::: {.column width="50%"}
**Point C**: (40 beach, 92 cultural)

**Point D**: (60 beach, 80 cultural)

**Change**:

- Gain: 20 beach units
- Loss: 12 cultural units
:::

::: {.column width="50%"}
**Opportunity Cost**:

$$OC = \frac{\text{Units Lost}}{\text{Units Gained}} = \frac{12}{20} = 0.6$$

:point_right: Each additional beach unit costs **0.6 cultural units**
:::
::::

::: {.fragment}
**Compare: D to E**

- Gain: 20 beach (60→80)
- Loss: 20 cultural (80→60)
- OC = 20/20 = **1.0** (higher!)

:bulb: Opportunity cost **increases** as we specialize more
:::

# Economic Growth & PPF Shifts

## Shifting the PPF Outward :rocket:

::: {.key-concept}
**Economic Growth**

An outward shift of the PPF represents economic growth — the ability to produce more of both goods. Caused by: more resources, better technology, or improved productivity.
:::

. . .

**Three ways to shift PPF outward**:

:arrow_up: **More resources** (population growth, discover new land)

:bulb: **Better technology** (innovation, improved processes)

:books: **Human capital** (education, training, skills)

## PPF Shift Example :chart_with_upwards_trend:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 6

# Two PPFs for shift
beach_1 <- seq(0, 100, by = 1)
cultural_1 <- 100 * sqrt(1 - (beach_1/100)^2)

beach_2 <- seq(0, 150, by = 1)
cultural_2 <- 150 * sqrt(1 - (beach_2/150)^2)

ppf_shift <- data.frame(
  beach = c(beach_1, beach_2),
  cultural = c(cultural_1, cultural_2),
  era = c(rep("2020: Before Innovation", length(beach_1)),
          rep("2025: After Innovation", length(beach_2)))
)

ggplot(ppf_shift, aes(x = beach, y = cultural, color = era)) +
  geom_line(linewidth = 2) +
  labs(title = "PPF Shift: Technological Innovation in Tourism",
       subtitle = "New technology (AI, digital platforms) expands production possibilities",
       x = "Beach Tourism Capacity",
       y = "Cultural Tourism Capacity",
       color = "") +
  scale_color_manual(values = c("2020: Before Innovation" = ue_gray, 
                                 "2025: After Innovation" = ue_red)) +
  coord_cartesian(xlim = c(0, 160), ylim = c(0, 160)) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom",
        plot.title = element_text(color = ue_red, face = "bold"),
        panel.grid.minor = element_blank()) +
  annotate("segment", x = 50, xend = 75, y = 87, yend = 130,
           arrow = arrow(length = unit(0.5, "cm"), type = "closed"),
           linewidth = 1.5, color = ue_black) +
  annotate("text", x = 55, y = 115, 
           label = "Economic Growth", 
           size = 6, color = ue_black, fontface = "bold")
```

## Investment Trade-off: Present vs Future :balance_scale:

**Key Economic Decision**: Consumption today vs. Growth tomorrow

:::: {.columns}
::: {.column width="50%"}
**More Consumption Goods Today**

:white_check_mark: Higher current standard of living

:white_check_mark: Immediate satisfaction

:x: Less investment in capital

:x: Slower future growth
:::

::: {.column width="50%"}
**More Capital Goods Today**

:white_check_mark: Faster PPF outward shift

:white_check_mark: Higher future production

:x: Lower current consumption

:x: Sacrifice today for tomorrow
:::
::::

::: {.fragment}
:bulb: **Tourism Example**: Build hotels now (consume less) → More capacity future → Higher tourism revenue later
:::

## Investment Trade-off Visualization :moneybag:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 6

# Two countries with different investment choices
consumption_vals <- seq(0, 100, by = 1)
capital_vals <- 100 * sqrt(1 - (consumption_vals/100)^2)

ppf_investment <- data.frame(
  consumption = consumption_vals,
  capital = capital_vals
)

# Points showing choices
choices <- data.frame(
  consumption = c(80, 40),
  capital = c(60, 92),
  country = c("Country A:\nHigh Consumption", "Country B:\nHigh Investment"),
  future_growth = c("Slower Growth", "Faster Growth")
)

ggplot(ppf_investment, aes(x = consumption, y = capital)) +
  geom_line(linewidth = 2, color = ue_red) +
  geom_point(data = choices, size = 8, color = ue_black) +
  geom_text(data = choices, aes(label = country), 
            nudge_x = c(-15, 15), nudge_y = c(-8, 8), 
            size = 5, fontface = "bold") +
  labs(title = "Consumption vs. Investment: Different Choices",
       subtitle = "Country B sacrifices consumption today for higher growth tomorrow",
       x = "Consumption Goods",
       y = "Capital/Investment Goods") +
  coord_cartesian(xlim = c(0, 110), ylim = c(0, 110)) +
  theme_minimal(base_size = 16) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(color = ue_red, face = "bold")) +
  annotate("text", x = 80, y = 45, 
           label = "Less future\ngrowth", color = ue_gray, size = 4, fontface = "italic") +
  annotate("text", x = 40, y = 75, 
           label = "More future\ngrowth", color = ue_black, size = 4, fontface = "italic")
```

## Tourism Investment Example :hotel:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 5.5

# Tourism investment categories
investment_categories <- data.frame(
  category = c("Marketing\nCampaigns", "Staff\nTraining", "Infrastructure\nDevelopment", "Digital\nTransformation"),
  immediate_impact = c(85, 60, 35, 40),
  longterm_impact = c(40, 75, 95, 90)
)

investment_categories$category <- factor(investment_categories$category,
                                          levels = investment_categories$category)

investment_long <- investment_categories %>%
  pivot_longer(cols = c(immediate_impact, longterm_impact),
               names_to = "timeframe",
               values_to = "score") %>%
  mutate(timeframe = recode(timeframe,
                            "immediate_impact" = "Immediate Revenue Impact",
                            "longterm_impact" = "Long-term Growth Potential"))

ggplot(investment_long, aes(x = category, y = score, fill = timeframe)) +
  geom_bar(stat = "identity", position = "dodge", color = ue_black, linewidth = 0.5) +
  labs(title = "Tourism Investment Categories: Trade-off Analysis",
       subtitle = "Infrastructure & Digital have high long-term impact but low immediate returns",
       x = "",
       y = "Impact Score (0-100)",
       fill = "") +
  theme_minimal(base_size = 16) +
  scale_fill_manual(values = c("Immediate Revenue Impact" = ue_red, 
                                "Long-term Growth Potential" = ue_black)) +
  theme(legend.position = "bottom",
        plot.title = element_text(color = ue_red, face = "bold"),
        axis.text.x = element_text(size = 13)) +
  geom_text(aes(label = score), position = position_dodge(width = 0.9), 
            vjust = -0.5, size = 5, fontface = "bold")
```

:point_right: **Strategic choice**: Prioritize long-term growth (infrastructure, digital) despite lower immediate returns

## Technological Change Example :computer:

```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 6

# Digital transformation PPF shift
tech_ppf <- data.frame(
  bookings = c(0, 100, 0, 150),
  customer_service = c(100, 0, 150, 0),
  era = rep(c("2015: Manual Systems", "2025: Digital + AI"), each = 2)
)

ggplot() +
  geom_line(data = tech_ppf, aes(x = bookings, y = customer_service, color = era), 
            linewidth = 2) +
  annotate("segment", x = 40, xend = 75, y = 60, yend = 75,
           arrow = arrow(length = unit(0.5, "cm"), type = "closed"),
           linewidth = 1.5, color = ue_black) +
  annotate("text", x = 50, y = 80, 
           label = "AI & Digital Shift", 
           size = 6, color = ue_black, fontface = "bold") +
  annotate("label", x = 110, y = 40, label = "New Feasible\nCombinations", 
           fill = ue_red, color = "white", fontface = "bold", size = 5) +
  scale_color_manual(values = c("2015: Manual Systems" = ue_gray, 
                                 "2025: Digital + AI" = ue_red)) +
  labs(title = "Technology Impact: Travel Agency PPF Shift",
       subtitle = "Digital transformation enables higher efficiency across all services",
       x = "Number of Bookings Processed (per day)",
       y = "Quality of Customer Service (index)",
       color = "") +
  coord_cartesian(xlim = c(0, 160), ylim = c(0, 160)) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom",
        plot.title = element_text(color = ue_red, face = "bold"),
        panel.grid.minor = element_blank())
```

## PPF Summary: Key Insights :bookmark_tabs:

:one: **Shows maximum production** given resources & technology

:two: **Points on PPF** = efficient (can't produce more without trade-off)

:three: **Points inside PPF** = inefficient (unemployment, waste)

:four: **Points outside PPF** = impossible (currently unattainable)

:five: **Moving along PPF** = reallocating resources (opportunity cost)

:six: **Shifting PPF outward** = economic growth (more resources, better technology)

:seven: **Investment trade-off** = consumption today vs. growth tomorrow

## Real-World Applications :earth_africa:

**PPF thinking helps answer**:

:question: Should Portugal build more hotels or preserve more heritage sites?

:question: How much should airlines invest in fuel efficiency vs. passenger comfort?

:question: Trade-off between mass tourism revenue and sustainable tourism?

:question: How much GDP to allocate to tourism vs. other sectors?

::: {.fragment}
:bulb: **All involve opportunity costs and efficiency considerations!**
:::

## Complete Picture: Integration :jigsaw:

**Today's Lecture Integration**:

:one: **Production Factors** (Land, Labor, Capital) are scarce inputs

:two: **Scarcity** forces choices between alternatives

:three: **PPF** visualizes society's production possibilities

:four: **Opportunity Cost** = moving along PPF (what's foregone)

:five: **Efficiency** = operating on (not inside) PPF

:six: **Economic Growth** = PPF shifts outward

:seven: **Investment** today enables faster growth tomorrow

::: {.fragment}
:link: **This connects to**: Lecture 3 (opportunity cost), Lecture 2 (WHAT/HOW/FOR WHOM), Lecture 1 (scarcity & efficiency)
:::

## Exercises :pencil: {.center}

**Application Time!**

PPF and opportunity cost calculations.

## Exercise 1: Multiple Choice

**An economy is producing at a point INSIDE its PPF. This indicates:**

A. The economy is efficiently using all resources

B. The economy has economic growth

C. There is unemployment or resource misallocation

D. It's impossible to produce more of any good

::: {.fragment}
::: {.highlight-box}
**Answer: C** - Inside PPF means inefficiency — resources are unemployed or misallocated. Could produce more of at least one good without sacrificing the other (move toward frontier).
:::
:::

## Exercise 2: Multiple Choice

**A country's PPF shifts outward. This could be caused by:**

A. Higher unemployment

B. Decrease in the labor force

C. Technological innovation

D. Producing less capital goods

::: {.fragment}
::: {.highlight-box}
**Answer: C** - Outward PPF shift = economic growth. Caused by: more resources, better technology, improved productivity. Options A, B, D would shift PPF inward or cause movement inside PPF.
:::
:::

## Exercise 3: Open Question - Setup

**Scenario**: The Portuguese Algarve region has limited land and labor. It can allocate resources between golf tourism and beach tourism. Current production possibilities:

| Combination | Golf Resorts | Beach Hotels |
|-------------|-------------|--------------|
| A | 0 | 50 |
| B | 5 | 48 |
| C | 10 | 44 |
| D | 15 | 38 |
| E | 20 | 30 |
| F | 25 | 0 |

## Exercise 3: Questions

**Questions**:

a) Draw the PPF with Golf Resorts on x-axis, Beach Hotels on y-axis

b) Calculate the opportunity cost of moving from:
   - Point B to Point C
   - Point D to Point E

c) Does the PPF exhibit increasing opportunity costs? Explain.

d) If region currently operates at point U (12 golf, 30 beach), what does this indicate?

e) A new tourism development technology is invented. How would this affect the PPF? Draw the new curve.

## Exercise 3: Solution - Part a (PPF Graph)
```{r}
#| echo: false
#| warning: false
#| fig-width: 13
#| fig-height: 6

# Data from the exercise
ppf_algarve <- data.frame(
  golf = c(0, 5, 10, 15, 20, 25),
  beach = c(50, 48, 44, 38, 30, 0),
  label = c("A", "B", "C", "D", "E", "F")
)

ggplot(ppf_algarve, aes(x = golf, y = beach)) +
  geom_line(linewidth = 2, color = ue_red) +
  geom_ribbon(aes(ymin = 0, ymax = beach), alpha = 0.1, fill = ue_gray) +
  geom_point(size = 7, color = ue_black) +
  geom_text(aes(label = label), nudge_y = 2.5, size = 6, fontface = "bold") +
  labs(title = "Algarve Tourism PPF: Golf Resorts vs. Beach Hotels",
       subtitle = "Bowed-out curve indicates increasing opportunity costs",
       x = "Golf Resorts",
       y = "Beach Hotels") +
  coord_cartesian(xlim = c(0, 27), ylim = c(0, 55)) +
  theme_minimal(base_size = 16) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(color = ue_red, face = "bold")) +
  annotate("text", x = 5, y = 10, 
           label = "FEASIBLE &\nINEFFICIENT", 
           color = ue_gray, alpha = 0.7, size = 5, fontface = "bold")
```

**a) The PPF shows all efficient combinations (points A-F) and the bowed-out shape indicating increasing opportunity costs.**

## Exercise 3: Solution - Part b

**b) Opportunity Cost Calculations**:

**B → C** (5 golf → 10 golf):
$$\text{Gain in Golf} = 10 - 5 = 5 \text{ resorts}$$
$$\text{Loss in Beach} = 48 - 44 = 4 \text{ hotels}$$
$$\text{Opportunity Cost} = \frac{4 \text{ hotels}}{5 \text{ resorts}} = 0.8 \text{ beach hotels per golf resort}$$

## Exercise 3: Solution - Part b continued

**D → E** (15 golf → 20 golf):
$$\text{Gain in Golf} = 20 - 15 = 5 \text{ resorts}$$
$$\text{Loss in Beach} = 38 - 30 = 8 \text{ hotels}$$
$$\text{Opportunity Cost} = \frac{8 \text{ hotels}}{5 \text{ resorts}} = 1.6 \text{ beach hotels per golf resort}$$

## Exercise 3: Solution - Part c & d

**c) Increasing Opportunity Costs?**

**YES!** Opportunity cost increases from 0.8 (B→C) to 1.6 (D→E).

**Explanation**: As more resources shift to golf, we use resources less suited for golf (originally better for beaches). Each additional golf resort requires sacrificing more beach hotels.

**d) Point U (12 golf, 30 beach)**:

This is **INSIDE the PPF** (compare to point D: 15 golf, 38 beach).

**Indicates**:

- **Inefficiency** (unemployment or resource misallocation)
- Could produce more golf without reducing beach (move to D: 15 golf, 38 beach)
- Could produce more beach without reducing golf

## Exercise 3: Solution - Part e

**e) New Technology Effect**:

**Technology shifts PPF OUTWARD** (economic growth).

New PPF might be:

| Combination | Golf Resorts | Beach Hotels |
|-------------|-------------|--------------|
| A' | 0 | 60 (+10) |
| C' | 12 | 55 (+11) |
| E' | 24 | 38 (+8) |
| F' | 30 | 0 (+5) |

**Effect**: Can now produce MORE of both types of tourism with same resources!

**Graph**: Draw original PPF, then new PPF further from origin (parallel or biased depending on where technology applies).

## Next Lecture :books:

**February 19, 2026**: Budget Set and Budget Constraint

:bulb: **Preparation**: Think about how you allocate your own limited budget across different goods!

## Thank You! :wave: {.center}

**Questions?**

:email: paulo.fagandini@ext.universidadeeuropeia.pt

---

*Next class: Thursday, February 19, 2026*

::: {.notes}
This completes the Fundamentals block. Students should understand: scarcity → trade-offs → opportunity cost → PPF → efficiency. These concepts are foundational for all future topics.
:::
