---
title: "Economics Fundamentals"
subtitle: "Lecture 1: Introduction, Macro vs Micro, Scarcity and Efficiency"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    theme: default
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: true
    footer: "Economics of Tourism | Lecture 1"
    logo: ../images/logo.png
---

## Welcome to Economics! {.center}

:::: {.columns}
::: {.column width="50%"}
**Why study Economics?**

- Understand government policies
- Navigate global markets
- Achieve business success
- Master supply & demand
- Understand inequality
:::

::: {.column width="50%"}
![](https://images.unsplash.com/photo-1526304640581-d334cdbbf45e?w=600)
:::
::::

::: {.notes}
Economics is essential for management professionals to make informed decisions in an increasingly complex business environment.
:::

## What is Economics?

**Economics studies how societies use scarce resources to produce valuable goods and distribute them among individuals**

## What is Economics?

Key aspects Economics addresses:

- How institutions and technology influence **prices** and **resource allocation**
- Behavior of **financial markets** and **capital allocation**
- **Income redistribution** and social welfare
- **Economic cycles**: unemployment and inflation
- **International trade** and globalization effects
- **Economic growth** in developing countries

## The Two Branches of Economics

:::: {.columns}
::: {.column width="50%"}
### Macroeconomics
**The Big Picture**

- Unemployment rates
- Inflation trends
- Budget deficits
- Economic growth
- International trade

*Focus: Entire economy*
:::

::: {.column width="50%"}
### Microeconomics
**Individual Decisions**

- Specific markets
- Consumer behavior
- Firm decisions
- Pricing strategies
- Competition analysis

*Focus: Individual agents*
:::
::::

## Macroeconomics in Tourism

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

library(ggplot2)
library(dplyr)

# Real data from INE Portugal & Turismo de Portugal
tourism_data <- data.frame(
  year = c(2019, 2020, 2021, 2022, 2023, 2024),
  guests_millions = c(27, 10.5, 15.8, 26.5, 30.5, 31.6),  # Real INE data
  revenue_billions = c(18.3, 7.6, 11.6, 22.0, 25.4, 27.7)  # Real Banco de Portugal data
)

ggplot(tourism_data, aes(x = year)) +
  geom_line(aes(y = guests_millions, color = "Guests (millions)"), size = 1.5) +
  geom_point(aes(y = guests_millions, color = "Guests (millions)"), size = 3) +
  geom_line(aes(y = revenue_billions, color = "Revenue (€ billions)"), size = 1.5) +
  geom_point(aes(y = revenue_billions, color = "Revenue (€ billions)"), size = 3) +
  labs(title = "Portugal Tourism: Recovery After COVID-19",
       subtitle = "Macroeconomic impact of tourism sector",
       x = "Year",
       y = "Value",
       color = "Metric",
       caption = "Sources: INE Portugal (guests); Banco de Portugal (revenue)") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Guests (millions)" = "#e74c3c", 
                                 "Revenue (€ billions)" = "#3498db")) +
  theme(legend.position = "bottom",
        plot.caption = element_text(size = 8, hjust = 0, color = "#666"))

```

**Source**: INE Portugal, 2025

## Microeconomics in Tourism

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Hotel occupancy rates by segment
hotel_data <- data.frame(
  segment = c("Budget", "Economy", "Midscale", "Upscale"),
  occupancy_2023 = c(72.1, 73.8, 72.7, 70.5),
  occupancy_2024 = c(70.1, 73.1, 73.1, 74.2)
)

library(tidyr)
hotel_long <- hotel_data %>%
  pivot_longer(cols = starts_with("occupancy"), 
               names_to = "year", 
               values_to = "occupancy") %>%
  mutate(year = ifelse(year == "occupancy_2023", "2023", "2024"))

ggplot(hotel_long, aes(x = segment, y = occupancy, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "European Hotel Occupancy Rates by Segment",
       subtitle = "Microeconomic analysis: firm-level performance",
       x = "Hotel Segment",
       y = "Occupancy Rate (%)",
       fill = "Year") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("2023" = "#95a5a6", "2024" = "#27ae60")) +
  theme(legend.position = "bottom") +
  ylim(0, 100)
```

**Source**: STR Europe, 2024

## The Fundamental Economic Problem

### Scarcity

> **Human wants are unlimited, but resources are limited**

**The Principle of Scarcity**: Having more of one good usually means having less of another

## The Fundamental Economic Problem

This leads to:

- **Trade-offs**: Choices involve compromises
- **Cost-Benefit Analysis**: Action taken only if benefits exceed costs
- **Efficiency**: Most effective use of society's resources

## Scarcity in Tourism: Real Example

**Lisbon Airport Capacity Constraints (2024)**

:::: {.columns}
::: {.column width="50%"}

- Capacity: ~31M passengers/year
- 2024: ~35M passengers
- Growing demand but limited slots
- Trade-off: More flights vs. congestion
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| warning: false
#| fig-width: 5
#| fig-height: 4

capacity_data <- data.frame(
  category = c("Capacity", "Actual (2024)", "Demand (2025)"),
  passengers = c(31, 35, 36.1)
)

ggplot(capacity_data, aes(x = category, y = passengers, fill = category)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 31, linetype = "dashed", color = "red", size = 1) +
  labs(title = "Lisbon Airport: Scarcity in Action",
       y = "Passengers (millions)",
       x = "") +
  theme_minimal(base_size = 12) +
  scale_fill_manual(values = c("#3498db", "#e74c3c", "#f39c12")) +
  theme(legend.position = "none") +
  annotate("text", x = 2, y = 32, label = "Capacity Limit", color = "red", size = 4)
```
:::
::::

## Efficiency

:::{.callout}

## Economic Efficiency

Situation where the economy produces and consumes goods at socially optimal levels:

* Getting the most out the available resources
* Using the least for the current level of consumption

:::

## Efficiency

**Characteristics**:

- Cannot increase one person's welfare without harming another (This is known as Pareto Efficiency)
- Resources used in most effective way
- Society operates on Production Possibilities Frontier (PPF) (forthcoming)

## Efficiency

**In Tourism Context**:

- Optimal allocation of hotel rooms
- Efficient pricing of airline tickets
- Balanced distribution of tourists across destinations

## Efficiency vs. Inefficiency

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Example: hotel room pricing efficiency
pricing_data <- data.frame(
  strategy = rep(c("Fixed Pricing", "Dynamic Pricing"), each = 12),
  month = rep(month.abb, 2),
  occupancy = c(45, 48, 55, 52, 50, 85, 95, 92, 70, 55, 50, 60,  # Fixed
                65, 68, 75, 72, 70, 90, 98, 95, 85, 75, 70, 78), # Dynamic
  month_num = rep(1:12, 2)
)

ggplot(pricing_data, aes(x = month_num, y = occupancy, color = strategy, group = strategy)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  labs(title = "Hotel Occupancy: Fixed vs. Dynamic Pricing",
       subtitle = "Dynamic pricing improves efficiency through better resource allocation",
       x = "Month",
       y = "Occupancy Rate (%)",
       color = "Pricing Strategy") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Fixed Pricing" = "#e74c3c", 
                                 "Dynamic Pricing" = "#27ae60")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## The Three Economic Questions

Every economy must answer:

:::: {.columns}
::: {.column width="33%"}
### 1. WHAT?
What goods to produce and in what quantity?

*Tourism example*: More hotels or more museums?
:::

::: {.column width="33%"}
### 2. HOW?
How should goods be produced?

*Tourism example*: Automated booking or human agents?
:::

::: {.column width="33%"}
### 3. FOR WHOM?
Who enjoys the goods produced?

*Tourism example*: Luxury tourism or mass tourism?
:::
::::

---

These questions require economic decisions based on:

- Available resources
- Technology
- Social preferences
- Market mechanisms

## Positive vs. Normative Economics

Two different ways in which we can approach Economics.

## Positive Economics
**Describes facts**

*Questions:*

- Why do hotel managers earn more than cleaners?
- Does tourism increase local incomes?
- What happens to prices during high season?

→ *Can be answered with data and analysis*



## Normative Economics
**Involves value judgments**

*Questions:*

- Should government limit Airbnb rentals?
- Should tourism revenue be redistributed?
- Should airlines be forced to reduce emissions?

→ *Requires ethical and political debate*

## Summary: Key Concepts

1. **Economics** studies how societies allocate scarce resources
2. **Macroeconomics** focuses on the entire economy (GDP, inflation, unemployment)
3. **Microeconomics** focuses on individual decisions (firms, consumers, markets)
4. **Scarcity** means unlimited wants with limited resources
5. **Efficiency** means optimal use of resources (no waste)
6. **Three Questions**: WHAT, HOW, FOR WHOM
7. **Positive** vs. **Normative** analysis

## Exercises {.center}

**Time for practice!**

Let's test your understanding with some questions.

## Exercise 1: Multiple Choice

**Question**: Which of the following is a **macroeconomic** question related to tourism?

A. How should a hotel set its room prices?  
B. Why did Portugal's tourism revenue fall in 2020?  
C. Should a restaurant hire more waiters?  
D. How do travelers choose between two airlines?

::: {.fragment}
**Answer: B**

Explanation: Portugal's overall tourism revenue is an aggregate measure affecting the entire economy (macroeconomic). The others focus on individual firms/consumers (microeconomic).
:::

## Exercise 2: Multiple Choice

**Question**: A museum in Lisbon has limited exhibition space. It must choose between hosting a modern art exhibition or a historical artifacts exhibition. This situation illustrates:

A. Efficiency  
B. Scarcity  
C. Positive economics  
D. Macroeconomics

::: {.fragment}
**Answer: B**

Explanation: Limited space (scarce resource) forces a choice between alternatives. This is the fundamental problem of scarcity.
:::

## Exercise 3: Open Question

**Scenario**: A low-cost airline operates flights between Lisbon and Paris. Currently, they charge €50 per ticket and sell 150 tickets per flight (capacity: 180 seats).

**Questions**:

a) Calculate current revenue per flight
b) If they increase price to €60, demand falls to 120 tickets. Calculate new revenue.
c) Which pricing is more efficient? Why?
d) Is this a microeconomic or macroeconomic analysis?

## Exercise 3: Solution

**a) Current revenue**:

$$\text{Revenue} = \text{Price} \times \text{Quantity} = 50 \times 150 = €7,500$$

**b) New revenue at higher price**:

$$\text{Revenue} = 60 \times 120 = €7,200$$

**c) Efficiency analysis**:

- First scenario (€50): Revenue €7,500, occupancy 83% (150/180)
- Second scenario (€60): Revenue €7,200, occupancy 67% (120/180)

The **first scenario is more efficient**: higher revenue AND better resource utilization (more seats filled).

**d)** This is **microeconomic** analysis (focuses on individual firm's pricing decision).

## Next Lecture

**Lecture 2** (February 6, 2026):

- Three Fundamental Economic Problems (detailed)
- Market, Centralized, and Mixed Economies
- Economic Systems Comparison

**Preparation**: Think about how tourism businesses make decisions in different economic systems!

## Thank You! {.center}

**Questions?**

*Contact*: paulo.fagandini@ext.universidadeeuropeia.pt

---

*Next class: Thursday, February 6, 2026*
