---
title: "Economics Fundamentals"
subtitle: "Lecture 2: Three Economic Problems & Economic Systems"
author: "Paulo Fagandini"
date:  today
date-format: "YYYY"
format:
  revealjs:
    theme: default
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: true
    footer: "Economics of Tourism | Lecture 2"
    logo: ../images/logo.png
---

<div style="font-size: 80%;">
</div>

## Recap: Lecture 1 {.center}

**Key concepts we covered**:

- Economics studies **scarcity** and **efficiency**
- **Macroeconomics** (big picture) vs. **Microeconomics** (individual decisions)
- Three fundamental questions: **WHAT**, **HOW**, **FOR WHOM**
- **Positive** (facts) vs. **Normative** (values) economics

Today: How do different economic systems answer these questions?

## The Three Fundamental Economic Problems
<div style="font-size: 75%;">
Every society must solve:

1. **WHAT goods and services to produce?**
   - Which products? How much of each?
   - Current consumption vs. future investment?

2. **HOW to produce them?**
   - Which technologies? Which resources?
   - Labor-intensive vs. capital-intensive?

3. **FOR WHOM are goods produced?**
   - How to distribute output?
   - Who gets what share of production?

</div>

## Problem 1: WHAT to Produce?

**Tourism Example: Portugal 2024**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

library(ggplot2)
library(dplyr)

# Tourism infrastructure investment allocation
investment_data <- data.frame(
  category = c("Hotels", "Restaurants", "Museums", "Transport", "Beach Facilities"),
  investment_2024 = c(450, 280, 120, 380, 150),
  visitors_impact = c(850, 720, 280, 900, 420)
)

investment_data$category <- factor(investment_data$category,
                                    levels = investment_data$category[order(investment_data$investment_2024)])

ggplot(investment_data, aes(x = category, y = investment_2024, fill = category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Tourism Infrastructure Investment in Portugal (2024)",
       subtitle = "WHAT to produce: Allocation decisions (€ millions)",
       x = "",
       y = "Investment (€ millions)") +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "Set3") +
  theme(legend.position = "none") +
  geom_text(aes(label = paste0("€", investment_2024, "M")), 
            hjust = -0.1, size = 4)
```

## Problem 2: HOW to Produce?

<div style="font-size: 70%;">

**Different Production Methods: Hotel Check-in**

:::: {.columns}
::: {.column width="50%"}
### Traditional Method
- Human receptionists
- Paper forms
- Phone bookings
- High labor cost
- Flexible service
- Personal touch

**Cost per check-in**: ~€8-12
:::

::: {.column width="50%"}
### Automated Method
- Self-service kiosks
- Online booking
- Mobile check-in
- High capital cost
- 24/7 availability
- Limited flexibility

**Cost per check-in**: ~€2-4
:::
::::

**Decision factors**: Initial investment, labor costs, customer preferences, technology reliability

</div>

## HOW: Production Cost Comparison

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Cost comparison over time
years <- 1:5
traditional_cost <- 10 * 10000 * years  # €10 per check-in, 10k check-ins/year
automated_cost <- 200000 + (3 * 10000 * years)  # €200k initial + €3 per check-in

cost_data <- data.frame(
  year = rep(years, 2),
  cost = c(traditional_cost, automated_cost) / 1000,  # Convert to thousands
  method = rep(c("Traditional", "Automated"), each = 5)
)

ggplot(cost_data, aes(x = year, y = cost, color = method, group = method)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  labs(title = "Cumulative Cost: Traditional vs. Automated Check-in",
       subtitle = "HOW to produce: Technology choice affects long-term costs",
       x = "Years of Operation",
       y = "Cumulative Cost (€ thousands)",
       color = "Method") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Traditional" = "#e74c3c", 
                                 "Automated" = "#3498db")) +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 2.86, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 3.2, y = 200, 
           label = "Break-even point", angle = 90, size = 4)
```

## Problem 3: FOR WHOM?

**Income Distribution & Tourism Access**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Tourism spending by income quintile
income_tourism <- data.frame(
  quintile = c("Lowest 20%", "Second 20%", "Third 20%", "Fourth 20%", "Highest 20%"),
  avg_vacation_spending = c(450, 950, 1650, 2800, 5200),
  trips_per_year = c(0.3, 0.8, 1.4, 2.1, 3.5)
)

income_tourism$quintile <- factor(income_tourism$quintile,
                                   levels = income_tourism$quintile)

ggplot(income_tourism, aes(x = quintile, y = avg_vacation_spending, fill = quintile)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Annual Vacation Spending by Income Group (EU 2024)",
       subtitle = "FOR WHOM: Income distribution determines who consumes tourism",
       x = "Income Quintile",
       y = "Average Spending (€)") +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "RdYlGn", direction = 1) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 25, hjust = 1)) +
  geom_text(aes(label = paste0("€", avg_vacation_spending)), 
            vjust = -0.5, size = 4)
```

## Economic Systems: Overview

<div style="font-size: 70%;">

Three fundamental ways to organize an economy:

1. **Market Economy** (Laissez-faire)
   - Decisions by individuals and firms
   - Prices and markets coordinate

2. **Centrally Planned Economy**
   - Government makes decisions
   - Central authority coordinates

3. **Mixed Economy** (Most common today)
   - Combination of market and government
   - Balance varies by country

</div>

## Market Economy

**Definition**: Individuals and companies make key production and consumption decisions through markets

**Key Features**:

- **Private ownership** of resources
- **Price mechanism** coordinates decisions
- **Profit motive** drives production
- **Competition** among firms
- **Consumer sovereignty** (monetary votes)

**Tourism Example**: Hotel chains decide what to build, where, and at what price based on profit expectations

## How Markets Work: Example

**Airbnb in Lisbon (Market Mechanism)**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Supply and demand for Airbnb rentals
library(tidyr)

# Price points
prices <- seq(20, 120, by = 10)

# Demand: higher price -> lower quantity
quantity_demanded <- 5000 - 30 * prices

# Supply: higher price -> higher quantity  
quantity_supplied <- -1000 + 40 * prices

market_data <- data.frame(
  price = rep(prices, 2),
  quantity = c(quantity_demanded, quantity_supplied),
  curve = rep(c("Demand", "Supply"), each = length(prices))
)

ggplot(market_data, aes(x = quantity, y = price, color = curve)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  labs(title = "Airbnb Market in Lisbon: Supply and Demand",
       subtitle = "Market economy: Prices coordinate decisions",
       x = "Number of Listings",
       y = "Average Price per Night (€)",
       color = "") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Demand" = "#e74c3c", "Supply" = "#27ae60")) +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 2200, linetype = "dashed", alpha = 0.3) +
  geom_hline(yintercept = 86, linetype = "dashed", alpha = 0.3) +
  annotate("point", x = 2200, y = 86, size = 5, color = "#f39c12") +
  annotate("text", x = 2200, y = 95, 
           label = "Equilibrium\nP=€86, Q=2,200", size = 4)
```

## Centrally Planned Economy

<div style="font-size: 80%;">

**Definition**: Government makes most production and distribution decisions

**Key Features**:

- **Government ownership** of resources
- **Central planning** replaces markets
- **Production quotas** set by authority
- **Price controls** by government
- **Administrative allocation** of resources

**Historical Example**: Soviet tourism industry (state-owned hotels, fixed prices, limited choice)

**Modern Examples**: Cuba, North Korea (though moving toward mixed systems)

</div>

## Market vs. Centralized: Tourism Example

<div style="font-size: 80%;">

:::: {.columns}
::: {.column width="50%"}
### Market System
**Hotel Construction Decision**:

✓ Private investor analyzes demand  
✓ Builds if profitable  
✓ Price determined by market  
✓ Competition ensures efficiency  

**Result**: Hotels where tourists want them
:::

::: {.column width="50%"}
### Centralized System
**Hotel Construction Decision**:

✓ Government planners decide  
✓ Build according to 5-year plan  
✓ Prices set administratively  
✓ No competition  

**Result**: May not match demand
:::
::::

</div>

## Mixed Economy

<div style="font-size: 80%;">

**Definition**: Combination of market mechanisms and government intervention

**Characteristics**:

- Markets direct daily economic activity
- Government regulates and provides public goods
- Balance varies by country and ideology

**Government Roles**:

1. Increase **efficiency** (regulate monopolies, address externalities)
2. Promote **equity** (redistribute income, social safety net)
3. Foster **stability** (manage inflation, unemployment, growth)

</div>

## Mixed Economy in Tourism

**Portugal Tourism (Mixed System)**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Government vs private sector roles
roles_data <- data.frame(
  sector = rep(c("Private Sector", "Government"), each = 5),
  activity = rep(c("Hotels", "Restaurants", "Transport", "Museums", "Promotion"), 2),
  percentage = c(95, 90, 60, 20, 30,  # Private
                 5, 10, 40, 80, 70)   # Government
)

ggplot(roles_data, aes(x = activity, y = percentage, fill = sector)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Public vs. Private Sector in Portuguese Tourism",
       subtitle = "Mixed economy: Balance between market and government",
       x = "",
       y = "Percentage of Activity",
       fill = "Sector") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("Private Sector" = "#3498db", 
                                "Government" = "#e74c3c")) +
  theme(legend.position = "bottom") +
  coord_flip() +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            color = "white", size = 4, fontface = "bold")
```

## Production Factors

<div style="font-size: 80%;">

To answer the three questions, economies use **factors of production** (inputs):

:::: {.columns}
::: {.column width="33%"}
### Land
**Natural resources**

- Agricultural land
- Energy resources
- Minerals
- Clean air & water

*Tourism*: Beaches, mountains, climate
:::

::: {.column width="33%"}
### Labor
**Human effort**

- Physical work
- Mental work
- Skills & education
- Time spent

*Tourism*: Hotel staff, guides, chefs
:::

::: {.column width="33%"}
### Capital
**Produced resources**

- Machinery
- Buildings
- Infrastructure
- Technology

*Tourism*: Hotels, airports, booking systems
:::
::::

</div>

## Production Factors in Tourism

**European Tourism Employment (2024)**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Labor factor in tourism
employment_data <- data.frame(
  country = c("Spain", "France", "Italy", "Portugal", "Greece"),
  tourism_jobs_thousands = c(2750, 2100, 1850, 950, 720),
  percentage_total_employment = c(13.8, 8.2, 7.9, 19.1, 24.3)
)

employment_data$country <- factor(employment_data$country,
                                   levels = employment_data$country)

ggplot(employment_data, aes(x = country, y = percentage_total_employment, 
                             fill = country)) +
  geom_bar(stat = "identity") +
  labs(title = "Tourism as % of Total Employment (Selected EU Countries)",
       subtitle = "Labor factor: Tourism's importance in national economies",
       x = "",
       y = "Tourism Employment (% of total)") +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none") +
  geom_text(aes(label = paste0(percentage_total_employment, "%")), 
            vjust = -0.5, size = 4.5, fontface = "bold") +
  ylim(0, 27)
```

**Source**: WTTC, 2024

## Comparing Economic Systems

<div style="font-size: 80%;">

| Aspect | Market Economy | Centrally Planned | Mixed Economy |
|--------|----------------|-------------------|---------------|
| **Ownership** | Private | State | Both |
| **Coordination** | Prices/markets | Central plan | Markets + regulation |
| **Incentives** | Profit | Quotas/orders | Profit + public interest |
| **Innovation** | High | Low | Medium-High |
| **Efficiency** | Generally high | Often low | Variable |
| **Equity** | Low (unequal) | High (equal poverty) | Medium |
| **Examples** | (Pure: rare) | Cuba, N. Korea | USA, EU, Portugal |

</div>

## Real-World Example: Tourism Recovery

<div style="font-size: 80%;">

**Different Systems, Different COVID-19 Responses**

:::: {.columns}
::: {.column width="50%"}
### Market-Led (Portugal)

- Private hotels closed voluntarily
- Government support (furlough)
- Market-driven reopening
- Rapid innovation (safety protocols)
- Fast recovery: 31.6M visitors (2024)
:::

::: {.column width="50%"}
### Centrally-Planned (Cuba)

- State-owned hotels closed by decree
- No market signals
- Political decision to reopen
- Limited innovation
- Slower recovery: ~50% of 2019 levels
:::
::::

</div>

## Summary: Key Concepts

1. **Three economic problems**: WHAT, HOW, FOR WHOM
2. **Market economy**: Decisions by individuals/firms via prices
3. **Centrally planned**: Government makes decisions
4. **Mixed economy**: Combination (most common today)
5. **Production factors**: Land, Labor, Capital
6. **Each system** has trade-offs (efficiency vs. equity)
7. **Portugal tourism**: Mixed system (mostly private, some government)

## Exercises {.center}

**Practice Time!**

Let's apply what we've learned.

## Exercise 1: Multiple Choice

**Question**: In a market economy, the question "FOR WHOM are goods produced?" is primarily answered by:

A. Government planners deciding who gets what  
B. Equal distribution to all citizens  
C. Willingness and ability to pay
D. First-come, first-served basis

::: {.fragment}
**Answer: C**

Explanation: In market economies, goods go to those who are willing to pay more for them, which answers "for whom."
:::

## Exercise 2: Multiple Choice

**Question**: Which of the following is an example of the **capital** factor of production in tourism?

A. A tour guide's knowledge and skills  
B. The beach at Algarve  
C. A hotel building in Lisbon  
D. The manager's decision-making ability

::: {.fragment}
**Answer: C**

Explanation: Capital consists of produced goods used to make other goods/services. The hotel building is capital. A is labor, B is land, D is labor/entrepreneurship.
:::

## Exercise 3: Open Question

<div style="font-size: 80%;">

**Scenario**: The Portuguese government is considering two policies for tourism development:

- **Policy A**: Let private investors freely build hotels wherever they want (market approach)
- **Policy B**: Government decides where hotels should be built and provides subsidies (planned approach)

**Questions**:

a) Identify which economic system each policy represents  
b) List 2 advantages and 2 disadvantages of each policy  
c) Calculate: If a private hotel costs €5M to build and generates €800k annual profit, what's the payback period?  
d) Recommend which policy Portugal should adopt (justify with economic reasoning)

</div>

## Exercise 3: Solution (Part 1)

<div style="font-size: 80%;">

**a) Economic systems**:

- **Policy A**: Market economy approach (laissez-faire)
- **Policy B**: Centrally planned economy approach

**b) Advantages and Disadvantages**:

| Policy A (Market) | Policy B (Planned) |
|-------------------|-------------------|
| ✓ Efficient allocation (demand-driven) | ✓ Can correct market failures |
| ✓ Innovation & competition | ✓ Ensures regional development |
| ✗ May ignore externalities | ✗ Information problems |
| ✗ Unequal regional development | ✗ Bureaucratic inefficiency |

</div>

## Exercise 3: Solution (Part 2)

<div style="font-size: 80%;">

**c) Payback period calculation**:

$$\text{Payback Period} = \frac{\text{Initial Investment}}{\text{Annual Profit}} = \frac{€5,000,000}{€800,000} = 6.25 \text{ years}$$

**d) Recommendation**: **Mixed approach** (combination)

**Reasoning**:

- Use **market mechanism** (Policy A) as primary driver
  - Private sector is more efficient at identifying demand
  - Competition drives innovation and service quality
  
- Add **selective government intervention** (Policy B elements)
  - Regional development incentives (e.g., tax breaks for interior regions)
  - Environmental regulations to prevent over-tourism
  - Infrastructure investment in under-served areas

**This is what Portugal actually does** – mostly private sector with strategic government support!

</div>

## Next Lecture

**Lecture 3** (February 12, 2026):

- Rationality and Economic Behavior
- Choices and Decision-Making
- **Opportunity Cost** (crucial concept!)

**Preparation**: Think about a recent choice you made. What did you give up to make that choice?

## Thank You! {.center}

**Questions?**

*Contact*: paulo.fagandini@ext.universidadeeuropeia.pt

---

*Next class: Wednesday, February 12, 2026*
