---
title: "Aggregation: The Concept"
subtitle: "Lecture 21: From Individual Transactions to the Economy as a Whole"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    from: markdown+emoji
    theme: [default, custom.scss]
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: false
    footer: "Economics of Tourism | Lecture 21"
    logo: ../images/logo.png
    width: 1600
    height: 900
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(eurostat)
library(tidyverse)
library(lubridate)

ue_red   <- "#C8102E"
ue_black <- "#1D1D1B"
ue_gray  <- "#6D6E71"
ue_white <- "#FFFFFF"
ue_light <- "#F5F5F5"
```

## Recap: The Macro Panorama :rewind:

Last lecture we identified the **six big macroeconomic issues**:

:::: {.columns}
::: {.column width="50%"}
:seedling: Economic growth & living standards

:factory: Productivity

:roller_coaster: Recessions & expansions

:::
::: {.column width="50%"}
:briefcase: Unemployment

:fire: Inflation

:globe_with_meridians: International interdependence
:::
::::

. . .

To study all of these, we need one fundamental tool:

:point_right: **Aggregation** — summing up millions of individual decisions into a single economy-wide picture.

# Part I: What is Aggregation?

## The Problem of Scale :microscope:

In microeconomics, we studied **one consumer**, **one firm**, **one market**.

. . .

But Portugal has:

:::: {.columns}
::: {.column width="50%"}
:busts_in_silhouette: ~10 million consumers

:factory: ~1.3 million registered firms

:handshake: Billions of transactions per year

:hotel: ~8,000 hotels and similar accommodations
:::
::: {.column width="50%"}
We cannot track every transaction. We need a way to **compress** all of this information into a manageable set of numbers.

:point_right: That is exactly what **aggregation** does.
:::
::::

## What is Aggregation? :jigsaw:

::: {.key-concept}
**Aggregation** is the process of **summing individual economic variables** — the spending decisions of millions of households, the output of thousands of firms — to obtain **totals for the economy as a whole**.
:::

. . .

**Examples:**

| Individual variable | Aggregate variable |
|---|---|
| Ana spends €80/month on restaurant meals | Household consumption |
| Pestana Hotels invests in a new resort | Business investment |
| Portuguese government builds a new highway | Government expenditure |
| A Lisbon hotel sells rooms to German tourists | Exports of services |

## Why Does Aggregation Work? :thinking:

Aggregation works because — even though individuals differ — their **combined behaviour** follows **predictable patterns**.

. . .

:::: {.columns}
::: {.column width="55%"}
Think of a **thermometer** :thermometer:

You cannot track every molecule of air in this room individually. But the **average kinetic energy** of all molecules gives you a reliable, useful number: **temperature**.

Similarly, we cannot track every euro spent in Portugal. But the **sum of all spending** gives us a reliable number: **GDP**.
:::
::: {.column width="45%"}
:bulb: **Key insight:**

Individual details get "averaged out". What remains is the **systematic, predictable** part of economic behaviour — which is exactly what policy needs to respond to.

:warning: **The cost:** Aggregation hides **distribution** — who gets what.
:::
::::

## The Trade-off of Aggregation :balance_scale:

:::: {.columns}
::: {.column width="50%"}
:heavy_check_mark: **What aggregation gives us:**

- A manageable picture of the whole economy
- Comparisons across time and countries
- Targets for economic policy
- Early warning indicators (recessions, inflation)
:::
::: {.column width="50%"}
:warning: **What aggregation hides:**

- Inequality: who earns the income?
- Regional differences: Lisbon vs interior Portugal
- Sectoral differences: tourism booming while manufacturing shrinks
- Quality of jobs: full-time vs precarious seasonal work
:::
::::

. . .

:point_right: A country's GDP can grow while most workers see their wages **fall** — if growth is concentrated at the top. Aggregation alone does not tell us this.

# Part II: The Circular Flow of Income

## Economies as a Flow :recycle:

The best way to understand aggregation is through the **circular flow of income** model.

. . .

::: {.key-concept}
The **circular flow model** shows how money, goods, and services continuously flow between the main economic agents — households, firms, the government, and the rest of the world.
:::

. . .

We will build it up **step by step**, starting simple and adding complexity.

## Step 1: Two Agents — Households & Firms :arrows_counterclockwise:

:::: {.columns}
::: {.column width="55%"}
In the simplest economy:

**Households** :family:

- Own the factors of production (labour, capital, land)
- Sell them to firms → earn **income**
- Spend that income on goods and services → **consumption**

**Firms** :factory:

- Buy factors of production → pay **wages, rents, profits**
- Produce goods and services → earn **revenue**

:point_right: Money flows in a **circle**: from firms to households (as income), and back from households to firms (as spending).
:::
::: {.column width="45%"}
```{r}
#| fig-width: 5
#| fig-height: 5

# Simple circular flow diagram using ggplot2 arrows
library(grid)

df_boxes <- data.frame(
  x    = c(0.5, 0.5),
  y    = c(0.82, 0.18),
  lab  = c("HOUSEHOLDS\n(Families)", "FIRMS\n(Companies)"),
  stringsAsFactors = FALSE
)

ggplot() +
  # boxes
  geom_tile(data = df_boxes, aes(x = x, y = y),
            width = 0.32, height = 0.14,
            fill = ue_red, color = NA) +
  geom_text(data = df_boxes, aes(x = x, y = y, label = lab),
            color = "white", fontface = "bold", size = 4.5) +
  # top arrow left: goods & services (firms → households)
  annotate("segment", x = 0.34, xend = 0.18, y = 0.77, yend = 0.77,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = ue_black, linewidth = 1) +
  annotate("text", x = 0.26, y = 0.83,
           label = "Goods &\nServices", size = 3.2, color = ue_black, fontface = "italic") +
  # top arrow right: spending (households → firms)
  annotate("segment", x = 0.66, xend = 0.82, y = 0.77, yend = 0.77,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = ue_red, linewidth = 1) +
  annotate("text", x = 0.74, y = 0.83,
           label = "Spending\n(C)", size = 3.2, color = ue_red, fontface = "bold") +
  # bottom arrow left: factors of prod (households → firms)
  annotate("segment", x = 0.34, xend = 0.18, y = 0.23, yend = 0.23,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = ue_black, linewidth = 1) +
  annotate("text", x = 0.26, y = 0.17,
           label = "Labour,\nCapital, Land", size = 3.2, color = ue_black, fontface = "italic") +
  # bottom arrow right: income (firms → households)
  annotate("segment", x = 0.66, xend = 0.82, y = 0.23, yend = 0.23,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = ue_red, linewidth = 1) +
  annotate("text", x = 0.74, y = 0.17,
           label = "Wages,\nProfits, Rents", size = 3.2, color = ue_red, fontface = "bold") +
  # left arc connecting
  annotate("segment", x = 0.18, xend = 0.18, y = 0.77, yend = 0.23,
           color = ue_black, linewidth = 0.8, linetype = "dashed") +
  # right arc connecting
  annotate("segment", x = 0.82, xend = 0.82, y = 0.77, yend = 0.23,
           color = ue_red, linewidth = 0.8, linetype = "dashed") +
  # labels for markets
  annotate("text", x = 0.12, y = 0.5,
           label = "GOODS\nMARKET", size = 3, color = ue_gray, angle = 90) +
  annotate("text", x = 0.88, y = 0.5,
           label = "FACTOR\nMARKET", size = 3, color = ue_gray, angle = 90) +
  xlim(0.05, 0.95) + ylim(0.05, 0.95) +
  theme_void() +
  theme(plot.background = element_rect(fill = ue_light, color = NA))
```
:::
::::

## Step 2: Adding the Government :classical_building:

In a real economy, the **government** plays a major role:

. . .

:::: {.columns}
::: {.column width="50%"}
**Government collects:**

:moneybag: **Taxes** — from households (income tax) and firms (corporate tax)

**Government spends:**

:school: Public services — education, health, infrastructure

:handshake: **Transfer payments** — pensions, unemployment benefits (money back to households without a service in return)
:::
::: {.column width="50%"}
**Key implication:**

Not all household income is spent on consumption. Some leaks out as **taxes**.

Not all firm revenue stays in the private sector. Government **injects** spending back in.

:point_right: The flow is still circular, but the government acts as a **redistribution node**.
:::
::::

## Step 3: Adding the Rest of the World :globe_with_meridians:

Real economies also trade with the outside world:

. . .

:::: {.columns}
::: {.column width="50%"}
**Exports** :outbox_tray:

Foreigners buy Portuguese goods and services

:point_right: Money **flows in** to Portugal

For tourism: foreign visitors spending in Portuguese hotels, restaurants, attractions = **export of services**

**Imports** :inbox_tray:

Portuguese households and firms buy foreign goods

:point_right: Money **flows out** of Portugal
:::
::: {.column width="50%"}
::: {.key-concept}
**Net Exports (NX)**

$$NX = \text{Exports} - \text{Imports}$$

If NX > 0: more money flowing **in** than out (trade surplus)

If NX < 0: more money flowing **out** than in (trade deficit)
:::

Portugal typically runs a **services surplus** (tourism!) but a **goods deficit**.
:::
::::

## The Full Circular Flow :earth_africa:

:::: {.columns}
::: {.column width="58%"}
The complete model has **four agents** and **four types of spending**:

| Agent | Type of spending | Symbol |
|---|---|---|
| Households | Consumption | C |
| Firms | Investment | I |
| Government | Government expenditure | G |
| Rest of world | Net exports | NX |

. . .

**The fundamental identity:**

$$Y = C + I + G + NX$$

:point_right: Every euro of output produced in Portugal must be **purchased** by someone — this equation always holds, by definition. This is the **expenditure approach** to GDP (next lecture).
:::
::: {.column width="42%"}
**Leakages vs Injections**

| Leakages :outbox_tray: | Injections :inbox_tray: |
|---|---|
| Savings (S) | Investment (I) |
| Taxes (T) | Government spending (G) |
| Imports (M) | Exports (X) |

In equilibrium:

$$S + T + M = I + G + X$$

:bulb: What leaks out of the circular flow must be matched by injections back in.
:::
::::

# Part III: Aggregating Output — What Counts?

## The Value-Added Principle :link:

When we aggregate output, we face a problem: **double counting**.

. . .

**Example — a tourist breakfast in Porto:**

| Stage | Seller | Revenue | Cost of inputs | **Value Added** |
|---|---|---|---|---|
| Wheat farm | Farmer | €0.50 | €0 | **€0.50** |
| Flour mill | Miller | €1.20 | €0.50 | **€0.70** |
| Bakery | Baker | €2.00 | €1.20 | **€0.80** |
| **Total GDP contribution** | | | | **€2.00** |

. . .

:point_right: We **do not** add €0.50 + €1.20 + €2.00 = €3.70 (that would count the bread three times).

We add only the **value added** at each stage: €0.50 + €0.70 + €0.80 = **€2.00**.

## What Gets Included in Aggregation? :white_check_mark:

:::: {.columns}
::: {.column width="50%"}
:heavy_check_mark: **Included:**

- Final goods and services (sold to end users)
- Government services (at cost)
- New housing construction
- Changes in business inventories
- Exports of goods and services

:hotel: *A German tourist staying at a Lisbon hotel: the room rate is included as an **export of services**.*
:::
::: {.column width="50%"}
:x: **Not included:**

- Intermediate goods (used to produce other goods)
- Sales of existing assets (second-hand house sales)
- Financial transactions (buying shares)
- Transfer payments (pensions, subsidies)
- Unpaid work (home cooking, volunteering)

:thinking: *Why not second-hand houses? The house was already counted when it was first built.*
:::
::::

## The Tourism Aggregation Problem :beach_umbrella:

Tourism is a **cross-cutting industry** — it does not have a single production line. This makes aggregation tricky.

. . .

:::: {.columns}
::: {.column width="55%"}
When a British tourist spends €1,000 in Portugal:

- €350 at hotels :hotel: → counted under **accommodation services**
- €200 at restaurants :fork_and_knife: → counted under **food & beverage**
- €150 on transport :bus: → counted under **transport services**
- €200 on retail :shopping_bags: → counted under **retail trade**
- €100 on cultural attractions :performing_arts: → counted under **arts & entertainment**

:point_right: That €1,000 is **spread across multiple sectors** in the national accounts. There is no single "tourism line" in GDP — it must be reconstructed using **Tourism Satellite Accounts (TSA)**.
:::
::: {.column width="45%"}
```{r}
#| fig-width: 4.8
#| fig-height: 4.5

# Hypothetical share breakdown — for illustration only
# Labelled clearly as illustrative
shares <- data.frame(
  sector = c("Accommodation", "F&B", "Transport", "Retail", "Culture & Other"),
  pct    = c(35, 20, 15, 20, 10)
)

ggplot(shares, aes(x = reorder(sector, pct), y = pct)) +
  geom_col(fill = ue_red, alpha = 0.85) +
  geom_text(aes(label = paste0(pct, "%")),
            hjust = -0.15, size = 3.8, color = ue_black) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 42)) +
  labs(
    title = "Tourist Spending by Sector",
    subtitle = "Hypothetical breakdown — illustrative only",
    x = NULL, y = "Share of total tourist spending (%)",
    caption = "Hypothetical data for illustration"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(color = ue_red, face = "bold"),
    plot.subtitle = element_text(color = ue_gray, size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )
```
:::
::::

## Tourism's Share in the Portuguese Economy :portugal:

```{r}
#| fig-width: 11
#| fig-height: 4.2

# Travel receipts as % of GDP — World Bank / Eurostat bop_its6_det
# Using Eurostat BOP: tourism receipts
receipts_raw <- get_eurostat("bop_its6_det",
  filters = list(
    geo      = "PT",
    bop_item = "SB",       # Travel (credit = receipts)
    currency = "MIO_EUR",
    stk_flow = "CRE",
    partner  = "EXT_EU27_2020"
  ),
  time_format = "num"
)

gdp_raw2 <- get_eurostat("nama_10_gdp",
  filters = list(geo = "PT", na_item = "B1GQ", unit = "CP_MEUR"),
  time_format = "num"
)

receipts <- receipts_raw |>
  filter(time >= 2010) |>
  select(time, receipts = values)

gdp2 <- gdp_raw2 |>
  filter(time >= 2010) |>
  select(time, gdp = values)

travel_gdp <- inner_join(receipts, gdp2, by = "time") |>
  mutate(share = receipts / gdp * 100) |>
  filter(!is.na(share))

ggplot(travel_gdp, aes(x = time, y = share)) +
  geom_col(fill = ue_red, alpha = 0.85) +
  geom_text(aes(label = sprintf("%.1f%%", share)),
            vjust = -0.4, size = 3.2, color = ue_black) +
  labs(
    title = "Tourism Receipts as % of GDP — Portugal",
    x = NULL, y = "% of GDP",
    caption = "Source: Eurostat (bop_its6_det, nama_10_gdp). Travel receipts from non-EU partners."
  ) +
  scale_x_continuous(breaks = seq(2010, 2024, 2)) +
  scale_y_continuous(limits = c(0, max(travel_gdp$share, na.rm = TRUE) * 1.2)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(color = ue_red, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

:point_right: Tourism represents a **significant and growing share** of Portugal's GDP — making macroeconomic performance especially important for this industry.

# Part IV: Aggregate Demand and Aggregate Supply

## From Individual to Aggregate Curves :chart_with_upwards_trend:

In microeconomics we drew **individual demand** and **supply curves** for single markets.

In macroeconomics we aggregate these into:

. . .

:::: {.columns}
::: {.column width="50%"}
:shopping_cart: **Aggregate Demand (AD)**

The total demand for all goods and services in an economy at each **price level**.

$$AD = C + I + G + NX$$

Downward sloping: when prices rise, real purchasing power falls, real wealth falls, and exports become less competitive.
:::
::: {.column width="50%"}
:factory: **Aggregate Supply (AS)**

The total output all firms in an economy are willing to produce at each **price level**.

Upward sloping in the **short run**: higher prices make production more profitable.

Vertical in the **long run**: output is determined by real factors (technology, labour, capital), not prices.
:::
::::

## The AD-AS Model (Preview) :bar_chart:

```{r}
#| fig-width: 10
#| fig-height: 4.8

x <- seq(0.1, 2, length.out = 200)

ad  <- 2.5 - x
sras <- 0.8 * x + 0.3
lras_x <- 1.1

df_curves <- data.frame(
  x    = c(x, x),
  y    = c(ad, sras),
  type = rep(c("Aggregate Demand (AD)", "Short-Run AS (SRAS)"), each = 200)
) |> filter(y > 0.1, y < 2.2)

ggplot() +
  geom_line(data = df_curves, aes(x = x, y = y, color = type, linetype = type),
            linewidth = 1.3) +
  geom_vline(xintercept = lras_x, color = ue_black, linewidth = 1.3, linetype = "solid") +
  annotate("text", x = lras_x + 0.07, y = 2.0,
           label = "Long-Run AS\n(LRAS)", color = ue_black, size = 3.5, hjust = 0) +
  # equilibrium point
  annotate("point", x = 1.1, y = 2.5 - 1.1, size = 4, color = ue_red) +
  annotate("text", x = 1.17, y = 2.5 - 1.1 + 0.1,
           label = "Equilibrium\n(Y*, P*)", color = ue_red, size = 3.3) +
  annotate("segment", x = 1.1, xend = 1.1, y = 0, yend = 2.5 - 1.1,
           linetype = "dashed", color = ue_gray, linewidth = 0.6) +
  annotate("segment", x = 0, xend = 1.1, y = 2.5 - 1.1, yend = 2.5 - 1.1,
           linetype = "dashed", color = ue_gray, linewidth = 0.6) +
  scale_color_manual(values = c("Aggregate Demand (AD)" = ue_red,
                                "Short-Run AS (SRAS)"   = ue_black)) +
  scale_linetype_manual(values = c("Aggregate Demand (AD)" = "solid",
                                   "Short-Run AS (SRAS)"   = "dashed")) +
  labs(
    title = "The AD-AS Model — Preview",
    subtitle = "Hypothetical illustration — formal treatment in later lectures",
    x = "Real Output (Y)", y = "Price Level (P)",
    color = NULL, linetype = NULL,
    caption = "Hypothetical data for illustration"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(color = ue_red, face = "bold"),
    plot.subtitle = element_text(color = ue_gray, size = 10),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

:point_right: This model gives us the macroeconomic equivalent of the supply-demand diagram we used in microeconomics. We will use it throughout the macro block.

## The Three Approaches to Measuring Aggregate Output :straight_ruler:

Any economy can be measured from **three equivalent angles** — they always give the same answer:

. . .

:::: {.columns}
::: {.column width="33%"}
:outbox_tray: **Expenditure approach**

Add up all **spending** on final goods and services

$$Y = C + I + G + NX$$

*Who buys the output?*
:::
::: {.column width="33%"}
:incoming_envelope: **Income approach**

Add up all **incomes** earned in production

$$Y = W + R + \Pi + \text{taxes}$$

*(Wages + Rents + Profits)*

*Who earns from production?*
:::
::: {.column width="33%"}
:factory: **Production (value-added) approach**

Add up **value added** at each stage of production

$$Y = \sum \text{Value Added}$$

*What is produced?*
:::
::::

. . .

:bulb: All three approaches measure the **same thing** — total economic activity — just from different perspectives. Next lecture we focus on the expenditure approach: **GDP**.

# Exercises

## :pencil: Exercise 1 — Multiple Choice

**A local tour operator in the Algarve buys sunscreen bottles from a wholesaler for €4 each, and sells them to tourists for €7 each. What is the value added by the tour operator, and how much does this contribute to GDP?**

. . .

**(A)** €7 — the full selling price is GDP

**(B)** €4 — only intermediate costs count

**(C)** €3 — only the value added by the tour operator counts

**(D)** €11 — the sum of all transactions counts

::: {.fragment}
::: {.highlight-box}
**Correct answer: (C)**

The tour operator's **value added** = €7 − €4 = **€3**. GDP counts only value added at each stage to avoid double-counting. The €4 paid to the wholesaler was already (or will be) counted as the wholesaler's value added. Option A double-counts the wholesale stage; D triple-counts it.
:::
:::

## :pencil: Exercise 2 — Multiple Choice

**Which of the following transactions would NOT be included in Portugal's GDP?**

. . .

**(A)** A German tourist pays €150 to stay in a Porto hotel

**(B)** The Portuguese government pays €2,000/month in pensions to a retired teacher

**(C)** A new Airbus A320 is purchased by TAP Air Portugal for its fleet

**(D)** A Portuguese construction company builds 10 new apartments

::: {.fragment}
::: {.highlight-box}
**Correct answer: (B)**

**Transfer payments** (pensions, unemployment benefits) are **not** included in GDP — they redistribute existing income but do not correspond to new production of goods or services. Option A is an export of services (counted). Option C is investment (counted). Option D is investment in new construction (counted).
:::
:::

## :pencil: Exercise 3 — Open Question

**Consider a simplified tourism economy with three firms:**

- **Quinta do Algarve** (wine producer): sells wine to a restaurant for **€60**. Inputs from outside the economy cost **€10**.
- **Restaurante Praia** (restaurant): buys wine for €60 and other ingredients for €20, and sells meals to tourists for **€160**.
- **Hotel Maré** (hotel): builds a new outdoor terrace at a cost of **€500** (inputs purchased from local suppliers for **€300**, rest is labour).

**(a)** Calculate the **value added** by each firm. *(3 marks)*

**(b)** What is the **total contribution to GDP** of these three firms? *(2 marks)*

**(c)** To which component of GDP (C, I, G, or NX) does each transaction belong? Justify briefly. *(3 marks)*

**(d)** If the meals at Restaurante Praia were purchased exclusively by **foreign tourists**, how would your classification in (c) change? *(2 marks)*

---

::: {.fragment}

**Solution:**

**(a)**
- Quinta do Algarve: €60 − €10 = **€50**
- Restaurante Praia: €160 − €60 − €20 = **€80**
- Hotel Maré: €500 − €300 = **€200** (value of new terrace less purchased inputs)

**(b)** Total GDP contribution = €50 + €80 + €200 = **€330**

**(c)** Restaurant meals sold to domestic consumers → **C** (consumption). Hotel terrace construction → **I** (investment, as it is a new capital asset). Wine sold to the restaurant is an intermediate good → **not counted directly** (captured via value added).

**(d)** If meals are purchased by foreign tourists, they represent an **export of services** → classified under **NX** (net exports), not C. The money flows **into** Portugal from abroad.

:::

## Summary :books:

**Today we covered:**

:white_check_mark: **Aggregation** — summing individual variables into economy-wide totals

:white_check_mark: The **circular flow of income** — households, firms, government, and the rest of the world

:white_check_mark: **Leakages** (savings, taxes, imports) and **injections** (investment, government spending, exports)

:white_check_mark: The **value-added principle** — how to avoid double counting

:white_check_mark: What is and is not **included** in aggregate output

:white_check_mark: The **three approaches** to measuring output (expenditure, income, production)

. . .

**Next lecture (Lecture 22):**

:mag: GDP in depth — the expenditure approach, nominal vs real GDP, and what GDP tells us (and doesn't tell us) about well-being.

## Thank You! :wave: {.center}

**Questions?**

:email: paulo.fagandini@ext.universidadeeuropeia.pt

---
*Next class: Friday, May 8th, 2026*
