---
title: "The Balance of Payments"
subtitle: "Lecture 26: The Current Account and International Transactions"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    from: markdown+emoji
    theme: [default, custom.scss]
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: false
    footer: "Economics of Tourism | Lecture 26"
    logo: ../images/logo.png
    width: 1600
    height: 900
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(eurostat)
library(tidyverse)
library(lubridate)
library(scales)

ue_red   <- "#C8102E"
ue_black <- "#1D1D1B"
ue_gray  <- "#6D6E71"
ue_white <- "#FFFFFF"
ue_light <- "#F5F5F5"
```

## Recap: Fiscal Policy :rewind:

Last lecture: the government's lever for managing the economy:

:white_check_mark: **Expansionary** — raise G / cut taxes → stimulate AD → fight recession

:white_check_mark: **Contractionary** — cut G / raise taxes → cool AD → fight inflation or debt

:white_check_mark: The **multiplier**: one euro of government spending triggers more than one euro of GDP

:white_check_mark: **Automatic stabilisers**: unemployment benefits and tax cycles cushion downturns automatically

:white_check_mark: **Portugal's story**: Troika austerity 2011–2014 → recovery → surplus 2019 → COVID stimulus 2020

. . .

:point_right: Today — the **last lecture** of the course :mortar_board:

We zoom out to the global picture: **how do we record all of a country's international transactions?**

# Part I: What is the Balance of Payments?

## Every Transaction With the World :globe_with_meridians:

Every time a Portuguese resident interacts economically with a foreign counterpart, that transaction needs to be recorded somewhere.

. . .

:hotel: A German tourist pays €200/night at a Lisbon hotel

:airplane: A Portuguese student buys a flight on Ryanair (Irish company)

:factory: Volkswagen sends dividend payments from its Portuguese plant to Germany

:bank: A Portuguese bank borrows from a French bank

:eu: Portugal receives EU structural funds

. . .

::: {.key-concept}
The **Balance of Payments (BoP)** is a systematic record of **all economic transactions** between residents of a country and the rest of the world over a given period of time. It always **balances** — every transaction has a debit and a credit entry.
:::

## The Structure of the Balance of Payments :bar_chart:

The BoP is divided into three main accounts:

. . .

:::: {.columns}
::: {.column width="33%"}
:shopping_cart: **Current Account**

Records flows of goods, services, income, and transfers

- Trade in goods
- Trade in services *(tourism is here)*
- Primary income (wages, dividends)
- Secondary income (remittances, aid)
:::
::: {.column width="33%"}
:construction: **Capital Account**

Records transfers of non-produced, non-financial assets

- EU structural funds & grants
- Debt forgiveness
- Relatively small for most countries

:::
::: {.column width="33%"}
:bank: **Financial Account**

Records flows of financial assets and liabilities

- Foreign Direct Investment (FDI)
- Portfolio investment (stocks, bonds)
- Other investment (bank loans)
- Reserve assets
:::
::::

. . .

::: {.key-concept}
**The fundamental identity:**
$$\text{Current Account} + \text{Capital Account} + \text{Financial Account} = 0$$
Every euro that leaves the current account must come back through the financial account — the BoP always balances.
:::

# Part II: The Current Account in Depth

## Current Account Components :receipt:

::: {.key-concept}
The **Current Account** records **all flows of real resources** (goods, services, income) between a country and the rest of the world.
$$CA = \text{Trade in Goods} + \text{Trade in Services} + \text{Primary Income} + \text{Secondary Income}$$
:::

. . .

| Component | Credit (inflow) | Debit (outflow) |
|---|---|---|
| **Goods** | Exports of goods | Imports of goods |
| **Services** | Tourism receipts, transport | Tourism abroad, freight |
| **Primary income** | Wages earned abroad, dividends received | Profits sent to foreign owners |
| **Secondary income** | Remittances received, EU transfers | Remittances sent abroad, contributions |

. . .

:beach_umbrella: **Tourism sits in "Services"** — one of the most important items in Portugal's current account.

## Trade in Goods vs Trade in Services :package:

:::: {.columns}
::: {.column width="50%"}
:package: **Trade in goods** (the "visible trade balance")

Physical products that cross borders:
- Portugal **exports**: wine, cork, footwear, machinery, olive oil
- Portugal **imports**: cars, petroleum, electronics, machinery

Portugal typically runs a **goods trade deficit** — imports more goods than it exports.
:::
::: {.column width="50%"}
:handshake: **Trade in services** (the "invisible trade balance")

Intangible services:
- Portugal **exports services**: tourism (foreigners visiting), transport, financial services
- Portugal **imports services**: travel abroad by Portuguese residents, shipping

Portugal runs a **services trade surplus** — and **tourism is the main reason**.

:point_right: Tourism receipts are one of Portugal's most powerful economic engines, consistently offsetting the goods deficit.
:::
::::

## Current Account Balance — Portugal :portugal:

```{r}
#| fig-width: 11
#| fig-height: 4.6

# Current account components as % of GDP — Eurostat bop_eu6_q annual
# Use bop_c6_q or bop_gdp6_q for annual CA data
# Most reliable: nama_10_gdp for NX, and separate CA from bop
# Using tipsii10 — current account balance as % of GDP (EDP)
ca_raw <- get_eurostat("tipsbp20",
  filters = list(geo = "PT"),
  time_format = "num"
) |>
  filter(time >= 2000, !is.na(values)) |>
  arrange(time)

ggplot(ca_raw, aes(x = time, y = values)) +
  geom_col(aes(fill = values >= 0), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "#5B8DB8", "FALSE" = ue_red)) +
  geom_hline(yintercept = 0, color = ue_black, linewidth = 0.5) +
  annotate("rect", xmin = 2010.5, xmax = 2014.5,
           ymin = -Inf, ymax = Inf, fill = ue_gray, alpha = 0.1) +
  annotate("text", x = 2012.5, y = -11,
           label = "Troika\nadjustment", color = ue_gray,
           size = 3, fontface = "italic") +
  annotate("rect", xmin = 2020, xmax = 2021,
           ymin = -Inf, ymax = Inf, fill = ue_gray, alpha = 0.1) +
  annotate("text", x = 2020.5, y = -11,
           label = "COVID", color = ue_gray,
           size = 3, fontface = "italic") +
  labs(
    title = "Portugal: Current Account Balance (% of GDP)",
    subtitle = "Blue = surplus, Red = deficit",
    x = NULL, y = "% of GDP",
    caption = "Source: Eurostat (tipsbp20)"
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = ue_red, face = "bold"),
    plot.subtitle    = element_text(color = ue_gray, size = 10),
    panel.grid.minor = element_blank()
  )
```

:point_right: Portugal ran persistent **current account deficits** before 2013 — a key driver of the debt crisis. The adjustment to surplus was painful but necessary.

# Part III: Tourism in the Balance of Payments

## Tourism as an Export of Services :airplane:

This is one of the most important conceptual points of the course:

::: {.key-concept}
When a **foreign tourist spends money in Portugal**, that spending counts as an **export of services** — money flows **into** Portugal from abroad, appearing as a **credit** in the current account.

When a **Portuguese resident travels abroad**, that spending counts as an **import of services** — money flows **out** of Portugal, appearing as a **debit** in the current account.
:::

. . .

:::: {.columns}
::: {.column width="50%"}
**Inbound tourism = export of services**

:de: German tourist in Algarve: spends €1,500

:gb: British tourist in Lisbon: spends €800

:us: American tourist in Porto: spends €2,000

→ All of these **increase Portugal's current account credit**

:point_right: Portugal does not need to ship anything — the tourist comes to us. Tourism is an "invisible export."
:::
::: {.column width="50%"}
**Outbound tourism = import of services**

:portugal: Portuguese family holidays in Spain: spends €1,200

:portugal: Portuguese student studies in UK: pays tuition €15,000

→ These **increase Portugal's current account debit**

:bulb: **Net tourism balance** = inbound receipts − outbound expenditure

Portugal's net tourism balance is **strongly positive** — we receive far more from foreign visitors than we spend abroad.
:::
::::

## Portugal's Tourism in the Current Account :chart_with_upwards_trend:

```{r}
#| fig-width: 11
#| fig-height: 4.5

# Trade in services — goods and services from nama_10_gdp for context
# For actual tourism share, use the exports of services from nama_10_gdp
# P6 = exports of goods and services total, already validated
# For services specifically, use bop via trade in services table

# Use validated dataset: nama_10_gdp P6 and P7 (total exports/imports) vs GDP
# and supplement with HICP services component as narrative support

# Simple validated approach: exports vs imports of goods+services as % of GDP
trade_raw <- get_eurostat("nama_10_gdp",
  filters = list(
    geo     = "PT",
    na_item = c("P6", "P7", "B1GQ"),
    unit    = "CP_MEUR"
  ),
  time_format = "num"
) |>
  filter(time >= 2000, !is.na(values)) |>
  select(time, na_item, values) |>
  pivot_wider(names_from = na_item, values_from = values) |>
  filter(!is.na(P6), !is.na(P7), !is.na(B1GQ)) |>
  mutate(
    exports_gdp = P6 / B1GQ * 100,
    imports_gdp = P7 / B1GQ * 100,
    nx_gdp      = (P6 - P7) / B1GQ * 100
  ) |>
  select(time, exports_gdp, imports_gdp, nx_gdp) |>
  pivot_longer(cols = c(exports_gdp, imports_gdp),
               names_to = "flow", values_to = "value") |>
  mutate(flow = case_when(
    flow == "exports_gdp" ~ "Exports (X)",
    flow == "imports_gdp" ~ "Imports (M)"
  ))

ggplot(trade_raw, aes(x = time, y = value, color = flow)) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = c(
    "Exports (X)" = ue_red,
    "Imports (M)" = ue_black
  )) +
  labs(
    title = "Portugal: Total Exports and Imports of Goods & Services (% of GDP)",
    x = NULL, y = "% of GDP", color = NULL,
    caption = "Source: Eurostat (nama_10_gdp). Includes goods and services."
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = ue_red, face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

## Why Tourism Matters So Much for Portugal :portugal:

:::: {.columns}
::: {.column width="55%"}
Tourism is Portugal's **most important single export** — larger than any manufactured good.

**What makes Portugal's tourism balance so powerful:**

:one: **High receipts per visitor** — Lisbon and Algarve attract high-spending segments

:two: **Low import content** — Portuguese hospitality relies heavily on **domestic** food, wine, labour, and culture. Unlike manufacturing, it doesn't require importing lots of components.

:three: **Year-round growth** — pre-COVID, tourism demand was growing every year (2013–2019 was a golden period)

:four: **Services deficit elsewhere** — Portugal pays for freight, insurance, and financial services imported from abroad; tourism is the main offset

:point_right: When tourism collapses (as in 2020), Portugal's **current account deteriorates sharply** — as clearly visible in the chart.
:::
::: {.column width="45%"}
::: {.key-concept}
**The structural importance:**

Portugal's historical **current account deficit** (pre-2013) was driven by importing more goods than it exported.

The **correction** came through:

1. Painful reduction of imports (recession cut demand)
2. **Massive growth of tourism exports** (2013–2019)

Tourism — as a service export — was essential to Portugal's economic rebalancing. This is a live example of the macroeconomic role tourism plays.
:::
:::
::::

# Part IV: Exchange Rates and the Current Account

## The Exchange Rate :currency_exchange:

::: {.key-concept}
The **exchange rate** is the price of one currency in terms of another.

A **depreciation** of the euro (€ becomes cheaper relative to £, $, etc.) → Portugal becomes cheaper for foreign visitors → **tourism exports rise**

An **appreciation** of the euro (€ becomes more expensive) → Portugal becomes pricier for non-euro visitors → **tourism exports fall**
:::

. . .

:::: {.columns}
::: {.column width="50%"}
**Portugal in the Eurozone:**

Since 1999, Portugal uses the euro — **no independent exchange rate**.

The euro/sterling and euro/dollar exchange rates are determined by ECB policy and market forces, not by Portugal alone.

:point_right: A weak euro (as in 2022–2023) **boosted** UK and US visitor numbers in Portugal — holidays became cheaper in pound and dollar terms.
:::
::: {.column width="50%"}
**Real vs Nominal Exchange Rate:**

The **nominal** rate is the raw currency ratio.

The **real** exchange rate adjusts for inflation differences:

$$q = \frac{e \times P^*}{P}$$

Where $e$ = nominal rate, $P^*$ = foreign prices, $P$ = domestic prices.

:point_right: If Portuguese inflation is higher than UK inflation, Portugal becomes **relatively more expensive** for British tourists — even if the nominal exchange rate is unchanged.
:::
::::

## Exchange Rates and Tourism Flows :airplane:

```{r}
#| fig-width: 11
#| fig-height: 4.4

# EUR/GBP and EUR/USD exchange rates from Eurostat ert_bil_eur_a
fx_raw <- get_eurostat("ert_bil_eur_a",
  filters = list(
    currency = c("GBP", "USD"),
    statinfo = "AVG"
  ),
  time_format = "num"
) |>
  filter(time >= 2000, !is.na(values)) |>
  mutate(label = case_when(
    currency == "GBP" ~ "EUR/GBP (€ per £)",
    currency == "USD" ~ "EUR/USD (€ per $)"
  ))

ggplot(fx_raw, aes(x = time, y = values, color = label)) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = c(
    "EUR/GBP (€ per £)" = ue_red,
    "EUR/USD (€ per $)" = ue_black
  )) +
  geom_hline(yintercept = 1, linetype = "dashed",
             color = ue_gray, linewidth = 0.6) +
  annotate("text", x = 2001.5, y = 1.02,
           label = "Parity (1:1)", color = ue_gray,
           size = 3, fontface = "italic") +
  labs(
    title = "EUR/GBP and EUR/USD Exchange Rates (annual average)",
    subtitle = "Higher value = euro is stronger (more expensive for UK/US visitors)",
    x = NULL, y = "Euros per foreign currency unit",
    color = NULL,
    caption = "Source: Eurostat (ert_bil_eur_a)"
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = ue_red, face = "bold"),
    plot.subtitle    = element_text(color = ue_gray, size = 10),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

:point_right: Brexit (2016) caused a **sharp GBP depreciation** — British tourists found Portugal more expensive overnight. A key example of exchange rate risk for tourism.

# Part V: Tying It All Together

## The Macro Block — What We Have Learned :books:

Over the last seven lectures, we have built a complete macroeconomic framework:

. . .

:::: {.columns}
::: {.column width="50%"}
**Measurement tools:**

:white_check_mark: **Aggregation** — summing individual transactions into economy-wide totals

:white_check_mark: **GDP** — the economy's total output (C+I+G+NX)

:white_check_mark: **Inflation (CPI/HICP)** — the general price level

:white_check_mark: **Balance of Payments** — recording all international transactions
:::
::: {.column width="50%"}
**Policy tools:**

:white_check_mark: **Monetary policy (ECB)** — interest rates control credit and spending

:white_check_mark: **Fiscal policy (Government)** — taxes and spending control aggregate demand

**Key tensions:**

:arrows_counterclockwise: Growth vs inflation

:arrows_counterclockwise: Stimulus vs debt sustainability

:arrows_counterclockwise: Competitiveness vs price stability
:::
::::

## Tourism Through Every Macro Lens :beach_umbrella:

| Macro concept | Tourism connection |
|---|---|
| **GDP** | Tourism receipts from abroad = export of services, part of NX |
| **Inflation** | Hotel/restaurant CPI tracked directly; cost competitiveness |
| **Unemployment** | Tourism creates 9%+ of Portuguese employment |
| **Monetary policy** | Low rates → investment boom; high rates → investment pause |
| **Fiscal policy** | Tourist taxes; COVID subsidies; VAT on hospitality |
| **Current account** | Tourism = Portugal's largest service export; key to external balance |
| **Exchange rates** | Euro strength vs £ and $ directly affects inbound visitor numbers |

. . .

:point_right: Tourism is not just an industry — it is a **macroeconomic phenomenon** that appears in every major indicator we have studied.

## Course Wrap-Up :mortar_board: {.center}

:::: {.columns}
::: {.column width="50%"}
**What you can now do:**

:microscope: Apply **microeconomic reasoning** — consumer choice, producer decisions, market equilibrium

:chess_pawn: Think **strategically** — game theory, Nash equilibrium, backward induction

:telescope: Read **macroeconomic data** — GDP, inflation, current account, policy stances

:beach_umbrella: Connect it all to **tourism** — the industry you are training for
:::
::: {.column width="50%"}
**What comes next:**

:calendar: **Test 2** — May 28th

Covers: Producer theory (Lectures 11–17) + Macroeconomics (Lectures 20–26)

:busts_in_silhouette: **Group Presentations** — May 29th

:books: For the **final exam**: the whole course

:muscle: You've covered more economics in one semester than most business graduates do in a full year. Well done.
:::
::::

# Exercises

## :pencil: Exercise 1 — Multiple Choice

**A British travel company brings 5,000 tourists to Portugal each summer, each spending an average of €1,200 in the country. From the perspective of Portugal's Balance of Payments, how are these transactions classified?**

. . .

**(A)** As imports of services in Portugal's current account — Portugal is buying tourism from the UK

**(B)** As exports of services in Portugal's current account — Portugal is providing tourism services to foreigners

**(C)** As Foreign Direct Investment in Portugal's financial account — British money entering the Portuguese economy

**(D)** As secondary income in Portugal's current account — it represents a transfer from the UK to Portugal

::: {.fragment}
::: {.highlight-box}
**Correct answer: (B)**

When foreigners spend money on services provided in Portugal, it is an **export of services** — Portugal is "selling" tourism to the UK. This creates a **credit** in the services component of the current account, improving Portugal's current account balance. Option A confuses the direction. Option C would apply if the British company was buying a Portuguese hotel. Option D (secondary income) covers transfers like remittances, not market transactions.
:::
:::

## :pencil: Exercise 2 — Multiple Choice

**In 2016, the UK voted to leave the EU (Brexit). The British pound immediately depreciated approximately 15% against the euro. Which of the following best describes the likely short-run effect on Portugal's tourism sector?**

. . .

**(A)** Positive — British tourists now find Portugal 15% cheaper, increasing inbound demand

**(B)** Negative — British tourists find Portugal 15% more expensive, reducing inbound demand

**(C)** Neutral — exchange rate changes do not affect tourism as prices are set in local currency

**(D)** Positive — a weaker pound means Portuguese tourists can holiday more cheaply in the UK, boosting Portugal's imports

::: {.fragment}
::: {.highlight-box}
**Correct answer: (B)**

A GBP depreciation means each pound buys **fewer euros**. From a British tourist's perspective, Portuguese prices (in euros) now cost **more pounds**. A €100 hotel night that cost £85 before depreciation now costs £100. This reduces the real purchasing power of British visitors in Portugal, lowering demand for Portuguese tourism. Option A gets the direction backwards. Option D describes an increase in Portuguese imports (outbound tourism), which worsens Portugal's current account — the opposite of helping the tourism sector.
:::
:::

## :pencil: Exercise 3 — Open Question

**The table below shows simplified Balance of Payments data for Portugal in a given year (€ billion):**

| Item | Value (€bn) |
|---|---|
| Goods exports | 65 |
| Goods imports | −85 |
| Services exports (incl. tourism) | 45 |
| Services imports | −22 |
| Primary income (net) | −8 |
| Secondary income (net) | +4 |
| Capital account (net) | +3 |
| Financial account (net) | ? |

**(a)** Calculate the **trade in goods balance**, the **trade in services balance**, and the **current account balance**. *(4 marks)*

**(b)** Using the fundamental BoP identity, calculate the **financial account balance**. *(2 marks)*

**(c)** Tourism receipts account for **€28bn** of the services exports figure. Calculate tourism's share of total goods and services exports, and briefly explain what this implies for Portugal's economic vulnerability. *(4 marks)*

---

::: {.fragment}

**Solution:**

**(a)**

Trade in goods balance = 65 − 85 = **−€20bn** (goods deficit)

Trade in services balance = 45 − 22 = **+€23bn** (services surplus)

Current account = (65−85) + (45−22) + (−8) + 4 = −20 + 23 − 8 + 4 = **−€1bn** (small deficit)

**(b)**

BoP identity: Current Account + Capital Account + Financial Account = 0

−1 + 3 + Financial Account = 0 → Financial Account = **−€2bn**

(A negative financial account = net capital outflow, meaning Portugal is acquiring more foreign assets than foreigners are acquiring Portuguese assets — consistent with a small current account deficit.)

**(c)**

Total goods + services exports = 65 + 45 = **€110bn**

Tourism share = 28 / 110 × 100 = **25.5%**

This implies significant **concentration risk**: a quarter of all export earnings comes from a single sector that is highly sensitive to external shocks (pandemics, exchange rates, terrorism, climate). When tourism collapses (as in 2020), Portugal's current account deteriorates sharply and the economy loses a critical income source. Diversifying the export base — developing manufacturing, tech, and other services — would reduce this vulnerability.

:::

## Summary :books:

**Today we covered:**

:white_check_mark: The **Balance of Payments** — a complete record of all international economic transactions

:white_check_mark: Three accounts: **Current** (goods, services, income), **Capital** (grants, transfers), **Financial** (investment flows)

:white_check_mark: The **fundamental identity**: CA + KA + FA = 0

:white_check_mark: **Tourism as an export of services** — foreign spending in Portugal = credit in current account

:white_check_mark: Portugal's **current account history** — chronic deficit pre-2013, adjustment via tourism growth

:white_check_mark: **Exchange rates** — EUR/GBP and EUR/USD directly affect inbound tourism competitiveness

:white_check_mark: Brexit as a live example of exchange rate risk for tourism

:trophy: **The macro block is complete!**

## Thank You — and Good Luck! :wave: :four_leaf_clover: {.center}

**Test 2: May 28th**

Producer theory (Lectures 11–17) + Macroeconomics (Lectures 20–26)

**Group Presentations: May 29th**

. . .

:email: paulo.fagandini@ext.universidadeeuropeia.pt

*"An economist is someone who sees something working in practice and wonders if it would work in theory."* — Ronald Reagan
