---
title: "Economics Fundamentals"
subtitle: "Lecture 3: Rationality, Choices and Opportunity Cost"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    theme: default
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: true
    footer: "Economics of Tourism | Lecture 3"
    logo: ../images/logo.png
---

## Recap: Previous Lectures {.center}

**Lecture 1**: Economics, Scarcity, Efficiency, Macro vs. Micro

**Lecture 2**: Three economic problems (WHAT, HOW, FOR WHOM) and economic systems (Market, Planned, Mixed)

**Today's Focus**:

- How do people make economic decisions?
- What is rationality?
- **Opportunity cost** – one of economics' most powerful concepts!

## Economic Rationality
<div style="font-size: 80%;">


**Assumption**: People are **rational decision-makers**

**What does "rational" mean in economics?**

- People have **clear preferences**
- They seek to **maximize their well-being** (utility)
- They respond **predictably** to incentives
- They use **cost-benefit analysis**

**Important**: Rational ≠ Perfect or emotionless!

People can be rational within their information and cognitive limits (bounded rationality)

</div>

## Cost-Benefit Principle

<div style="font-size: 80%;">

**The Foundation of Rational Choice**:

> An action should be undertaken **if and only if** the benefits exceed the costs.

$$\text{Decision Rule: } \begin{cases} 
\text{Do it} & \text{if } Benefits \geq Costs \\
\text{Don't do it} & \text{if } Benefits < Costs
\end{cases}$$

**In practice**:

- Benefits = What you gain
- Costs = What you give up
- Compare net benefit of alternatives

</div>

## Tourism Example: Cost-Benefit

<div style="font-size: 80%;">

**Should a tourist visit the Belém Tower in Lisbon?**

:::: {.columns}
::: {.column width="50%"}
### Benefits
- Historical experience
- Photos/memories
- Educational value
- Satisfaction

**Estimated value**: €20
:::

::: {.column width="50%"}
### Costs
- Entry ticket: €6
- Travel time: 30 min
- Waiting time: 20 min
- Value of time: €15/hr

**Total time cost**: €12.50
**Total cost**: €18.50
:::
::::

**Decision**: Benefits (€20) > Costs (€18.50) → **Visit!**

But if waiting time increases to 60 min, costs rise to €28.50 → **Don't visit!**

</div>

## Marginal Analysis

**Key Concept**: Decisions are made **at the margin**

**Marginal** = Additional/incremental change from one more unit

- **Marginal Benefit (MB)**: Benefit from one additional unit
- **Marginal Cost (MC)**: Cost of one additional unit

**Optimal Decision Rule**:

$$\text{Continue activity while } MB \geq MC$$

Stop when $MB < MC$

## Marginal Analysis: Hotel Rooms

**Should a hotel accept one more booking?**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

library(ggplot2)

# Marginal cost and price for hotel rooms
rooms <- seq(50, 200, by = 10)

# Marginal cost increases with occupancy (more cleaning, energy, wear)
marginal_cost <- 30 + 0.3 * rooms

# Market price (what customers pay)
market_price <- rep(80, length(rooms))

cost_data <- data.frame(
  rooms = rep(rooms, 2),
  value = c(marginal_cost, market_price),
  type = rep(c("Marginal Cost", "Market Price"), each = length(rooms))
)

ggplot(cost_data, aes(x = rooms, y = value, color = type, linetype = type)) +
  geom_line(size = 1.5) +
  labs(title = "Hotel Room Pricing: Marginal Analysis",
       subtitle = "Accept bookings while Price > Marginal Cost",
       x = "Number of Rooms Booked",
       y = "Price/Cost (€)",
       color = "",
       linetype = "") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Marginal Cost" = "#e74c3c", 
                                 "Market Price" = "#27ae60")) +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 167, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 167, y = 90, 
           label = "Optimal: ~167 rooms\n(P = MC)", size = 4) +
  annotate("rect", xmin = 50, xmax = 167, ymin = 0, ymax = 120, 
           alpha = 0.1, fill = "green") +
  annotate("text", x = 100, y = 110, label = "Accept\n(P > MC)", size = 4) +
  annotate("rect", xmin = 167, xmax = 200, ymin = 0, ymax = 120, 
           alpha = 0.1, fill = "red") +
  annotate("text", x = 185, y = 110, label = "Reject\n(P < MC)", size = 4)
```

## Opportunity Cost: Definition
<div style="font-size: 80%;">

**The Most Important Concept in Economics!**

> **Opportunity Cost** = The value of the best alternative foregone when making a choice

**Key Points**:

- It's what you **give up**, not what you pay
- It's the **next-best** alternative (not all alternatives)
- It includes both **explicit** (money) and **implicit** costs (time, foregone opportunities)

</div>

## Opportunity Cost: Definition

<div style="font-size: 80%;">

**When deciding**: You should choose option A over B if:

$$\overbrace{B_A}^{\text{Gross Benefit of}\ A} \geq \underbrace{\overbrace{C_A}^{\text{Monetary Cost}} + \underbrace{(B_B-C_b)}_{\text{Surplus of best alternative}}}_{\text{Opportunity Cost of}\ A}$$

</div>

## Opportunity Cost Examples

<div style="font-size: 80%;">

**Simple Example**: You have €100 and 4 hours free on Saturday afternoon.

**Options**:

A. Visit museum (€10 entry, 3 hours)  
B. Beach trip (€0 entry, 4 hours)  
C. Movie (€12 ticket, 2 hours)  
D. Stay home and study (€0, 4 hours)

**If you choose A (museum)**:

- **Explicit cost**: €10
- **Opportunity cost**: Value of best alternative foregone
  - Can't go to beach (assume you value it at €20)
  - **Total opportunity cost**: €10 paid + €20 value foregone = **€30**

</div>

## Opportunity Cost in Tourism Decisions

<div style="font-size: 80%;">

**Example: Hotel Investment**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Investment alternatives
investment_options <- data.frame(
  option = c("Build Hotel\nin Lisbon", "Build Hotel\nin Algarve", "Invest in\nStock Market", "Government\nBonds"),
  initial_investment = c(5000, 4500, 5000, 5000),
  expected_annual_return = c(450, 520, 400, 175),
  return_percentage = c(9.0, 11.6, 8.0, 3.5)
)

investment_options$option <- factor(investment_options$option,
                                     levels = investment_options$option)

ggplot(investment_options, aes(x = option, y = return_percentage, fill = option)) +
  geom_bar(stat = "identity") +
  labs(title = "Investment Alternatives: Expected Returns",
       subtitle = "Opportunity cost of choosing hotel = foregone return from best alternative",
       x = "",
       y = "Expected Annual Return (%)") +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "Set3") +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 12)) +
  geom_text(aes(label = paste0(return_percentage, "%\n(€", 
                                expected_annual_return, "k/yr)")), 
            vjust = -0.3, size = 3.5) +
  ylim(0, 14)
```

**If you build hotel in Lisbon**: Opportunity cost = foregone €520k from Algarve (best alternative) (We are assuming the same investment in both places)

</div>

## Sunk Costs vs. Opportunity Costs

**Sunk Cost**: Cost already incurred that **cannot be recovered**

**Critical Rule**: **Ignore sunk costs in decisions!**

They're already spent – only future costs and benefits matter.

---

**Tourism Example**:

You bought a non-refundable €100 train ticket to Porto. On the travel day, you feel sick.

**Sunk cost**: €100 (already spent, can't get back)

**Decision**: Should you go?

- ❌ Wrong thinking: "I paid €100, so I must go"
- ✅ Right thinking: "€100 is gone. Will I enjoy the trip enough given how I feel?"

**The €100 is irrelevant to the decision!**

## Real Example: Sunk Costs in Airlines

<div style="font-size: 80%;">

**Case**: TAP Air Portugal orders new aircraft

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Hypothetical airline decision timeline
timeline_data <- data.frame(
  stage = c("Order Placed\n(2022)", "Deposit Paid\n(2023)", "Delivery Due\n(2025)", "Decision Point\n(2024)"),
  cost_incurred = c(0, 150, 0, 0),
  cumulative_cost = c(0, 150, 150, 150),
  future_payment = c(1500, 1350, 1350, 1350)
)

timeline_data$stage <- factor(timeline_data$stage, levels = timeline_data$stage)
timeline_data$stage_num <- 1:4

ggplot(timeline_data, aes(x = stage_num, y = cumulative_cost)) +
  geom_line(size = 1.5, color = "#e74c3c") +
  geom_point(size = 4, color = "#e74c3c") +
  geom_line(aes(y = future_payment), size = 1.5, color = "#3498db", linetype = "dashed") +
  geom_point(aes(y = future_payment), size = 4, color = "#3498db") +
  labs(title = "Aircraft Purchase Decision: Sunk Cost Problem",
       subtitle = "At Decision Point: €150M is SUNK (ignore it). Focus on future €1,350M payment.",
       x = "",
       y = "Amount (€ millions)") +
  scale_x_continuous(breaks = 1:4, labels = timeline_data$stage) +
  theme_minimal(base_size = 14) +
  annotate("text", x = 2, y = 300, label = "Sunk Cost\n(€150M paid)", 
           color = "#e74c3c", size = 4, fontface = "bold") +
  annotate("text", x = 4, y = 1500, label = "Future Payment\n(€1,350M)", 
           color = "#3498db", size = 4, fontface = "bold") +
  geom_rect(aes(xmin = 3.7, xmax = 4.3, ymin = 0, ymax = 1600), 
            alpha = 0.01, fill = "yellow", color = "orange", size = 1.5)
```

**Correct Decision Process**: Compare future benefits (passenger revenue, efficiency) vs. future costs (€1,350M), **ignoring** the sunk €150M.

</div>

## Opportunity Cost in Time Allocation

<div style="font-size: 80%;">

**Very Important**: Your **time** has opportunity cost!

**Example**: Restaurant owner deciding whether to hire manager or self-manage

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Time allocation decision
time_decision <- data.frame(
  scenario = c("Self-Manage", "Hire Manager"),
  restaurant_profit = c(120000, 110000),
  manager_salary = c(0, -45000),
  owner_consulting = c(0, 65000),
  net_income = c(120000, 130000)
)

time_decision$scenario <- factor(time_decision$scenario, 
                                  levels = time_decision$scenario)

# Reshape for stacked bar
library(tidyr)
library(dplyr)
time_long <- time_decision %>%
  pivot_longer(cols = c(restaurant_profit, manager_salary, owner_consulting),
               names_to = "component", values_to = "value")

time_long$component <- factor(time_long$component,
                               levels = c("restaurant_profit", "owner_consulting", "manager_salary"),
                               labels = c("Restaurant Profit", "Owner Consulting Income", "Manager Salary"))

ggplot(time_long, aes(x = scenario, y = value, fill = component)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Restaurant Owner's Time Allocation Decision",
       subtitle = "Opportunity cost of self-managing = foregone consulting income (€65k)",
       x = "",
       y = "Annual Amount (€)",
       fill = "") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("Restaurant Profit" = "#27ae60", 
                                "Owner Consulting Income" = "#3498db",
                                "Manager Salary" = "#e74c3c")) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0, color = "black") +
  geom_text(data = time_decision, 
            aes(x = scenario, y = net_income + 5000, 
                label = paste0("Net: €", format(net_income, big.mark = ",")),
                fill = NULL),  # Add this line
            size = 5, fontface = "bold", inherit.aes = FALSE)  # Add inherit.aes = FALSE
```

**Analysis**: Hiring manager is better (€130k > €120k) because owner's time has opportunity cost!

</div>

## Production Possibilities Frontier (Preview)

<div style="font-size: 80%;">

**Visual representation of opportunity cost at societal level**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Simple PPF for tourism economy
# Two goods: Beach tourism vs. Cultural tourism (museum capacity)
library(ggplot2)

# 1. Create Linear PPF Data
# Equation: cultural = 100 - beach
beach <- seq(0, 100, by = 5)
cultural <- 100 - beach
ppf_data <- data.frame(beach, cultural)

# 2. Define Points EXACTLY on the line (x + y = 100)
points <- data.frame(
  beach = c(30, 70, 90),
  cultural = c(70, 30, 10), # Adjusted to sum to 100
  label = c("A", "B", "C")
)

ggplot(ppf_data, aes(x = beach, y = cultural)) +
  geom_line(linewidth = 1.5, color = "#3498db") +
  geom_ribbon(aes(ymin = 0, ymax = cultural), alpha = 0.2, fill = "#3498db") +
  geom_point(data = points, aes(x = beach, y = cultural), size = 5, color = "#e74c3c") +
  geom_text(data = points, aes(x = beach, y = cultural, label = label), 
            vjust = -1.5, size = 6, fontface = "bold") +
  labs(title = "Tourism Development: Linear PPF",
       subtitle = "Constant Opportunity Cost: 1 unit of beach = 1 unit of culture",
       x = "Beach Tourism Capacity (units)",
       y = "Cultural Tourism Capacity (units)") +
  theme_minimal(base_size = 14) +
  # Arrows showing the trade-off from A to B
  annotate("segment", x = 30, xend = 70, y = 70, yend = 70,
           arrow = arrow(length = unit(0.3, "cm")), size = 1, color = "#e67e22") +
  annotate("text", x = 50, y = 74, label = "+40 beach units", size = 4, color = "#e67e22") +
  annotate("segment", x = 70, xend = 70, y = 70, yend = 30,
           arrow = arrow(length = unit(0.3, "cm")), size = 1, color = "#e67e22") +
  annotate("text", x = 85, y = 50, label = "-40 cultural units\n(Opportunity Cost)", 
           size = 4, color = "#e67e22", fontface = "bold")
```

**Next lecture**: We'll explore PPF in detail!

</div>

## Trade-offs in Tourism: Real Data

<div style="font-size: 80%;">

**Wakanda 2024: Investment Trade-offs**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Government tourism investment allocation
allocation_data <- data.frame(
  category = c("Mass Tourism\nInfrastructure", "Cultural Heritage\nPreservation", "Sustainable Tourism\nInitiatives", "Digital\nInnovation"),
  budget_2024 = c(180, 95, 75, 50),
  budget_2025_proposed = c(160, 110, 95, 65)
)

allocation_data$category <- factor(allocation_data$category,
                                     levels = allocation_data$category)

library(tidyr)
allocation_long <- allocation_data %>%
  pivot_longer(cols = starts_with("budget"),
               names_to = "year",
               values_to = "amount") %>%
  mutate(year = ifelse(year == "budget_2024", "2024 (Actual)", "2025 (Proposed)"))

ggplot(allocation_long, aes(x = category, y = amount, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Tourism Investment Allocation: Wakanda Government",
       subtitle = "Trade-offs: More cultural preservation = Less mass tourism infrastructure",
       x = "",
       y = "Budget (millions)",
       fill = "") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("2024 (Actual)" = "#95a5a6", 
                                "2025 (Proposed)" = "#27ae60")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 11)) +
  geom_text(aes(label = paste0("", amount, "M")), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, size = 3.5)
```

**Opportunity cost of increasing cultural preservation by €15M** = €20M less for mass tourism infrastructure

</div>

## Making Better Decisions

<div style="font-size: 80%;">

**Checklist for Rational Economic Decisions**:

1. ✓ Identify **all alternatives**
2. ✓ Estimate **benefits** of each alternative
3. ✓ Estimate **costs** (including opportunity costs!)
4. ✓ **Ignore sunk costs** (they're irrelevant)
5. ✓ Consider **marginal** changes (not just totals)
6. ✓ Choose alternative with **highest net benefit**

**Remember**: Good decisions maximize net benefit (Benefits - Opportunity Costs)


</div>

## Common Decision Mistakes

<div style="font-size: 80%;">

**Mistake 1**: Ignoring opportunity costs

- Thinking only about money paid, not value of alternatives foregone

**Mistake 2**: Counting sunk costs

- "I've already invested so much, I must continue"

**Mistake 3**: Ignoring marginal analysis

- Looking at totals instead of additional benefits/costs

**Mistake 4**: Not comparing net benefits

- Choosing option with highest benefit (but also highest cost)

</div>

## Real Business Case: Ryanair Strategy

**Ryanair's Opportunity Cost Thinking (2024)**

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Ryanair route profitability
route_data <- data.frame(
  route = c("Lisbon-London", "Lisbon-Paris", "Lisbon-Milan", "Lisbon-Dublin", "Lisbon-Brussels"),
  revenue_per_flight = c(12500, 11200, 10800, 9500, 8900),
  cost_per_flight = c(7500, 7200, 7000, 7100, 7300),
  profit_per_flight = c(5000, 4000, 3800, 2400, 1600),
  frequency_per_week = c(14, 12, 10, 7, 5)
)

route_data$route <- factor(route_data$route,
                            levels = route_data$route[order(route_data$profit_per_flight, 
                                                             decreasing = TRUE)])

ggplot(route_data, aes(x = route, y = profit_per_flight, fill = route)) +
  geom_bar(stat = "identity") +
  labs(title = "Ryanair Route Profitability Analysis (Hypothetical)",
       subtitle = "Opportunity cost: Each flight to Brussels = foregone €3,400 from London route",
       x = "",
       y = "Profit per Flight (€)") +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "RdYlGn", direction = 1) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1)) +
  geom_text(aes(label = paste0("€", format(profit_per_flight, big.mark = ","))), 
            vjust = -0.5, size = 4) +
  geom_segment(aes(x = 1, xend = 5, y = 5000, yend = 5000), 
               linetype = "dashed", color = "#e74c3c", size = 1) +
  annotate("text", x = 3, y = 5300, 
           label = "Best alternative profit", color = "#e74c3c", size = 4)
```

**Decision**: Cancel Brussels route, add London frequency → Better use of aircraft (opportunity cost logic)

## Summary: Key Concepts

1. **Rationality**: People seek to maximize well-being through cost-benefit analysis
2. **Cost-Benefit Principle**: Do it if benefits ≥ costs
3. **Marginal Analysis**: Decisions at the margin (additional units)
4. **Opportunity Cost**: Value of best alternative foregone
5. **Sunk Costs**: Ignore them! (already spent, irrelevant)
6. **Trade-offs**: Choosing one thing means giving up another
7. **Good Decisions**: Maximize net benefit considering opportunity costs

## Exercises {.center}

**Practice Time!**

Real-world applications of opportunity cost.

## Exercise 1: Multiple Choice

**Question**: A hotel owner can use her restaurant space for:

- Option A: Fine dining (€80k annual profit)
- Option B: Buffet restaurant (€60k annual profit)  
- Option C: Rent to external operator (€50k annual rent)

She chooses Option A. What is the opportunity cost of this decision?

A. €80,000  
B. €60,000  
C. €50,000  
D. €110,000

::: {.fragment}
**Answer: B (€60,000)**

Explanation: Opportunity cost = value of **best alternative foregone**. The next-best alternative to fine dining (€80k) is buffet (€60k), NOT the sum of all alternatives.
:::

## Exercise 2: Multiple Choice

**Question**: You bought a non-refundable €200 ticket to a concert. On the day of the concert, you feel ill. A friend offers you €50 for the ticket (transferable). What is the opportunity cost of **attending** the concert?

A. €200  
B. €150  
C. €50 + discomfort from illness  
D. €250

::: {.fragment}
**Answer: C (€50 + discomfort)**

Explanation: 

- €200 is SUNK (can't recover even if you don't go)
- If you attend: you give up €50 (could sell) + endure discomfort
- Opportunity cost = what you give up by attending
:::

## Exercise 3: Open Question

**Scenario**: A tour operator in Porto must decide between two summer season strategies:

**Strategy A**: Focus on luxury tours

- Expected revenue: €500,000
- Variable costs: €280,000
- Fixed costs already committed: €80,000

**Strategy B**: Focus on budget tours

- Expected revenue: €420,000
- Variable costs: €200,000
- Same fixed costs: €80,000

**Questions**:

a) Calculate profit for each strategy  
b) What is the opportunity cost of choosing Strategy A?  
c) Which strategy should be chosen and why?  
d) Are fixed costs relevant to the decision? Explain.

## Exercise 3: Solution

**a) Profit calculation**:

**Strategy A (Luxury)**:
$$\text{Profit}_A = €500,000 - €280,000 - €80,000 = €140,000$$

**Strategy B (Budget)**:
$$\text{Profit}_B = €420,000 - €200,000 - €80,000 = €140,000$$

**Both strategies have equal profit!**

**b) Opportunity cost of Strategy A**:

Opportunity cost = Profit from best alternative foregone = **€140,000** (Strategy B profit)

## Exercise 3: Solution (continued)

**c) Which strategy to choose?**

**From pure profit perspective**: Indifferent (both €140,000)

**But consider other factors**:

- **Risk**: Luxury tours may have more variable demand
- **Reputation**: Does company have luxury experience?
- **Market trends**: Which segment is growing?
- **Resource requirements**: Which uses existing capabilities better?

**Recommendation**: Choose Strategy B (budget) because:

- Lower variable costs (€200k vs €280k) = less risk
- Higher revenue-to-variable-cost ratio: 2.1 vs. 1.79
- More resilient to demand fluctuations

**d) Are fixed costs relevant?**

**NO!** Fixed costs (€80,000) are **SUNK** – already committed regardless of strategy choice.

Only **variable costs** matter for this decision (they differ between strategies).

This is key example of ignoring sunk costs in decision-making!

## Next Lecture

**Lecture 4** (February 13, 2026):

- **Production Factors** (detailed)
- **Production Possibilities Frontier (PPF)**
- Efficiency, Trade-offs, and Economic Growth

**Preparation**: Review today's opportunity cost concept – it's the foundation for understanding PPF!

## Thank You! {.center}

**Questions?**

*Contact*: paulo.fagandini@ext.universidadeeuropeia.pt

---

*Next class: Thursday, February 13, 2026*

::: {.notes}
Opportunity cost is one of the most powerful concepts in economics. Make sure students understand it deeply – it will appear throughout the entire course!
:::
