---
title: "The Panorama of Macroeconomics"
subtitle: "Lecture 20: The Main Macroeconomic Issues"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    from: markdown+emoji
    theme: [default, custom.scss]
    transition: slide
    slide-number: true
    chalkboard: true
    scrollable: false
    footer: "Economics of Tourism | Lecture 20"
    logo: ../images/logo.png
    width: 1600
    height: 900
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(eurostat)
library(tidyverse)
library(lubridate)

ue_red    <- "#C8102E"
ue_black  <- "#1D1D1B"
ue_gray   <- "#6D6E71"
ue_white  <- "#FFFFFF"
ue_light  <- "#F5F5F5"
```

## Recap: Where We Have Been :rewind:

:::: {.columns}
::: {.column width="50%"}
**Microeconomics block** :microscope:

:one: Fundamentals, scarcity, opportunity cost

:two: Consumer theory — preferences, utility, demand

:three: Producer theory — costs, profit, supply

:four: Market equilibrium — price & quantity
:::
::: {.column width="50%"}
**Game Theory block** :chess_pawn:

:five: Strategic interaction — Nash equilibrium

:six: Sequential games — backward induction

. . .

:point_right: **Today we zoom out** :telescope:

We move from individual markets to the **economy as a whole**
:::
::::

# Part I: From Micro to Macro

## The Key Difference :mag:

:::: {.columns}
::: {.column width="48%"}
:microscope: **Microeconomics**

Studies **individual** agents and **specific** markets

- One firm's pricing decision
- One consumer's demand for hotel nights
- Equilibrium in the Lisbon restaurant market
:::
::: {.column width="4%"}
:::
::: {.column width="48%"}
:telescope: **Macroeconomics**

Studies the **economy as a whole**

- Total output of the entire Portuguese economy
- Overall unemployment rate
- The general level of prices (inflation)
:::
::::

::: {.fragment}
::: {.key-concept}
**Aggregation** is the tool that bridges the two. Macroeconomists **sum up** individual variables — millions of consumers, thousands of firms — into economy-wide totals.
:::
:::

## Why Aggregation? :jigsaw:

:::: {.columns}
::: {.column width="55%"}
Imagine trying to manage a flight by tracking every single **air molecule** in the cabin.

Instead, pilots watch **aggregate** gauges:

:thermometer: Cabin temperature

:wind_face: Air pressure

:fuel_pump: Total fuel remaining

. . .

Macroeconomics works the same way. We cannot track every transaction, so we build **aggregate indicators** that summarize the economy's health.
:::
::: {.column width="45%"}
| Individual variable | Aggregate |
|---|---|
| Your wage | Total wages in the economy |
| One firm's output | GDP |
| One hotel's prices | Overall inflation |
| One worker unemployed | Unemployment rate |
:::
::::

## The Macroeconomic Dashboard :bar_chart:

::: {.key-concept}
**Macroeconomics** is the study of the performance of the **national economy** and the policy measures used to improve it.
:::

. . .

**Six big questions macroeconomists ask:**

:seedling: Economic growth & living standards

:factory: Productivity

:roller_coaster: Recessions & expansions

:briefcase: Unemployment

:fire: Inflation

:globe_with_meridians: Economic interdependence of nations

# Part II: The Six Main Issues

## :seedling: 1. Economic Growth & Living Standards

:::: {.columns}
::: {.column width="55%"}
Over the past century, living standards in industrialised countries have improved **dramatically**.

- Life expectancy up
- Access to health, education, technology
- Real incomes higher

**Macroeconomics asks:**

- Why do some countries grow faster than others?
- What drives long-run improvements in well-being?
- Can tourism-dependent economies sustain growth?
:::
::: {.column width="45%"}
```{r}
#| fig-width: 5.5
#| fig-height: 4.5

# Real GDP per capita Portugal — Eurostat nama_10_pc
gdp_raw <- get_eurostat("nama_10_pc",
  filters = list(
    geo   = "PT",
    na_item = "B1GQ",
    unit  = "CP_EUR_HAB"
  ),
  time_format = "num"
)

gdp <- gdp_raw |>
  filter(time >= 2000) |>
  arrange(time)

ggplot(gdp, aes(x = time, y = values / 1000)) +
  geom_line(color = ue_red, linewidth = 1.3) +
  geom_area(fill = ue_red, alpha = 0.1) +
  annotate("rect", xmin = 2008.5, xmax = 2013.5,
           ymin = -Inf, ymax = Inf,
           fill = ue_gray, alpha = 0.15) +
  annotate("text", x = 2011, y = 12,
           label = "Financial\nCrisis", color = ue_gray,
           size = 3, fontface = "italic") +
  annotate("rect", xmin = 2019.8, xmax = 2021,
           ymin = -Inf, ymax = Inf,
           fill = ue_gray, alpha = 0.15) +
  annotate("text", x = 2020.4, y = 12,
           label = "COVID", color = ue_gray,
           size = 3, fontface = "italic") +
  labs(
    title = "Portugal: GDP per capita",
    x = NULL, y = "Thousand €",
    caption = "Source: Eurostat (nama_10_pc)"
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(color = ue_red, face = "bold"),
    panel.grid.minor = element_blank()
  )
```
:::
::::

## :factory: 2. Productivity

::: {.key-concept}
**Labour productivity** = Output per worker employed

It is the **single most important** determinant of living standards in the long run.
:::

. . .

:::: {.columns}
::: {.column width="50%"}
**Why does it matter for tourism?**

:beach_umbrella: A hotel with better-trained staff can serve **more guests** at **higher quality**

:airplane: A tour operator using better software books more itineraries per employee

:moneybag: Higher productivity → higher wages → higher living standards
:::
::: {.column width="50%"}
**What drives productivity?**

:books: Education & training

:computer: Technology & capital

:gear: Better management practices

:recycle: Infrastructure (transport, broadband)

:point_right: Countries that invest in these grow **faster and richer**
:::
::::

## :roller_coaster: 3. Recessions & Expansions

:::: {.columns}
::: {.column width="55%"}
Economies do **not** grow smoothly. They experience:

**Expansions** :chart_with_upwards_trend: — periods of rising output & employment

**Recessions** :chart_with_downwards_trend: — periods of falling output (two or more consecutive quarters of negative growth)

**Why does this matter for tourism?**

Tourism is highly **cyclical** — one of the first sectors to suffer in recessions and one of the first to bounce back in expansions.
:::
::: {.column width="45%"}
```{r}
#| fig-width: 5.5
#| fig-height: 4.5

# GDP growth rate Portugal — Eurostat nama_10_gdp
growth_raw <- get_eurostat("namq_10_gdp",
  filters = list(
    geo    = "PT",
    na_item = "B1GQ",
    unit   = "CLV_PCH_PRE",
    s_adj  = "SCA"
  ),
  time_format = "date"
)

growth <- growth_raw |>
  filter(time >= as.Date("2005-01-01")) |>
  arrange(time)

ggplot(growth, aes(x = time, y = values)) +
  geom_col(aes(fill = values >= 0), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = ue_red, "FALSE" = ue_black)) +
  geom_hline(yintercept = 0, color = ue_gray, linewidth = 0.6) +
  labs(
    title = "Portugal: GDP Growth Rate (quarterly)",
    x = NULL, y = "% change (prev. quarter)",
    caption = "Source: Eurostat (namq_10_gdp)"
  ) +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(color = ue_red, face = "bold"),
    panel.grid.minor = element_blank()
  )
```
:::
::::

## :briefcase: 4. Unemployment

::: {.key-concept}
**Unemployment rate** = % of people who are **willing and able** to work but **cannot find a job**
:::

. . .

:::: {.columns}
::: {.column width="55%"}
**Key facts:**

- Unemployment rises sharply in recessions
- Even in booms, some unemployment always exists (people switching jobs, new graduates entering the market)
- Youth unemployment tends to be much higher than overall unemployment
- Tourism creates a lot of employment — but often **seasonal and precarious**
:::
::: {.column width="45%"}
```{r}
#| fig-width: 5.5
#| fig-height: 4.5

unemp_raw <- get_eurostat("une_rt_a",
  filters = list(
    geo  = c("PT", "EU27_2020"),
    age  = "Y15-74",
    sex  = "T",
    unit = "PC_ACT"
  ),
  time_format = "num"
)

unemp <- unemp_raw |>
  filter(time >= 2000) |>
  mutate(geo_label = case_when(
    geo == "PT" ~ "Portugal",
    TRUE        ~ "EU-27"
  ))

ggplot(unemp, aes(x = time, y = values, color = geo_label)) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = c("Portugal" = ue_red, "EU-27" = ue_black)) +
  labs(
    title = "Unemployment Rate",
    x = NULL, y = "%", color = NULL,
    caption = "Source: Eurostat (une_rt_a)"
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(color = ue_red, face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```
:::
::::

## :fire: 5. Inflation

::: {.key-concept}
**Inflation** = the rate at which the **general level of prices** rises over time. If inflation is 3%, a basket of goods that cost €100 today will cost **€103** next year.
:::

. . .

:::: {.columns}
::: {.column width="50%"}
**Why does it matter for tourism?**

:hotel: Higher hotel prices reduce the purchasing power of tourists

:airplane: Flight costs affect travel decisions

:currency_exchange: Inflation differences between countries affect **exchange rates** and **competitiveness**

:point_right: Portugal's inflation relative to Germany or the UK affects how attractive it is as a destination
:::
::: {.column width="50%"}
```{r}
#| fig-width: 5.5
#| fig-height: 4.5

hicp_raw <- get_eurostat("prc_hicp_aind",
  filters = list(
    geo   = "PT",
    coicop = "CP00",
    unit  = "RCH_A_AVG"
  ),
  time_format = "num"
)

hicp <- hicp_raw |>
  filter(time >= 2000) |>
  arrange(time)

ggplot(hicp, aes(x = time, y = values)) +
  geom_col(aes(fill = values >= 0), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = ue_red, "FALSE" = ue_black)) +
  geom_hline(yintercept = 0, color = ue_gray, linewidth = 0.5) +
  labs(
    title = "Portugal: HICP Inflation (annual)",
    x = NULL, y = "%",
    caption = "Source: Eurostat (prc_hicp_aind)"
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(color = ue_red, face = "bold"),
    panel.grid.minor = element_blank()
  )
```
:::
::::

## :globe_with_meridians: 6. Economic Interdependence of Nations

:::: {.columns}
::: {.column width="55%"}
Modern economies are **deeply interconnected**:

:ship: Trade in goods and services

:bank: International capital flows

:currency_exchange: Exchange rate fluctuations

:handshake: Trade agreements (EU single market, WTO)

**For tourism, this is especially visible:**

Portugal receives visitors from Germany, the UK, the US, Brazil — each with their own economic cycles, exchange rates, and purchasing power.

:point_right: A recession in Germany directly **reduces tourism demand** in the Algarve.
:::
::: {.column width="45%"}
```{r}
#| fig-width: 5.5
#| fig-height: 4.5

# Tourism receipts from Eurostat (tour_dem_extot or bop_its6_det)
# Use trade openness as a robust proxy: exports + imports / GDP
trade_raw <- get_eurostat("nama_10_gdp",
  filters = list(
    geo     = "PT",
    na_item = c("P6", "P7", "B1GQ"),
    unit    = "CP_MEUR"
  ),
  time_format = "num"
)

trade <- trade_raw |>
  filter(time >= 2000) |>
  select(time, na_item, values) |>
  pivot_wider(names_from = na_item, values_from = values) |>
  mutate(openness = (P6 + P7) / B1GQ * 100) |>
  filter(!is.na(openness))

ggplot(trade, aes(x = time, y = openness)) +
  geom_line(color = ue_red, linewidth = 1.3) +
  geom_area(fill = ue_red, alpha = 0.1) +
  labs(
    title = "Portugal: Trade Openness",
    subtitle = "(Exports + Imports) / GDP × 100",
    x = NULL, y = "%",
    caption = "Source: Eurostat (nama_10_gdp)"
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(color = ue_red, face = "bold"),
    plot.subtitle = element_text(color = ue_gray, size = 10),
    panel.grid.minor = element_blank()
  )
```
:::
::::

# Part III: The Macro Toolkit

## How Do Macroeconomists Think? :wrench:

Macroeconomics does not just describe problems — it tries to **explain and fix** them.

. . .

:::: {.columns}
::: {.column width="50%"}
**Two types of analysis:**

:straight_ruler: **Positive analysis**

Objective study of cause and effect.
"If interest rates rise, what happens to investment?"

:balance_scale: **Normative analysis**

Value judgments about what *should* be done.
"Should the government reduce unemployment even if it raises inflation?"
:::
::: {.column width="50%"}
**Two main policy levers:**

:bank: **Monetary policy**

Central banks control money supply and interest rates (e.g., the ECB for Portugal)

:receipt: **Fiscal policy**

Government controls taxes and public spending

:point_right: We will study both in detail in the coming lectures.
:::
::::

## The Macroeconomic Policy Triangle :triangular_flag_on_post:

:::: {.columns}
::: {.column width="55%"}
Policymakers juggle three goals that are often **in tension**:

:chart_with_upwards_trend: **Growth** — increase output and living standards

:briefcase: **Low unemployment** — keep people employed

:fire: **Low inflation** — keep prices stable

. . .

**The tension:**

Policies that boost growth and cut unemployment often *raise* inflation.

Policies that reduce inflation often *slow* growth and *raise* unemployment.

:point_right: This is the classic **Phillips Curve** trade-off (we will revisit this).
:::
::: {.column width="45%"}
```{r}
#| fig-width: 5
#| fig-height: 4.5

# Scatter: unemployment vs inflation Portugal
u_raw <- get_eurostat("une_rt_a",
  filters = list(geo = "PT", age = "Y15-74", sex = "T",
                 unit = "PC_ACT"),
  time_format = "num"
) |> filter(time >= 2000) |> select(time, unemp = values)

p_raw <- get_eurostat("prc_hicp_aind",
  filters = list(geo = "PT", coicop = "CP00", unit = "RCH_A_AVG"),
  time_format = "num"
) |> filter(time >= 2000) |> select(time, infl = values)

scatter <- inner_join(u_raw, p_raw, by = "time") |>
  filter(!is.na(unemp), !is.na(infl))

ggplot(scatter, aes(x = unemp, y = infl)) +
  geom_point(color = ue_red, size = 3, alpha = 0.8) +
  geom_text(aes(label = time), nudge_y = 0.3, size = 2.8,
            color = ue_gray) +
  geom_smooth(method = "lm", se = FALSE,
              color = ue_black, linewidth = 0.8, linetype = "dashed") +
  labs(
    title = "Unemployment vs Inflation — Portugal",
    x = "Unemployment rate (%)",
    y = "Inflation rate (%)",
    caption = "Source: Eurostat"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(color = ue_red, face = "bold"),
    panel.grid.minor = element_blank()
  )
```
:::
::::

## Tourism and Macroeconomics :beach_umbrella:

Macroeconomic variables directly shape tourism:

. . .

:::: {.columns}
::: {.column width="33%"}
:chart_with_upwards_trend: **GDP Growth**

When incomes rise, households spend **more** on tourism — especially international trips.

Tourism is a **normal good** with income elasticity > 1 (luxury good behaviour).
:::
::: {.column width="33%"}
:briefcase: **Unemployment**

High unemployment → lower household incomes → **less discretionary spending** on travel.

Tourism jobs are often the **first cut** in downturns.
:::
::: {.column width="33%"}
:fire: **Inflation**

Rising prices in destination countries reduce **price competitiveness**.

Exchange rate shifts alter the **real cost** of foreign travel.
:::
::::

. . .

:point_right: The 2008 financial crisis, COVID-19 in 2020, and the 2022 inflation spike all hit tourism **hard and fast** — illustrating exactly why macro matters for this industry.

## The Road Ahead :map:

Over the next lectures, we will build the **macro toolkit** piece by piece:

| Lecture | Topic |
|---|---|
| 21 | Aggregation: the concept |
| 22 | GDP: measuring total production |
| 23 | Inflation |
| 24 | Central Banks & Monetary Policy |
| 25 | Fiscal Policy |
| 26 | Balance of Payments |

:calendar: **Test 2** covers producer theory and macroeconomics — May 28th.

# Exercises

## :pencil: Exercise 1 — Multiple Choice

**Which of the following best describes macroeconomics?**

. . .

**(A)** The study of how an individual firm decides how much to produce

**(B)** The study of the performance of the national economy and the policies used to improve it

**(C)** The analysis of how a single consumer chooses between two goods given a budget constraint

**(D)** The study of price formation in a single competitive market

::: {.fragment}
::: {.highlight-box}
**Correct answer: (B)**

Macroeconomics studies economy-wide outcomes — GDP, unemployment, inflation — and the policies that affect them. Options A, C, and D are microeconomic questions.
:::
:::

## :pencil: Exercise 2 — Multiple Choice

**Portugal's unemployment rate rose sharply after 2008 and again in 2020. Which macroeconomic concept best explains why tourism revenues typically fall during such periods?**

. . .

**(A)** Tourism is an inferior good, so demand rises when incomes fall

**(B)** Inflation is always lower during recessions, making tourism cheaper

**(C)** Tourism has high income elasticity; falling incomes reduce discretionary travel spending

**(D)** Monetary policy automatically increases tourism demand during downturns

::: {.fragment}
::: {.highlight-box}
**Correct answer: (C)**

Tourism behaves as a luxury/normal good with income elasticity > 1. When unemployment rises and household incomes fall, discretionary spending — like holidays — is cut first. Option A is incorrect (it is a normal good), B is unrelated, and D is wrong.
:::
:::

## :pencil: Exercise 3 — Open Question

**Consider the following macroeconomic context for Portugal:**

A severe recession hits the Eurozone. GDP contracts by 3%, unemployment rises to 15%, and inflation falls to near zero. The British pound also depreciates 15% against the euro.

**(a)** Using the macroeconomic issues covered in this lecture, identify **three** distinct channels through which this scenario affects Portugal's tourism sector. *(6 marks)*

**(b)** For each channel, indicate whether the effect on tourism demand is **positive or negative**, and briefly justify. *(6 marks)*

**(c)** The recession in the Eurozone is accompanied by a **strong growth period in Brazil**. How might this partially offset the effects identified in (a)? *(3 marks)*

---

::: {.fragment}

**Solution outline:**

**(a)** Channels: (1) GDP/income of EU tourists falls → less demand from Germany, France, Spain; (2) Unemployment rises → fewer households can afford international travel; (3) GBP depreciation → UK tourists face higher real price for Portuguese holidays.

**(b)** (1) Negative — lower incomes reduce outbound tourism; (2) Negative — unemployed households cut discretionary travel first; (3) Negative — Portugal becomes 15% more expensive for British visitors.

**(c)** Brazilian GDP growth → Brazilian tourists have higher incomes and more demand for European travel. Brazil is a key source market for Portugal (language, cultural ties). Positive demand shock from non-EU markets can partially offset EU-origin declines.

:::

## Summary :books:

**Today we covered:**

:white_check_mark: The micro-macro distinction — individual markets vs economy-wide aggregates

:white_check_mark: Aggregation as the key tool of macroeconomics

:white_check_mark: The **six main macroeconomic issues**: growth, productivity, recessions/expansions, unemployment, inflation, and international interdependence

:white_check_mark: The two policy levers: monetary and fiscal policy

:white_check_mark: Why all of this matters directly for **tourism**

. . .

**Next lecture (Lecture 21):**

:jigsaw: Aggregation — the concept in depth; building up from individual transactions to economy-wide totals

## Thank You! :wave: {.center}

**Questions?**

:email: paulo.fagandini@ext.universidadeeuropeia.pt

---
*Next class: Wednesday, May 7th, 2026*
