---
title: "Producer Theory"
subtitle: "Lecture 15: Elasticity — Demand Recap & Supply Elasticity"
author: "Paulo Fagandini"
date: today
date-format: "YYYY"
format:
  revealjs:
    from: markdown+emoji
    theme: [default, custom.scss]
    transition: slide
    slide-number: true
    chalkboard: false        # ⚠️ must disable — incompatible with embed-resources
    scrollable: false
    footer: "Economics of Tourism | Lecture 15"
    logo: ../images/logo.png
    width: 1600
    height: 900
    embed-resources: true
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyverse)
library(patchwork)

ue_red <- "#C8102E"
ue_black <- "#1D1D1B"
ue_gray <- "#6D6E71"
ue_white <- "#FFFFFF"
ue_blue <- "#2E86AB"
ue_green <- "#27ae60"
ue_orange <- "#E67E22"
```

## Recap: Lecture 14 :rewind:

<div style="font-size: 80%;">

**What we covered last time**:

- **Producer surplus**: $PS = TR - VC$ = area above supply, below price
- **Market supply**: horizontal sum of individual supply curves (kinks when firms enter)
- **Linear supply**: $P = c + dQ$, with $PS = \frac{1}{2} \times Q^* \times (P^* - c)$

. . .

:dart: **Today**: How **responsive** are quantities to price changes?

Before we tackle **supply** elasticity, we need a quick recap of the **demand** elasticity concept — we ran out of time for it earlier in the course. The good news: the idea is the same on both sides!

</div>

# Demand Elasticity: The Key Ideas

## Price Elasticity of Demand :chart_with_downwards_trend: {.smaller}

<div style="font-size: 76%;">

::: {.key-concept}
**Price Elasticity of Demand ($\varepsilon_d$)**

The **percentage change in quantity demanded** resulting from a **1% change in price**:

$$\varepsilon_d = \frac{\%\Delta Q_d}{\%\Delta P} = \frac{\Delta Q / Q}{\Delta P / P}$$
:::

Since demand slopes downward (P↑ → Q↓), $\varepsilon_d$ is **negative**. We often use the **absolute value** $|\varepsilon_d|$.

| $|\varepsilon_d|$ | Classification | Meaning | Tourism example |
|:---:|---|---|---|
| $> 1$ | **Elastic** | Q responds **more** than P | Leisure beach holidays |
| $= 1$ | **Unit elastic** | Q responds **exactly** as P | — (theoretical midpoint) |
| $< 1$ | **Inelastic** | Q responds **less** than P | Business travel |
| $= 0$ | **Perfectly inelastic** | Q doesn't respond at all | Emergency medical flights |
| $= \infty$ | **Perfectly elastic** | Infinite response | Identical commodity, many sellers |

</div>

## Calculating Demand Elasticity :pencil2:

<div style="font-size: 74%;">

**Two methods** (same as we'll use for supply):

:::: {.columns}
::: {.column width="50%"}
:one: **Point elasticity** (at one point):

$$\varepsilon_d = \frac{\Delta Q}{\Delta P} \times \frac{P}{Q} = \frac{1}{\text{slope}} \times \frac{P}{Q}$$

**Example**: Demand is $P = 50 - 2Q$.

At $P = 30$, $Q = 10$: slope = $-2$

$$\varepsilon_d = \frac{1}{-2} \times \frac{30}{10} = -1.5$$

$|\varepsilon_d| = 1.5 > 1$ → **elastic**
:::

::: {.column width="50%"}
:two: **Arc elasticity** (midpoint method):

$$\varepsilon_d = \frac{Q_2 - Q_1}{(Q_1 + Q_2)/2} \div \frac{P_2 - P_1}{(P_1 + P_2)/2}$$

**Example**: A museum raises price from €10 to €12. Visitors fall from 1,000 to 800.

$$\varepsilon_d = \frac{-200}{900} \div \frac{2}{11} = -1.22$$

$|\varepsilon_d| = 1.22 > 1$ → **elastic**
:::
::::

</div>

:bulb: Key insight for linear demand: elasticity **varies** along the curve — elastic at the top (high P, low Q), inelastic at the bottom (low P, high Q), unit elastic at the midpoint.



## The Revenue Test & Determinants :moneybag:

<div style="font-size: 72%;">

:::: {.columns}
::: {.column width="50%"}
**The Revenue Test**:

How does total revenue ($TR = P \times Q$) respond to a price increase?

| Elasticity | P ↑ → TR? | Intuition |
|---|---|---|
| Elastic ($|\varepsilon_d| > 1$) | TR **falls** :arrow_down: | Q drops more than P rises |
| Unit elastic ($|\varepsilon_d| = 1$) | TR **unchanged** | Effects cancel |
| Inelastic ($|\varepsilon_d| < 1$) | TR **rises** :arrow_up: | Q drops less than P rises |

* :point_right: Hotel prices up 10%, but bookings only drop 3%? Demand is inelastic → revenue **rises**!
:::

::: {.column width="50%"}
**Five determinants of demand elasticity**:

1. :arrows_counterclockwise: **Substitutes**: more substitutes → more elastic
2. :moneybag: **Budget share**: bigger expense → more elastic
3. :gem: **Necessity vs luxury**: luxuries more elastic
4. :hourglass: **Time horizon**: long run → more elastic
5. :world_map: **Market definition**: narrower → more elastic

**Tourism**: Leisure travel is very elastic ($|\varepsilon_d| \approx 2$–3), business travel is inelastic ($|\varepsilon_d| \approx 0.3$). Airlines use this to price-discriminate!
:::
::::

</div>

## Demand Elasticity: Quick Practice :dart:

<div style="font-size: 78%;">

**Quick check — try these mentally before revealing the answers!**

**Q1**: Algarve resort demand: $P = 500 - 0.05Q$. At $P = 100$, $Q = 8{,}000$. Is demand elastic or inelastic?

::: {.fragment}
$\varepsilon_d = \frac{1}{-0.05} \times \frac{100}{8000} = -20 \times 0.0125 = -0.25$ → **Inelastic** ($|\varepsilon_d| = 0.25 < 1$). Raising prices would **increase** revenue!
:::

**Q2**: If luxury cruise demand has $|\varepsilon_d| = 2.5$ and operators raise prices by 8%, what happens to revenue?

::: {.fragment}
Elastic demand → price and revenue move in **opposite directions**. Revenue **falls**. ($\%\Delta Q \approx -20\%$, much larger than the 8% price rise.)
:::

. . .

:point_right: Now let's apply the **exact same logic** to the supply side!

</div>

# Price Elasticity of Supply

## Definition :straight_ruler:

<div style="font-size: 78%;">

::: {.key-concept}
**Price Elasticity of Supply ($\varepsilon_s$)**

The **percentage change in quantity supplied** resulting from a **1% change in price**:

$$\varepsilon_s = \frac{\%\Delta Q_s}{\%\Delta P} = \frac{\Delta Q / Q}{\Delta P / P}$$
:::

**Key differences from demand elasticity**:

| | Demand elasticity ($\varepsilon_d$) | Supply elasticity ($\varepsilon_s$) |
|---|---|---|
| **Sign** | Negative (P↑ → Q↓) | **Positive** (P↑ → Q↑) |
| **We use absolute value?** | Yes ($|\varepsilon_d|$) | No — it's already positive |
| **Interpretation** | Responsiveness of buyers | Responsiveness of sellers |

:bulb: The formula is **identical** — the only difference is that supply slopes upward, so $\varepsilon_s > 0$.

</div>

## Classification :bar_chart: {.smaller}

<div style="font-size: 78%;">

| Elasticity value | Classification | Meaning |
|:---:|---|---|
| $\varepsilon_s > 1$ | **Elastic** | Quantity supplied responds **more** than proportionally to price |
| $\varepsilon_s = 1$ | **Unit elastic** | Quantity supplied responds **exactly** proportionally |
| $\varepsilon_s < 1$ | **Inelastic** | Quantity supplied responds **less** than proportionally |
| $\varepsilon_s = 0$ | **Perfectly inelastic** | Quantity doesn't respond at all (vertical supply) |
| $\varepsilon_s = \infty$ | **Perfectly elastic** | Any quantity at one fixed price (horizontal supply) |

. . .

**Tourism intuition**:

- :hotel: Hotel rooms in Lisbon in the short run → **inelastic** (can't build new rooms overnight)
- :fork_and_knife: Street food vendors in a tourist area → **elastic** (easy to set up a new stall)
- :classical_building: UNESCO heritage sites → **perfectly inelastic** (supply is fixed!)

</div>

# Calculating Supply Elasticity

## Point and Arc Methods :pencil2:

<div style="font-size: 72%;">

:::: {.columns}
::: {.column width="50%"}
:one: **Point elasticity**:

$$\varepsilon_s = \frac{1}{\text{slope}} \times \frac{P}{Q}$$

**Example**: Supply is $P = 5 + 0.25Q$. At $P = 30$, $Q = 100$:

Slope = $0.25$, so $\frac{1}{\text{slope}} = 4$

$$\varepsilon_s = 4 \times \frac{30}{100} = 1.2$$

**Elastic** — a 1% price increase leads to a 1.2% increase in quantity supplied.
:::

::: {.column width="50%"}
:two: **Arc elasticity** (midpoint):

$$\varepsilon_s = \frac{Q_2 - Q_1}{(Q_1 + Q_2)/2} \div \frac{P_2 - P_1}{(P_1 + P_2)/2}$$

**Example**: Tour operator increases tours from 40 to 50/month when price rises from €80 to €100.

$$\varepsilon_s = \frac{10}{45} \div \frac{20}{90} = 0.222 \div 0.222 = 1.0$$

**Unit elastic** — quantity and price change by the same percentage.
:::
::::

:point_right: Same methods as demand elasticity — just applied to the supply curve!

</div>

## Elasticity Along a Linear Supply Curve :straight_ruler:

<div style="font-size: 72%;">

```{r}
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 5
#| fig-align: center

df_origin <- data.frame(Q = seq(0, 18, by = 0.5)) %>%
  mutate(P = Q / 3)

df_noorigin <- data.frame(Q = seq(0.5, 18, by = 0.5)) %>%
  mutate(P = 2 + Q)

pts_no <- data.frame(
  Q = c(2, 6, 14),
  P = c(4, 8, 16)
) %>%
  mutate(eps = P / Q,
         label = paste0("ε = ", round(eps, 2)))

p1 <- ggplot() +
  geom_line(data = df_origin, aes(x = Q, y = P),
            color = ue_red, linewidth = 1.5) +
  annotate("text", x = 16, y = 4, label = "S (through origin)",
           color = ue_red, fontface = "bold", size = 4) +
  annotate("text", x = 9, y = 1.5,
           label = "ε = 1 everywhere!", color = ue_red,
           fontface = "bold.italic", size = 4.5) +
  coord_cartesian(ylim = c(0, 7), xlim = c(0, 20)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Q", y = "P",
       title = "Supply Through the Origin: ε = 1 Always") +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(color = ue_red, face = "bold", hjust = 0.5, size = 13),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p2 <- ggplot() +
  geom_line(data = df_noorigin, aes(x = Q, y = P),
            color = ue_blue, linewidth = 1.5) +
  geom_point(data = pts_no, aes(x = Q, y = P), size = 3.5, color = ue_black) +
  geom_text(data = pts_no, aes(x = Q + 1.5, y = P - 0.8, label = label),
            fontface = "bold", size = 3.5, color = ue_black, hjust = 0) +
  annotate("text", x = 15, y = 12, label = "S (positive intercept)",
           color = ue_blue, fontface = "bold", size = 4) +
  annotate("text", x = 1.5, y = 2, label = "c = 2",
           color = ue_gray, fontface = "italic", size = 3.5) +
  coord_cartesian(ylim = c(0, 20), xlim = c(0, 20)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Q", y = "P",
       title = "Supply with Positive Intercept: ε Decreases") +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(color = ue_blue, face = "bold", hjust = 0.5, size = 13),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p1 + p2
```

**Left**: Supply through the origin → $\varepsilon_s = 1$ at **every** point (P and Q always change proportionally).

**Right**: Supply with positive intercept ($c > 0$) → $\varepsilon_s > 1$ near the intercept, **falling** as Q increases.

</div>

## The Origin Rule & Quick Guide :bulb:

<div style="font-size: 78%;">

**Why does supply through the origin always have $\varepsilon_s = 1$?**

If $P = dQ$: $\quad \varepsilon_s = \frac{1}{d} \times \frac{P}{Q} = \frac{1}{d} \times \frac{dQ}{Q} = 1$

The $P/Q$ ratio always equals $d$, which cancels perfectly with $1/d$.

. . .

::: {.key-concept}
**Quick Elasticity Guide for Linear Supply $P = c + dQ$**

| Intercept | Elasticity | Shortcut formula |
|---|---|---|
| $c = 0$ (through origin) | $\varepsilon_s = 1$ everywhere | — |
| $c > 0$ (positive intercept) | $\varepsilon_s > 1$, falling toward 1 | $\varepsilon_s = \frac{P}{P - c}$ |
| $c < 0$ (hits Q-axis) | $\varepsilon_s < 1$ everywhere | $\varepsilon_s = \frac{P}{P - c}$ (where $c < 0$) |
:::

:point_right: **Contrast with demand**: demand elasticity varies from $\infty$ to $0$ along a linear curve. Supply elasticity is more "stable."

</div>

## Extreme Cases :triangular_ruler:

<div style="font-size: 72%;">

```{r}
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 4.5
#| fig-align: center

p_inelastic <- ggplot() +
  geom_vline(xintercept = 50, color = ue_red, linewidth = 2) +
  annotate("text", x = 52, y = 7, label = "S", color = ue_red,
           fontface = "bold.italic", size = 6) +
  annotate("text", x = 30, y = 5, label = "ε = 0\nPerfectly\nInelastic",
           color = ue_red, fontface = "bold", size = 4.5) +
  coord_cartesian(ylim = c(0, 10), xlim = c(0, 100)) +
  scale_x_continuous(expand = c(0, 0), breaks = 50, labels = expression(bar(Q))) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Q", y = "P", title = "Perfectly Inelastic Supply") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(color = ue_red, face = "bold", hjust = 0.5, size = 13),
        axis.title = element_text(face = "bold"), panel.grid.minor = element_blank())

p_elastic <- ggplot() +
  geom_hline(yintercept = 5, color = ue_blue, linewidth = 2) +
  annotate("text", x = 80, y = 6.5, label = "S", color = ue_blue,
           fontface = "bold.italic", size = 6) +
  annotate("text", x = 50, y = 8, label = "ε = ∞\nPerfectly\nElastic",
           color = ue_blue, fontface = "bold", size = 4.5) +
  coord_cartesian(ylim = c(0, 10), xlim = c(0, 100)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), breaks = 5, labels = expression(bar(P))) +
  labs(x = "Q", y = "P", title = "Perfectly Elastic Supply") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(color = ue_blue, face = "bold", hjust = 0.5, size = 13),
        axis.title = element_text(face = "bold"), panel.grid.minor = element_blank())

p_inelastic + p_elastic
```

:::: {.columns}
::: {.column width="50%"}
**Perfectly inelastic** ($\varepsilon_s = 0$): Quantity fixed regardless of price.

*Tourism*: Hotel rooms in a fully built-up area, UNESCO sites, beach space.
:::

::: {.column width="50%"}
**Perfectly elastic** ($\varepsilon_s = \infty$): Any quantity at one price — constant MC.

*Tourism*: Digital products (e-tickets, online guides) — near-zero MC per copy.
:::
::::

</div>

# Determinants of Supply Elasticity

## What Makes Supply More or Less Elastic? :thinking:

<div style="font-size: 72%;">

The textbook identifies **four key determinants** (plus time):

:::: {.columns}
::: {.column width="50%"}
:one: **Flexibility of production factors**

Can inputs be used for other things? If yes → elastic.

- :white_check_mark: **Elastic**: Unskilled labor (waiters can switch from retail to tourism)
- :x: **Inelastic**: Specialized skills (brain surgeons, master sommeliers)

:two: **Mobility of production factors**

Can inputs move between locations? If yes → elastic.

- :white_check_mark: **Elastic**: Touring musicians, portable equipment
- :x: **Inelastic**: Beachfront land, historical buildings
:::

::: {.column width="50%"}
:three: **Ability to produce substitute inputs**

Can inputs be replicated over time? If yes → elastic.

- :white_check_mark: **Elastic**: Train more tour guides, build more hotel rooms
- :x: **Inelastic**: Rare natural resources (diamond mines, volcanic hot springs)

:four: **Time horizon** :hourglass:

- **Short run**: At least one input is fixed → more **inelastic**
- **Long run**: All inputs adjustable → more **elastic**
- **Very short run** (market period): supply is nearly **vertical**
:::
::::

</div>

## Time and Supply Elasticity :hourglass:

<div style="font-size: 70%;">

```{r}
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 5
#| fig-align: center

df_vsr <- data.frame(Q = seq(0, 120, by = 1)) %>% mutate(P = 5 + 0.8 * Q)
df_sr <- data.frame(Q = seq(0, 120, by = 1)) %>% mutate(P = 5 + 0.25 * Q)
df_lr <- data.frame(Q = seq(0, 120, by = 1)) %>% mutate(P = 5 + 0.08 * Q)

ggplot() +
  geom_line(data = df_vsr, aes(x = Q, y = P), color = ue_red,
            linewidth = 1.3, linetype = "dashed") +
  geom_line(data = df_sr, aes(x = Q, y = P), color = ue_orange, linewidth = 1.3) +
  geom_line(data = df_lr, aes(x = Q, y = P), color = ue_green,
            linewidth = 1.3, linetype = "longdash") +
  annotate("text", x = 35, y = 38, label = "Very Short Run\n(nearly fixed)",
           color = ue_red, fontface = "bold", size = 3.8) +
  annotate("text", x = 85, y = 28, label = "Short Run",
           color = ue_orange, fontface = "bold", size = 3.8) +
  annotate("text", x = 105, y = 11, label = "Long Run\n(all inputs adjustable)",
           color = ue_green, fontface = "bold", size = 3.8) +
  coord_cartesian(ylim = c(0, 45), xlim = c(0, 120)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Quantity (Q)", y = "Price (€)",
       title = "Supply Elasticity Increases with Time (Hypothetical)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(color = ue_red, face = "bold", hjust = 0.5, size = 15),
        axis.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

The **longer the time horizon**, the more firms can adjust all inputs → supply becomes **flatter** (more elastic).

</div>

# Tourism Applications

## Elasticity in Tourism: Both Sides :airplane:

<div style="font-size: 72%;">

:::: {.columns}
::: {.column width="50%"}
**Inelastic supply** ($\varepsilon_s$ small):

- :hotel: **Hotel rooms in central Lisbon**: limited space, takes years to build
- :classical_building: **Museum capacity**: Jerónimos can't add rooms
- :beach_umbrella: **Beach access in Algarve**: fixed coastline

**Implication**: When demand rises (peak season), price rises **sharply** because supply can't respond much.
:::

::: {.column width="50%"}
**Elastic supply** ($\varepsilon_s$ large):

- :bus: **Tour buses**: can add more buses relatively quickly
- :fork_and_knife: **Food trucks at festivals**: quick to set up
- :iphone: **Online booking services**: near-zero cost to scale

**Implication**: When demand rises, quantity adjusts more than price — price stays relatively stable.
:::
::::

. . .

* :point_right: **Both elasticities together** determine **who bears the burden** of taxes and price shocks! The side with **less elastic** response absorbs more of the burden.

</div>

## The Full Parallel: Demand vs Supply Elasticity :handshake:

<div style="font-size: 70%;">

| Concept | Demand Elasticity | Supply Elasticity |
|---|---|---|
| **Formula** | $\varepsilon_d = \frac{\Delta Q/Q}{\Delta P/P}$ | $\varepsilon_s = \frac{\Delta Q/Q}{\Delta P/P}$ |
| **Sign** | Negative (use $|\varepsilon_d|$) | **Positive** |
| **Elastic** | $|\varepsilon_d| > 1$ | $\varepsilon_s > 1$ |
| **Revenue test** | Elastic: P↑ → TR↓ | No direct revenue test |
| **Along linear curve** | Varies: ∞ → 1 → 0 | More stable; through origin = 1 |
| **Key determinants** | Substitutes, budget share, necessity, time, market definition | **Input flexibility, mobility, replicability, time** |
| **Perfectly inelastic** | Vertical demand | Vertical supply |
| **Perfectly elastic** | Horizontal demand | Horizontal supply |

* :point_right: Both sides use the **same formula** and **same classification**. The determinants differ because they reflect different underlying decisions (buying vs producing).

</div>

## Summary :memo:

<div style="font-size: 70%;">

**Today's Key Takeaways**:

**Demand elasticity recap**:

1. $\varepsilon_d = \frac{\%\Delta Q}{\%\Delta P}$ — negative, use $|\varepsilon_d|$. Elastic > 1, inelastic < 1
2. **Revenue test**: elastic → P↑ means TR↓; inelastic → P↑ means TR↑
3. **Determinants**: substitutes, budget share, necessity/luxury, time, market definition

**Supply elasticity (new)**:

4. $\varepsilon_s = \frac{\%\Delta Q_s}{\%\Delta P}$ — always positive. Same formula, same classification
5. **Linear supply through origin**: $\varepsilon_s = 1$ everywhere. Positive intercept: $\varepsilon_s > 1$
6. **Determinants**: flexibility, mobility, and replicability of inputs, **time horizon**
7. **Time**: very short run (inelastic) → short run → long run (elastic)
8. Both elasticities together determine **price volatility** and **tax burden sharing**

**Next (Lecture 16, April 16)**: Short Run and Long Run Market Equilibrium :balance_scale:

</div>

## Exercises {.center}

**Practice Time!** :pencil2:

Demand and supply elasticity calculation.

## Exercise 1: Multiple Choice

<div style="font-size: 82%;">

**Question**: The supply of hotel rooms in Porto's city center has $\varepsilon_s = 0.3$. If room prices increase by 20%, what happens to the quantity of rooms supplied?

A. Quantity increases by 60%

B. Quantity increases by 6%

C. Quantity increases by 0.6%

D. Quantity increases by 20%

::: {.fragment}
::: {.highlight-box}
**Answer: B**
:::

$\%\Delta Q_s = \varepsilon_s \times \%\Delta P = 0.3 \times 20\% = 6\%$

Supply is **inelastic** ($\varepsilon_s = 0.3 < 1$): even a large price increase (20%) produces only a small quantity response (6%). You can't build new hotel rooms quickly!
:::

</div>

## Exercise 2: Multiple Choice

<div style="font-size: 82%;">

**Question**: Demand for TAP flights from Lisbon to London has $|\varepsilon_d| = 1.8$. If TAP raises fares by 10%, what happens to total revenue from this route?

A. Revenue increases by 18%

B. Revenue increases because higher price means more money per ticket

C. Revenue falls because passenger numbers drop by more than 10%

D. Revenue stays the same

::: {.fragment}
::: {.highlight-box}
**Answer: C**
:::

Demand is **elastic** ($|\varepsilon_d| = 1.8 > 1$): a 10% fare increase causes approximately $1.8 \times 10\% = 18\%$ fewer passengers. Since quantity drops **proportionally more** than price rises, $TR = P \times Q$ **falls**. This is the **revenue test** for elastic demand.
:::

</div>

## Exercise 3: Open Question :writing_hand:

<div style="font-size: 60%;">

**The market for *tuk-tuk* tours in Lisbon** has the following estimated curves:

- **Demand**: $P = 36 - 0.2Q$ (€ per tour)
- **Supply**: $P = 6 + 0.1Q$ (€ per tour)

**a)** Find the equilibrium price and quantity.

**b)** Calculate the **price elasticity of demand** at the equilibrium point. Is demand elastic or inelastic?

**c)** Calculate the **price elasticity of supply** at the equilibrium point. Is supply elastic or inelastic?

**d)** At the equilibrium, if the city council imposes a small price increase (say €1), would you expect total revenue to rise or fall? Use the revenue test from part (b).

**e)** The supply has a positive intercept ($c = 6$). Using the shortcut formula $\varepsilon_s = \frac{P}{P - c}$, verify your answer from part (c). At what price would supply elasticity equal exactly 2?

**f)** The city council considers a €3 per-tour tax on operators. This shifts supply to $P = 9 + 0.1Q$. Find the new equilibrium. Who bears more of the tax burden — tourists or operators? Relate your answer to the elasticities you calculated.

</div>

## Exercise 3: Solution — Parts a, b & c

<div style="font-size: 66%;">

**a)** Equilibrium: $36 - 0.2Q = 6 + 0.1Q$ → $30 = 0.3Q$ → $Q^* = 100$, $P^* = 36 - 0.2(100) = €16$

**b)** Demand elasticity at $(Q^* = 100, P^* = 16)$:

Demand slope = $-0.2$, so $\frac{1}{\text{slope}} = \frac{1}{-0.2} = -5$

$$\varepsilon_d = -5 \times \frac{16}{100} = -0.8$$

$|\varepsilon_d| = 0.8 < 1$ → Demand is **inelastic** at the equilibrium.

**c)** Supply elasticity at $(Q^* = 100, P^* = 16)$:

Supply slope = $0.1$, so $\frac{1}{\text{slope}} = 10$

$$\varepsilon_s = 10 \times \frac{16}{100} = 1.6$$

$\varepsilon_s = 1.6 > 1$ → Supply is **elastic** at the equilibrium.

</div>

## Exercise 3: Solution — Parts d & e

<div style="font-size: 70%;">

**d)** Since demand is **inelastic** ($|\varepsilon_d| = 0.8 < 1$), the revenue test says: a price increase → total revenue **rises**. Quantity falls, but by a smaller percentage than price rises. The €1 increase would increase TR.

**e)** Shortcut: $\varepsilon_s = \frac{P}{P - c} = \frac{16}{16 - 6} = \frac{16}{10} = 1.6$ :white_check_mark: Matches part (c)!

For $\varepsilon_s = 2$: $\frac{P}{P - 6} = 2$ → $P = 2P - 12$ → $P = €12$

At $P = 12$: $Q = \frac{12 - 6}{0.1} = 60$ tours. So supply is more elastic at **lower prices** (closer to the intercept).

</div>

## Exercise 3: Solution — Part f

<div style="font-size: 68%;">

**f)** New supply with €3 tax: $P = 9 + 0.1Q$

New equilibrium: $36 - 0.2Q = 9 + 0.1Q$ → $27 = 0.3Q$ → $Q^* = 90$, $P^* = 36 - 0.2(90) = €18$

| | Before tax | After tax | Change |
|---|:---:|:---:|:---:|
| **Price paid by tourists** | €16 | €18 | +€2 |
| **Price received by operators** (net of tax) | €16 | €18 − €3 = €15 | −€1 |
| **Quantity** | 100 | 90 | −10 |

Of the €3 tax: tourists pay **€2 more**, operators receive **€1 less**.

:point_right: **Tourists bear more** (€2 out of €3) because **demand is more inelastic** ($|\varepsilon_d| = 0.8$) than supply is ($\varepsilon_s = 1.6$). The side with **less elastic** response absorbs more of the tax.

:bulb: This is a **general rule**: tax burden falls more heavily on the side of the market that is **less responsive** to price changes.

</div>

## Next Lecture {.center}

**April 16, 2026**: Short Run and Long Run Market Equilibrium :balance_scale:

We finally put **demand and supply together** to find prices and quantities!

. . .

:point_right: This is the payoff of Lectures 5–15: consumer theory meets producer theory!

## Thank You! {.center}

**Questions?** :raising_hand:

:email: paulo.fagandini@ext.universidadeeuropeia.pt

---

*Next class: Thursday, April 16, 2026*

::: {.notes}
This lecture serves double duty: it provides the demand elasticity content that was skipped (L9) in condensed form, then teaches supply elasticity as its mirror. The demand recap is kept to 3-4 slides covering definition, calculation, revenue test, and determinants — just enough for students to use the concept. The supply elasticity section follows the same structure. The exercise integrates both sides with a tax incidence analysis that requires comparing demand and supply elasticities, which is the conceptual payoff of learning both.
:::
